---
title: "Autogenous - Linkage network analysis"
author: "Luciano V Cosme"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: breezedark
    css:
      - "styles.css"
    toc: yes
    toc_float: no
    toc_depth: 5
editor_options:
  markdown:
    wrap: 120
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval                        = TRUE,
  echo                        = TRUE,
  cache                       = FALSE,
  class.output                = "bg-success"
)
knitr::opts_knit$set(
  root.dir = here::here()
)
```

> **Population label key:** The code labels used in the analysis below correspond to the following manuscript labels: AUT = AUTO, MAN = NON-AUTO, NEW = NON-AUTO-FIELD.

## 1. Install LDna and load the libraries

```{r install-ldna, eval=FALSE}
# Load devtools
library(devtools)


# Install LDna from GitHub
devtools::install_github("petrikemppainen/LDna", ref = 'v.2.15')
```

Clean env and memory
```{r clean-env-1}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```


Restart the R session if needed
```{r restart-r-session, eval=FALSE}
#.rs.restartR() # RStudio only
```

If you are using a MacOS you can download and install the library colorout https://github.com/jalvesaq/colorout to make the console fonte colorful.

To install libraries
```{r install-cran-packages, eval=FALSE}
# 1. Define the list of CRAN packages you want to install (this is what I like or use frequently)
cran_packages <- c(
  "tidyverse",
  "here",
  "dplyr",
  "flextable",
  "ggplot2",
  "scales",
  "reticulate",
  "extrafont",
  "stringr",
  "officer",
  "ggrepel",
  "Cairo",
  "reshape2",
  "viridis",
  "igraph",
  "writexl"
)

# 2. Function to check and install missing CRAN packages
install_if_missing <- function(packages) {
  # Identify packages that are not yet installed
  missing_packages <- packages[!(packages %in% installed.packages()[,"Package"])]

  # If there are missing packages, install them
  if(length(missing_packages)) {
    install.packages(missing_packages, dependencies = TRUE)
  } else {
    message("All CRAN packages are already installed.")
  }
}

# 3. Install missing CRAN packages
install_if_missing(cran_packages)
```


Load the libraries
```{r libraries, message=FALSE, warning=FALSE, results='hide'}
# Load tidyverse FIRST to avoid rlang version conflict
library(tidyverse)
library(here)
library(scales)
library(ggrepel)

# Then add LDna v.2.15 library path and load
.libPaths(c("/workspace/lib/ldna_v215", .libPaths()))
library(LDna)
library(reshape2)
library(writexl)
```


## 2. Data Tidying

We will have to estimate LD with Plink and prepare the data to use it with LDna. The first step is to decided with SNP set we will use. For example, we did LD pruning for some analysis, so we should not use those files to start the linkage network analysis. We also have to decided about the minor allele frequency threshold we will use.

We can remove the individuals that failed some of our tests. One mosquito, AUT 399, failed the heterozygosity test. None failed the relatedness test, and 867 SNPs failed the HWE test.
We can start with file4 from the quality control. We can remove the mosquito and the SNPs that did not pass the HWE test.

Create file to remove mosquito

```{bash}
echo "AUT 399" > output/quality_control/remove_aut_399.txt
```

Remove mosquito and SNPs
```{bash}
plink2 \
--allow-extra-chr \
--bfile output/quality_control/file4 \
--remove output/quality_control/remove_aut_399.txt \
--extract output/quality_control/passed_hwe.txt \
--make-bed \
--out output/ldna/files/file1 \
--silent;

grep "samples\|variants" output/ldna/files/file1.log
```

The next step is to create the chromosomal scale for this data set. Check the bim file now to see that it is using the scaffold scale

```{bash}
head output/ldna/files/file1.bim
```

### 2.1 Create chromosomal scale

Import the .bim file with the SNPs to create a new chromosomal scale.

```{r import-bim-file}
# Import the function
source(
  here(
    "notebooks", "helpers", "import_bim.R")
)

# Import the data
snps <- import_bim(here("output", "ldna", "files", "file1.bim"))

# Check it
head(snps)
```


Separate the tibbles into each chromosome.

```{r separate_data_into_chroms}
#   ____________________________________________________________________________
#   separate the SNP data per chromosome                                    ####
# chr1
chr1_snps <-
  snps |>
  filter(
    str_detect(
      Scaffold, "^1."
    )
  ) |> # here we get only Scaffold rows starting with 1
  as_tibble() # save as tibble
#
# chr2
chr2_snps <-
  snps |>
  filter(
    str_detect(
      Scaffold, "^2."
    )
  ) |>
  as_tibble()
#
# chr3
chr3_snps <-
  snps |>
  filter(
    str_detect(
      Scaffold, "^3."
    )
  ) |>
  as_tibble()
```

Import the file with sizes of each scaffold.

```{r import-scaffold-sizes}
#   ____________________________________________________________________________
#   import the file with the scaffold sizes                                 ####
sizes <-
  read_delim(
    here(
      "data", "genome", "scaffold_sizes.txt"
    ),
    col_names      = FALSE,
    show_col_types = FALSE,
    col_types      = "cd"
  )
#
# set column names
colnames(
  sizes
) <- c(
  "Scaffold", "Size"
)
#   ____________________________________________________________________________
#   create new column with the chromosome number                            ####
sizes <-
  sizes |>
  mutate(
    Chromosome = case_when( # we use mutate to create a new column called Chromosome
      startsWith(
        Scaffold, "1"
      ) ~ "1", # use startsWith to get Scaffold rows starting with 1 and output 1 on Chromosome column
      startsWith(
        Scaffold, "2"
      ) ~ "2",
      startsWith(
        Scaffold, "3"
      ) ~ "3"
    )
  ) |>
  arrange(
    Scaffold
  )                   # to sort the order of the scaffolds, fixing the problem we have with scaffold 1.86
# check it
head(sizes)
```

Create new scale. Get the scaffolds for each chromosome.

```{r get_list_scaffolds_per-chr}
#   ____________________________________________________________________________
#   separate the scaffold sizes tibble per chromosome                       ####
# chr1
chr1_scaffolds <-
  sizes |>
  filter(
    str_detect(
      Scaffold, "^1" # we use library stringr to get scaffolds starting with 1 (chromosome 1)
    )
  ) |>
  as_tibble()
#
# chr2
chr2_scaffolds <-
  sizes |>
  filter(
    str_detect(
      Scaffold, "^2" # we use library stringr to get scaffolds starting with 2 (chromosome 2)
    )
  ) |>
  as_tibble()
#
# # chr3
chr3_scaffolds <-
  sizes |>
  filter(
    str_detect(
      Scaffold, "^3" # we use library stringr to get scaffolds starting with 3 (chromosome 3)
    )
  ) |>
  as_tibble()
```

Create a scale for each chromosome.

```{r create_chr_scale}
#   ____________________________________________________________________________
#   create a new scale for each chromosome                                  ####
# chr1
chr1_scaffolds$overall_size_before_bp <-
  0                                                                        # we create a new column with zeros
for (i in 2:nrow(
  chr1_scaffolds
)
) {                                                                        # loop to start on second line
  chr1_scaffolds$overall_size_before_bp[i] <-                              # set position on the scale
    chr1_scaffolds$overall_size_before_bp[i - 1] + chr1_scaffolds$Size[i - # add the scaffold size and the location to get position on new scale
      1]
}
#
# chr2
chr2_scaffolds$overall_size_before_bp <- 0
for (i in 2:nrow(
  chr2_scaffolds
)
) {
  chr2_scaffolds$overall_size_before_bp[i] <-
    chr2_scaffolds$overall_size_before_bp[i - 1] + chr2_scaffolds$Size[i -
      1]
}
#
# chr3
chr3_scaffolds$overall_size_before_bp <- 0
for (i in 2:nrow(
  chr3_scaffolds
)
) {
  chr3_scaffolds$overall_size_before_bp[i] <-
    chr3_scaffolds$overall_size_before_bp[i - 1] + chr3_scaffolds$Size[i -
      1]
}
```

Merge the data frames scaffolds and SNPs.

```{r merge_snps_scales}
#   ____________________________________________________________________________
#   merge the data sets using the tidyverse function left_join              ####
# chr1
chr1_scale <-
  chr1_snps |>          # create data frame for each chromosome, get chr1_snps
  left_join(            # use lef_join function to merge it with chr1_scaffolds
    chr1_scaffolds,
    by = "Scaffold"
  ) |>                  # set column to use for merging (Scaffold in this case)
  na.omit() |>          # remove NAs, we don't have SNPs in every scaffold
  mutate(
    midPos_fullseq = as.numeric(
      Position
    ) +                 # make new columns numeric
      as.numeric(
        overall_size_before_bp
      )
  )
#
# chr2
chr2_scale <-
  chr2_snps |>
  left_join(
    chr2_scaffolds,
    by = "Scaffold"
  ) |>
  na.omit() |>
  mutate(
    midPos_fullseq = as.numeric(
      Position
    ) +
      as.numeric(
        overall_size_before_bp
      )
  )
#
# chr3
chr3_scale <-
  chr3_snps |>
  left_join(
    chr3_scaffolds,
    by = "Scaffold"
  ) |>
  na.omit() |>
  mutate(
    midPos_fullseq = as.numeric(
      Position
    ) +
      as.numeric(
        overall_size_before_bp
      )
  )
```

Merge all chromosome scales.

```{r bind_chr_scales}
#   ____________________________________________________________________________
#   merge the data sets, and select only the columns we need                ####
chroms <- rbind(
  chr1_scale, chr2_scale, chr3_scale
) |>
  dplyr::select(
    Chromosome, SNP, Cm, midPos_fullseq, Allele1, Allele2
  )
# check it
head(chroms)
```

Save the new .bim file

```{r save-new-bim-file}
#   ____________________________________________________________________________
#   save the new bim file with a new name, I added "B"                      ####
write.table(
  chroms,
  file      = here(
    "output", "ldna", "files", "file1B.bim"
  ),
  sep       = "\t",
  row.names = FALSE,
  col.names = FALSE,
  quote     = FALSE
)
```

Rename the .bim files
```{bash}
# change the name of the first .bim file, for example, append _backup.bim, and then replace the original file
mv output/ldna/files/file1.bim output/ldna/files/file1_backup.bim;
# than change the new bim we create to the original name (do it only once, otherwise it will mess up)
mv output/ldna/files/file1B.bim output/ldna/files/file1.bim
```

Create a new bed file with Plink2 to see if it works. For example, to see if the variants are in the
right order. Plink2 will give us a warning.

```{bash}
plink2 \
--bfile output/ldna/files/file1 \
--make-bed \
--out output/ldna/test01;
# then we remove the files
rm output/ldna/test01.*
```
No warnings from Plink2. Now, we can go ahead with our analysis.


Clean env and memory
```{r clean-env-2}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```

### 2.2 Subset by family

Estimate frequency
```{bash}
plink \
--keep-allele-order \
--bfile output/ldna/files/file1 \
--make-bed \
--freqx \
--out output/ldna/files/frq \
--silent;

grep "people\|variants" output/ldna/files/frq.log
```

Now we can use bash to check if there SNPs with low heterozygosity
```{bash}
# count SNPs with het < 0.5 (we have 60 mosquitoes)
cat output/ldna/files/frq.frqx | awk '{ if ($6 <= 30) print }' | wc -l
```

How many SNPs
```{r count-het-snps}
110353 - 104625
```
We can remove this SNPs since a limiting factor for the linkage network analysis is memory

Create a file with the SNPs
```{bash}
# get list of snps
cat output/ldna/files/frq.frqx | awk '{ if ($6 <= 30) print }' | awk '{print $2}' > output/ldna/files/snps_het.txt
```

We can start our analysis with MAF of 5%. If we have memory issues we can use 10% next. We will also set the genotyping missingness to zero.
```{bash}
# Filter MAF 5% and do not allow missing genotypes
plink \
--keep-allele-order \
--bfile output/ldna/files/file1 \
--out output/ldna/files/file2 \
--maf 0.05 \
--geno 0 \
--make-bed \
--extract output/ldna/files/snps_het.txt \
--silent;

grep "people\|variants" output/ldna/files/file2.log
```

Check the fam file families
```{bash}
# get the list of families
cat output/ldna/files/file2.fam | awk '!seen[$1]++' | awk '{print $1}'
```

We can check how many mosquitoes per population as well
```{bash get_pops_n, cache=TRUE}
awk '{print $1}' output/ldna/files/file2.fam | sort | uniq -c | awk '{print $2, $1}'
```

I did the analysis using the 3 populations, however the results of MAN were odd because we have only 10 mosquitoes. Therefore, I decided not to do linkage network analysis with MAN. I finished the analysis to see the odd pattern, but I came back and changed the code from this part foward to include only AUT and NEW for the linkage network analysis. I will still create files for each family, however the I will use only the shared SNPs between NEW and AUT. We will not consider MAN. I left it in the code, but we will not use it for comparisons.

Create a file for each of them
```{bash}
# make a text file with the name of each population
for pop in $(cat output/ldna/files/file2.fam | awk '!seen[$1]++' | awk '{print $1}');
do
  echo $pop > output/ldna/files/$pop\.txt
done
```

Now use Plink to create a file for each family. Now we set a MAF threshold of 5% within each family
```{bash}
# use plink to create a plink file for each population
for pop in $(cat output/ldna/files/file2.fam | awk '!seen[$1]++' | awk '{print $1}');
do
  plink --keep-allele-order --allow-no-sex --bfile output/ldna/files/file2 --make-bed --keep-fam output/ldna/files/$pop\.txt --out output/ldna/files/$pop --geno 0 --maf 0.05 --silent
done
```

Now we can check the number of SNPs and samples in each file

AUT
```{bash}
grep "people\|variants" output/ldna/files/AUT.log
```

MAN
```{bash}
grep "people\|variants" output/ldna/files/MAN.log
```

NEW
```{bash}
grep "people\|variants" output/ldna/files/NEW.log
```


We can see that the number of SNPs kept is not the same. We need to get the shared SNPs. We can import the bim files and use R to get the intersect

```{r import-pop-bim-files}
# Import the data
AUT <- import_bim(here("output", "ldna", "files", "AUT.bim"))
MAN <- import_bim(here("output", "ldna", "files", "MAN.bim")) # not using MAN because it has only 10 mosquitoes
NEW <- import_bim(here("output", "ldna", "files", "NEW.bim"))
```

Now get the shared SNPs

```{r find-common-snps}
# Identify common SNPs
# common_snps <- Reduce(intersect, list(AUT$SNP, MAN$SNP, NEW$SNP))
common_snps <- Reduce(intersect, list(AUT$SNP, NEW$SNP))

# Create a data frame with the common SNPs
common_snps_df <- data.frame(SNP = common_snps)

# Count them
length(common_snps)

# Write the data frame to a file
write.table(common_snps_df, here("output", "ldna", "files","common_snps.txt"), quote = FALSE, row.names = FALSE, col.names = FALSE)
```

We have 32,732 SNPs. It seems we remove a lot but still too much for LDna in a laptop. We can use the cluster but first I will try using a laptop.

Now we have to repeat the previous step when we create a file for each family but now using only the SNPs that have 5% MAF in each family

```{bash}
# use plink to create a plink file for each population
for pop in $(cat output/ldna/files/file2.fam | awk '!seen[$1]++' | awk '{print $1}');
do
  plink --keep-allele-order --allow-no-sex --bfile output/ldna/files/file2 --make-bed --keep-fam output/ldna/files/$pop\.txt --out output/ldna/files/$pop --extract output/ldna/files/common_snps.txt --silent
done
```


Now we can check the number of SNPs and they all should be the same

AUT
```{bash}
grep "people\|variants" output/ldna/files/AUT.log
```

MAN
```{bash}
grep "people\|variants" output/ldna/files/MAN.log
```

NEW
```{bash}
grep "people\|variants" output/ldna/files/NEW.log
```

Perfect. Now we can split the data by chromosome for each family.

### 2.3 Subset by chromosome within each family

We can create a new directory
```{bash}
# Create a directory
mkdir -p output/ldna/pop;

# We can copy the family files there
cp output/ldna/files/AUT* output/ldna/pop;
cp output/ldna/files/MAN* output/ldna/pop;
cp output/ldna/files/NEW* output/ldna/pop;

# remove the files we dont need
rm output/ldna/pop/*.txt;
rm output/ldna/pop/*.nosex
```

We can now get a list of SNPs for each chromosome. All the populations have the same SNPs, so we can use any of the files.

```{bash}
cat output/ldna/pop/AUT.bim | awk '$1 == 1' | awk '{print $2}' > output/ldna/pop/chr1_snps.txt;
cat output/ldna/pop/AUT.bim | awk '$1 == 2' | awk '{print $2}' > output/ldna/pop/chr2_snps.txt;
cat output/ldna/pop/AUT.bim | awk '$1 == 3' | awk '{print $2}' > output/ldna/pop/chr3_snps.txt
```

### 2.4 Estimate LD with Plink for each chromosome

```{bash}
# chr1
for pop in $(ls -1 output/ldna/pop/*.bim | sed 's/output\/ldna\/pop\///' | sed 's/\.[^.]*$//');
do
  plink --allow-no-sex --keep-allele-order --bfile output/ldna/pop/$pop --out output/ldna/pop/$pop\.chr1 --r2 triangle gz --extract output/ldna/pop/chr1_snps.txt --silent
done;

# chr2
for pop in $(ls -1 output/ldna/pop/*.bim | sed 's/output\/ldna\/pop\///' | sed 's/\.[^.]*$//');
do
  plink --allow-no-sex --keep-allele-order --bfile output/ldna/pop/$pop --out output/ldna/pop/$pop\.chr2 --r2 triangle gz --extract output/ldna/pop/chr2_snps.txt --silent
done;

# chr3
for pop in $(ls -1 output/ldna/pop/*.bim | sed 's/output\/ldna\/pop\///' | sed 's/\.[^.]*$//');
do
  plink --allow-no-sex --keep-allele-order --bfile output/ldna/pop/$pop --out output/ldna/pop/$pop\.chr3 --r2 triangle gz --extract output/ldna/pop/chr3_snps.txt --silent
done
```

Create directory for each chromosome and move the files
```{bash}
# make dirs and remove unnecessary files
rm output/ldna/pop/*.nosex;
mkdir -p output/ldna/pop/chr1; mv output/ldna/pop/*chr1.* output/ldna/pop/chr1/;
mkdir -p output/ldna/pop/chr2; mv output/ldna/pop/*chr2.* output/ldna/pop/chr2/;
mkdir -p output/ldna/pop/chr3; mv output/ldna/pop/*chr3.* output/ldna/pop/chr3/;

# copy files
cp output/ldna/pop/chr1*.txt output/ldna/pop/chr1/;
cp output/ldna/pop/chr2*.txt output/ldna/pop/chr2/;
cp output/ldna/pop/chr3*.txt output/ldna/pop/chr3/
```



### 2.5 Prepare LD matrices

Prepare files for LDna
```{bash}
# chr1
cat output/ldna/pop/chr1_snps.txt | awk '{print $1}' > output/ldna/pop/chr1/snps1.txt;
echo "" > output/ldna/pop/chr1/snps2.txt;
cat output/ldna/pop/chr1/snps2.txt output/ldna/pop/chr1/snps1.txt | gzip -9 > output/ldna/pop/chr1/snps3.txt.gz;
cat output/ldna/pop/chr1/snps1.txt | tr '\n' ' ' |  awk -v OFS='\t' '{$1=$1}1' | gzip -9 > output/ldna/pop/chr1/header.txt.gz;

# chr2
cat output/ldna/pop/chr2_snps.txt | awk '{print $1}' > output/ldna/pop/chr2/snps1.txt;
echo "" > output/ldna/pop/chr2/snps2.txt;
cat output/ldna/pop/chr2/snps2.txt output/ldna/pop/chr2/snps1.txt | gzip -9 > output/ldna/pop/chr2/snps3.txt.gz;
cat output/ldna/pop/chr2/snps1.txt | tr '\n' ' ' |  awk -v OFS='\t' '{$1=$1}1' | gzip -9 > output/ldna/pop/chr2/header.txt.gz;

# chr3
cat output/ldna/pop/chr3_snps.txt | awk '{print $1}' > output/ldna/pop/chr3/snps1.txt;
echo "" > output/ldna/pop/chr3/snps2.txt;
cat output/ldna/pop/chr3/snps2.txt output/ldna/pop/chr3/snps1.txt | gzip -9 > output/ldna/pop/chr3/snps3.txt.gz;
cat output/ldna/pop/chr3/snps1.txt | tr '\n' ' ' |  awk -v OFS='\t' '{$1=$1}1' | gzip -9 > output/ldna/pop/chr3/header.txt.gz;
```

Add header to the matrices
```{bash}
# chr1
for pop in $(ls -1 output/ldna/pop/chr1/*.ld.gz | sed 's/output\/ldna\/pop\/chr1\///' | sed 's/\.[^.]*$//');
do
  cat output/ldna/pop/chr1/header.txt.gz output/ldna/pop/chr1/$pop\.gz > output/ldna/pop/chr1/$pop\.txt.gz
done;

# chr2
for pop in $(ls -1 output/ldna/pop/chr2/*.ld.gz | sed 's/output\/ldna\/pop\/chr2\///' | sed 's/\.[^.]*$//');
do
  cat output/ldna/pop/chr2/header.txt.gz output/ldna/pop/chr2/$pop\.gz > output/ldna/pop/chr2/$pop\.txt.gz
done;

# chr3
for pop in $(ls -1 output/ldna/pop/chr3/*.ld.gz | sed 's/output\/ldna\/pop\/chr3\///' | sed 's/\.[^.]*$//');
do
  cat output/ldna/pop/chr3/header.txt.gz output/ldna/pop/chr3/$pop\.gz > output/ldna/pop/chr3/$pop\.txt.gz
done
```


We can rename the files
```{bash}
# rename the files (remove the ld)
for f in output/ldna/pop/chr1/*ld.txt.gz; do [ -f "$f" ] && mv "$f" "${f/ld.txt.gz/txt.gz}"; done
for f in output/ldna/pop/chr2/*ld.txt.gz; do [ -f "$f" ] && mv "$f" "${f/ld.txt.gz/txt.gz}"; done
for f in output/ldna/pop/chr3/*ld.txt.gz; do [ -f "$f" ] && mv "$f" "${f/ld.txt.gz/txt.gz}"; done
```

Now we can add the row names (this takes time and the output files are near 1Gb)
```{bash}
# chr1
for pop in $(ls -1 output/ldna/pop/chr1/*.chr1.txt.gz | sed 's/output\/ldna\/pop\/chr1\///' | sed 's/\.chr1\.txt\.gz$//');
do
  paste <(gzip -d < output/ldna/pop/chr1/snps3.txt.gz) <(gzip -d < output/ldna/pop/chr1/$pop.chr1.txt.gz) > output/ldna/pop/chr1/${pop}.chr1.txt
done

# chr2
for pop in $(ls -1 output/ldna/pop/chr2/*.chr2.txt.gz | sed 's/output\/ldna\/pop\/chr2\///' | sed 's/\.chr2\.txt\.gz$//');
do
  paste <(gzip -d < output/ldna/pop/chr2/snps3.txt.gz) <(gzip -d < output/ldna/pop/chr2/$pop.chr2.txt.gz) > output/ldna/pop/chr2/${pop}.chr2.txt
done

# chr3
for pop in $(ls -1 output/ldna/pop/chr3/*.chr3.txt.gz | sed 's/output\/ldna\/pop\/chr3\///' | sed 's/\.chr3\.txt\.gz$//');
do
  paste <(gzip -d < output/ldna/pop/chr3/snps3.txt.gz) <(gzip -d < output/ldna/pop/chr3/$pop.chr3.txt.gz) > output/ldna/pop/chr3/${pop}.chr3.txt
done
```

I deleted the MAN files for chromosome 2 and 3 since we will not estimate LD for them. I kept chromosome 1 to show how small sample size bias the results.


## 3. LDna chromosome 1

Clean env and memory
```{r clean-env-chr1}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```


### 3.1 NEW
Clean env and memory
```{r clean-env-new-chr1}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```

Import the data
```{r new-chr1-load-ld-matrix, eval=TRUE}
ld1 <-
  read.delim(
    here("output", "ldna", "pop", "chr1", "NEW.chr1.txt"),
    sep = "\t",
    header = T,
    row.names = 1,
    stringsAsFactors = F,
    check.names = FALSE
  )
```

Check the LD matrix we built using bash
```{r new-chr1-check-ld-matrix, eval=TRUE}
head(ld1)
```

The row and column names are correct. Now we can convert the data frame to matrix

```{r new-chr1-convert-to-matrix, eval=TRUE}
# Create the matrix for ldna
ld2<- as.matrix(sapply(ld1, as.numeric))

# Get names
names<- rownames(ld1)
#names<-gsub("-", ".", names)

# Add names
row.names(ld2)<-names

# Clean non-numeric values
ld2[is.nan(ld2)] <- NA
ld2[is.infinite(ld2)] <- NA

# Remove diagonal values = 1
diag(ld2)=NA

# Make sure it is only the lower triangle matrix
ld2[!lower.tri(ld2)] <- NA
```

Create a LDna object
```{r new-chr1-run-ldna-raw, eval=TRUE}
# create the LDna object (change the number of cores as needed)
diag(ld2) <- NA
ld2[!lower.tri(ld2)] <- NA
ldna <- LDnaRaw(ld2, mc.cores=2, method = "single")
```

Save ldna
```{r new-chr1-save-ldna, eval=TRUE}
saveRDS(ldna, file = here("output", "ldna", "pop", "chr1", "NEW.rds"))
saveRDS(ld2, file = here("output", "ldna", "pop", "chr1", "NEW_ld2.rds"))
```

Read the data
```{r new-chr1-read-ldna}
ldna <- readRDS(file = here("output", "ldna", "pop", "chr1", "NEW.rds"))
ld2 <- readRDS(file = here("output", "ldna", "pop", "chr1", "NEW_ld2.rds"))
```


Check the object
```{r new-chr1-check-ldna-str}
str(ldna)
```
```{r new-chr1-check-ld2-str}
str(ld2)
```

We can use the same parameters that we used for the other population
```{r new-chr1-set-edge-params}
# snp_count <- ncol(ld2)
edges_max <- 100
edges_min <- 20
```

We can test different number of edges. Here we set lambda.lim to 1. We use branch.traversal=TRUE so the clusters are extracted at the base of the branch
```{r new-chr1-extract-branches-loop}
pdf(file = here("output", "ldna", "pop", "chr1", paste(edges_min, "1.NEW_loop.edges.pdf", sep = "_")), width = 20, height = 12)
op <- par(mfcol=c(1,4))

# Adjust the step size here
step_size <- 10  # for example, try a smaller step like 10

# NOTE: This loop always runs extractClusters for ALL edges_estimate values
# because clusters_count_prev is reset to 100 on every iteration.
# The final saved result uses min.edges=100 (the last iteration),
# which matches the clean pipeline and manuscript results.
for (edges_estimate in seq(edges_min, edges_max, by = step_size)) {
  clusters_count_prev <- 100
  clusters_count <- 50

  # The 'if' condition seems redundant here since clusters_count_prev is always set to 100 before the check
  if (clusters_count_prev > clusters_count) {
    print(edges_estimate) # This prints the current edges_estimate, showing that the loop is working
    clusters_count_prev <- clusters_count

    # Extract clusters
    clusters <- extractBranches(ldna, min.edges=edges_estimate)

    # Summarize the clusters
    summary <- summaryLDna(ldna, clusters, ld2)

    # Write the summary to a file
    write.table(summary, file = here("output", "ldna", "pop", "chr1", "summary_NEW.txt"), row.names=FALSE, sep="\t", quote=FALSE)

    # Update clusters_count with the number of clusters just written
    clusters_count <- nrow(read.delim(here("output", "ldna", "pop", "chr1", "summary_NEW.txt")))
  }
}

# Reset graphical parameters and close all open graphic devices
par(op)
while (!is.null(dev.list())) dev.off()
```

Get cluster information
```{r new-chr1-get-cluster-info}
# to get the list of names of the clusters
# Find the length of the longest vector
max_length <- max(sapply(clusters, length))

# Function to pad vectors with NAs to make them the same length
pad_vector_to_max_length <- function(vec, max_length) {
  c(vec, rep(NA, max_length - length(vec)))
}

# Apply the padding function to each vector in the list and combine into a data frame
do1 <- as.data.frame(do.call(cbind, lapply(clusters, pad_vector_to_max_length, max_length)))

# to get list of snps for each cluster
do <- do.call(cbind, unname(lapply(clusters, `length<-`, max(lengths(clusters)))))
colnames(do) <- colnames(do1)
```

Create an objet with the clusters and the SNPs in each cluster
```{r new-chr1-build-cluster-df}
max_length <- max(sapply(clusters, length))
# Initialize an empty list
cluster_df_list <- list()

# Loop through each cluster and pad with NAs
for (cluster_name in names(clusters)) {
  cluster_length <- length(clusters[[cluster_name]])
  padded_cluster <- c(clusters[[cluster_name]], rep(NA, max_length - cluster_length))
  cluster_df_list[[cluster_name]] <- padded_cluster
}

# Combine the lists into a data frame
df <- data.frame(cluster_df_list)
# Remove 'X' from column names
names(df) <- make.names(names(df), unique = FALSE)
names(df) <- sub("X", "", names(df), fixed = TRUE)

# Save it
saveRDS(df, file = here("output", "ldna", "pop", "chr1", "NEW_clusters_snps.rds"))
head(df)
```
We can save it in a different format as well
```{r new-chr1-clusters-long-format}
# Assuming your data frame is named df
long_df <- gather(df, Cluster, SNP, everything())

# View the first few rows of the transformed data
head(long_df)

# Save it
saveRDS(long_df, file = here("output", "ldna", "pop", "chr1", "NEW_clusters_snps1.rds"))
```


Make long format
```{r new-chr1-melt-clusters}
new_chr1 <- melt(do, na.rm = T, value.name = "value")
head(new_chr1)
```


Update names
```{r new-chr1-update-colnames}
colnames(new_chr1)<- c("v1", "cluster", "SNP")
head(new_chr1)
```



Import the bim file to get the SNP positions
```{r new-chr1-import-bim-positions}
# Import the data
new_snps_chr1 <- import_bim(here("output", "ldna", "pop", "NEW.bim")) |>
  dplyr::select(
    SNP, Scaffold, Position
  ) |>
  dplyr::rename(
    Chromosome = Scaffold
  )

# Check it
head(new_snps_chr1)
```

Merge the cluster data and the SNP data
```{r new-chr1-merge-clusters-positions}
# merge dataframes
merged<- merge(new_chr1, new_snps_chr1, by="SNP", all.x=TRUE)
merged<-na.omit(merged)
head(merged)
```

Select the columns we need
```{r new-chr1-select-cluster-cols}
# subset
merged <- subset(merged, select = c(cluster, SNP, Position))
head(merged)
```

Sort by distance
```{r new-chr1-sort-by-position}
merged<- merged[order(merged$Position),,drop=FALSE]
head(merged)
```


```{r new-chr1-prep-sushi-data}
# for parsing later
sushi1<- subset(merged, select = c(cluster, Position))
head(sushi1)
```

Save the data
```{r new-chr1-save-cluster-files, warning=FALSE}
# Create directory
new_directory_path <- here("output", "ldna", "pop", "chr1", "clusters")
dir.create(new_directory_path, recursive = TRUE)

# save file
write.table(merged, file = here("output", "ldna", "pop", "chr1", "SNPs_clusters_NEW_chr1.txt"), row.names = F, sep = "\t", quote = F)
write.table(sushi1, file = here("output", "ldna", "pop", "chr1", "clusters", "sushi_NEW_chr1.txt"), row.names = F, sep = "\t", quote = F)
```

Parse the file1 to get the start and end of the clusters as well as their size. Remember, recombination make it a mosaic, so the clusters are "mixed".
```{bash}
awk '
function print_row() {
    if ( feature != "" )
       print feature, start, end, (end - start)
}

BEGIN {
    FS=OFS="\t";
    print "Cluster", "Start", "End", "Size";
}
NR == 1 { next } # Skip the first line
$1 != feature {
    print_row();
    feature = $1;
    start = $2;
    end = $2;
    next;
}
{
    end = $2;
}
END {
    print_row();
}
' output/ldna/pop/chr1/clusters/sushi_NEW_chr1.txt > output/ldna/pop/chr1/clusters/sushi2_NEW_chr1.txt;
head output/ldna/pop/chr1/clusters/sushi2_NEW_chr1.txt
```

Count how many segments
```{r new-chr1-count-cluster-segments}
# Define the path
input_path <- here("output", "ldna", "pop", "chr1", "clusters", "sushi2_NEW_chr1.txt")

# Read the data into R
clusters_data <- read_table(input_path, col_names = TRUE, col_types = NULL)

# Arrange by position
clusters_data <- clusters_data |>
  arrange(Start)

# Count segments for each cluster - if you want only the counts
segment_counts <- clusters_data %>%
  group_by(Cluster) %>%
  summarize(nSegments = n(), .groups = 'drop')

# View the result
head(segment_counts)
```


Plot it
```{r new-chr1-plot-clusters-basic, fig.height=5, fig.width=8}
# Define the path
input_path <-
  here("output",
       "ldna",
       "pop",
       "chr1",
       "clusters",
       "sushi2_NEW_chr1.txt")

# Read the data into R
clusters_data <- read.table(input_path, header = TRUE, sep = "\t") |>
  arrange(Start)

# Filter out rows with Size equal to 0
clusters_data <- subset(clusters_data, Size > 100000) # show only bigger than 100kb


# Calculate the maximum size for scaling
max_size <- max(clusters_data$Size)

# Create Start_Mb and End_Mb within clusters_data
clusters_data$Start_Mb <- clusters_data$Start / 1e6
clusters_data$End_Mb <- clusters_data$End / 1e6

# Create the plot
ggplot(clusters_data, aes(xmin = Start_Mb, xmax = End_Mb, ymin = 0, ymax = Size / 1e6)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = label_number(unit = "M"), breaks = pretty_breaks(n = 10)) +
  scale_y_continuous(labels = label_number(unit = "M"), breaks = pretty_breaks(n = 10)) +
  labs(x = "Chromosome 1 (Mb)", y = "Cluster Size (Mb)", fill = "Cluster ID") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    panel.grid.minor.x = element_blank(),
    legend.position = "top",
    legend.box = "horizontal"
  ) +
  guides(fill = guide_legend(nrow = 4, title.position = "top", title.hjust = 0.5))


# # Use ggsave to save the plot as a PDF
# ggsave(
#   filename = here("output", "ldna", "pop", "chr1", "clusters", "NEW_chr1.pdf"),
#   device = "pdf",
#   width = 8,
#   height = 5,
#   units = "in"
# )
```


We can annotate the number of segments in the legend.

```{r new-chr1-plot-clusters-annotated, fig.height=5, fig.width=8}
# Merge SNP counts into clusters_data while specifying suffixes
annotated_data <- merge(clusters_data, segment_counts, by = "Cluster")

# Create a new column with Cluster names and SNP counts
annotated_data$Cluster_with_SNPs <- with(annotated_data, paste(Cluster, " (", nSegments, ")", sep=""))

# Now the plot uses the new Cluster_with_SNPs for the fill legend
ggplot(annotated_data, aes(xmin = Start_Mb, xmax = End_Mb, ymin = 0, ymax = Size / 1e6)) +
  geom_rect(aes(fill = as.factor(Cluster_with_SNPs)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = scales::label_number(unit = "M"), breaks = scales::pretty_breaks(n = 10)) +
  scale_y_continuous(labels = scales::label_number(unit = "M"), breaks = scales::pretty_breaks(n = 10)) +
  labs(x = "Chromosome 1 (Mb)", y = "Cluster Size (Mb)", fill = "Cluster ID") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    panel.grid.minor.x = element_blank(),
    legend.position = "top",
    legend.box = "horizontal"
  ) +
  guides(fill = guide_legend(nrow = 5, title.position = "top", title.hjust = 0.5))

# Use ggsave to save the plot as a PDF
ggsave(
  filename = here("output", "ldna", "pop", "chr1", "clusters", "NEW_chr1.pdf"),
  device = "pdf",
  width = 8,
  height = 5,
  units = "in"
)
```


We can save the data to plot all chromosomes together using a facet plot
```{r new-chr1-check-annotated-data}
head(annotated_data)
```
Save plotting data, we can arrange the column order first
```{r new-chr1-save-plot-data}
plot_data <- annotated_data |>
  dplyr::select(
    Cluster, nSegments, Start, End, Size, Start_Mb, End_Mb
  )

# Create a new column with Chromosome
plot_data <- data.frame(Chromosome = rep(1, nrow(plot_data)), plot_data)

# Split the 'Cluster' column at "_", and create the 'r2' column with the second part
plot_data$r2 <- sapply(strsplit(as.character(plot_data$Cluster), "_"), `[`, 2)

# Move the 'r2' column to be the second column, right after 'Chromosome'
plot_data <- plot_data[, c(1:2, ncol(plot_data), 3:(ncol(plot_data)-1))]


# Arrange
plot_data <- plot_data |>
  dplyr::arrange(Start)

# Check it
head(plot_data)

# Save it
saveRDS(plot_data, file = here("output", "ldna", "pop", "chr1", "NEW_plot.rds"))
```

We can also import the summary file to create a object with SNP counts per segment
```{r new-chr1-build-cluster-seg-snp}
# Define the path
input_path <-
  here("output",
       "ldna",
       "pop",
       "chr1",
       "summary_NEW.txt")

# Read the data into R
clusters_snps <- read.table(input_path, header = TRUE, sep = "\t")

# We subset and rename the column 1
clusters_snps <- clusters_snps |>
  dplyr::select(
    Name, nLoci
  ) |>
  dplyr::rename(
    Cluster = Name,
    nSNPs = nLoci
  )

# Merge
cluster_seg_snp <- inner_join(segment_counts, clusters_snps, by = "Cluster")
cluster_seg_snp <- inner_join(clusters_data, cluster_seg_snp, by = "Cluster")

# Split the 'Cluster' column at "_", and create the 'r2' column with the second part
cluster_seg_snp$r2 <- sapply(strsplit(as.character(cluster_seg_snp$Cluster), "_"), `[`, 2)

# Add chromosome number and population name
cluster_seg_snp <- data.frame(Chromosome = rep(1, nrow(cluster_seg_snp)), cluster_seg_snp)
cluster_seg_snp <- data.frame(Population = rep("NEW", nrow(cluster_seg_snp)), cluster_seg_snp)


# Reorder the columns
cluster_seg_snp <- cluster_seg_snp |>
  dplyr::select(
    Chromosome, Cluster, r2, Start, End, nSegments, nSNPs
  ) |>
  dplyr::arrange(
    Chromosome, Start
  )

head(cluster_seg_snp)


# Save the data for plotting later
saveRDS(cluster_seg_snp, file = here("output", "ldna", "pop", "chr1", "NEW_plot2.rds"))
```


### 3.2 AUT

Clean env and memory
```{r clean-env-aut-chr1}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```

Import the data
```{r aut-chr1-load-ld-matrix, eval=TRUE}
ld1 <-
  read.delim(
    here("output", "ldna", "pop", "chr1", "AUT.chr1.txt"),
    sep = "\t",
    header = T,
    row.names = 1,
    stringsAsFactors = F,
    check.names = FALSE
  )
```

Check the LD matrix we built using bash
```{r aut-chr1-check-ld-matrix, eval=TRUE}
head(ld1)
```

The row and column names are correct. Now we can convert the data frame to matrix

```{r aut-chr1-convert-to-matrix, eval=TRUE}
# Create the matrix for ldna
ld2<- as.matrix(sapply(ld1, as.numeric))

# Get names
names<- rownames(ld1)
#names<-gsub("-", ".", names)

# Add names
row.names(ld2)<-names

# Clean non-numeric values
ld2[is.nan(ld2)] <- NA
ld2[is.infinite(ld2)] <- NA

# Remove diagonal values = 1
diag(ld2)=NA

# Make sure it is only the lower triangle matrix
ld2[!lower.tri(ld2)] <- NA
```

Create a LDna object
```{r aut-chr1-run-ldna-raw, eval=TRUE}
# create the LDna object (change the number of cores as needed)
diag(ld2) <- NA
ld2[!lower.tri(ld2)] <- NA
ldna <- LDnaRaw(ld2, mc.cores=2, method = "single")
```

Check the object
```{r aut-chr1-check-ldna-str, eval=TRUE}
str(ldna)
```

Save it
```{r aut-chr1-save-ldna, eval=TRUE}
saveRDS(ldna, file = here("output", "ldna", "pop", "chr1", "AUT.rds"))
saveRDS(ld2, file = here("output", "ldna", "pop", "chr1", "AUT_ld2.rds"))
```

Import data
```{r aut-chr1-read-ldna}
ldna <- readRDS(here("output", "ldna", "pop", "chr1", "AUT.rds"))
ld2 <- readRDS(here("output", "ldna", "pop", "chr1", "AUT_ld2.rds"))
```


We can use the same parameters that we used for the other population
```{r aut-chr1-set-edge-params}
edges_max <- 100
edges_min <- 20
```

We can test the number of edges in increments of 10 and use the default phi=2
```{r aut-chr1-extract-branches-loop}
pdf(file = here("output", "ldna", "pop", "chr1", paste(edges_min, "1.AUT_loop.edges.pdf", sep = "_")), width = 20, height = 12)
op <- par(mfcol=c(1,4))

# Adjust the step size here
step_size <- 10  # for example, try a smaller step like 10

for (edges_estimate in seq(edges_min, edges_max, by = step_size)) {
  clusters_count_prev <- 100
  clusters_count <- 50

  # The 'if' condition seems redundant here since clusters_count_prev is always set to 100 before the check
  if (clusters_count_prev > clusters_count) {
    print(edges_estimate) # This prints the current edges_estimate, showing that the loop is working
    clusters_count_prev <- clusters_count

    # Extract clusters
    clusters <- extractBranches(ldna, min.edges=edges_estimate)

    # Summarize the clusters
    summary <- summaryLDna(ldna, clusters, ld2)

    # Write the summary to a file
    write.table(summary, file = here("output", "ldna", "pop", "chr1", "summary_AUT.txt"), row.names=FALSE, sep="\t", quote=FALSE)

    # Update clusters_count with the number of clusters just written
    clusters_count <- nrow(read.delim(here("output", "ldna", "pop", "chr1", "summary_AUT.txt")))
  }
}

# Reset graphical parameters and close all open graphic devices
par(op)
while (!is.null(dev.list())) dev.off()
```

Get cluster information
```{r aut-chr1-get-cluster-info}
# to get the list of names of the clusters
# Find the length of the longest vector
max_length <- max(sapply(clusters, length))

# Function to pad vectors with NAs to make them the same length
pad_vector_to_max_length <- function(vec, max_length) {
  c(vec, rep(NA, max_length - length(vec)))
}

# Apply the padding function to each vector in the list and combine into a data frame
do1 <- as.data.frame(do.call(cbind, lapply(clusters, pad_vector_to_max_length, max_length)))

# to get list of snps for each cluster
do <- do.call(cbind, unname(lapply(clusters, `length<-`, max(lengths(clusters)))))
colnames(do) <- colnames(do1)
```

Save it
```{r aut-chr1-build-cluster-df}
max_length <- max(sapply(clusters, length))
# Initialize an empty list
cluster_df_list <- list()

# Loop through each cluster and pad with NAs
for (cluster_name in names(clusters)) {
  cluster_length <- length(clusters[[cluster_name]])
  padded_cluster <- c(clusters[[cluster_name]], rep(NA, max_length - cluster_length))
  cluster_df_list[[cluster_name]] <- padded_cluster
}

# Combine the lists into a data frame
df <- data.frame(cluster_df_list)
# Remove 'X' from column names
names(df) <- make.names(names(df), unique = FALSE)
names(df) <- sub("X", "", names(df), fixed = TRUE)

# Save it
saveRDS(df, file = here("output", "ldna", "pop", "chr1", "AUT_clusters_snps.rds"))
head(df)
```
We can save it in a different format as well
```{r aut-chr1-clusters-long-format}
# Assuming your data frame is named df
long_df <- gather(df, Cluster, SNP, everything())

# View the first few rows of the transformed data
head(long_df)

# Save it
saveRDS(long_df, file = here("output", "ldna", "pop", "chr1", "AUT_clusters_snps1.rds"))
```


Make long format
```{r aut-chr1-melt-clusters}
aut_chr1 <- melt(do, na.rm = T, value.name = "value")
head(aut_chr1)
```


Update names
```{r aut-chr1-update-colnames}
colnames(aut_chr1)<- c("v1", "cluster", "SNP")
head(aut_chr1)
```


Import the bim file to get the SNP positions
```{r aut-chr1-import-bim-positions}
# Import the data
aut_snps_chr1 <- import_bim(here("output", "ldna", "pop", "AUT.bim")) |>
  dplyr::select(
    SNP, Scaffold, Position
  ) |>
  dplyr::rename(
    Chromosome = Scaffold
  )

# Check it
head(aut_snps_chr1)
```

Merge the cluster data and the SNP data
```{r aut-chr1-merge-clusters-positions}
# merge dataframes
merged<- merge(aut_chr1, aut_snps_chr1, by="SNP", all.x=TRUE)
merged<-na.omit(merged)
head(merged)
```

Select the columns we need
```{r aut-chr1-select-cluster-cols}
# subset
merged <- subset(merged, select = c(cluster, SNP, Position))
head(merged)
```

Sort by distance
```{r aut-chr1-sort-by-position}
merged<- merged[order(merged$Position),,drop=FALSE]
head(merged)
```


```{r aut-chr1-prep-sushi-data}
# for parsing later
sushi1<- subset(merged, select = c(cluster, Position))
head(sushi1)
```

Save the data
```{r aut-chr1-save-cluster-files, warning=FALSE}
# Create directory
new_directory_path <- here("output", "ldna", "pop", "chr1", "clusters")
dir.create(new_directory_path, recursive = TRUE)

# save file
write.table(merged, file = here("output", "ldna", "pop", "chr1", "SNPs_clusters_AUT_chr1.txt"), row.names = F, sep = "\t", quote = F)
write.table(sushi1, file = here("output", "ldna", "pop", "chr1", "clusters", "sushi_AUT_chr1.txt"), row.names = F, sep = "\t", quote = F)
```

Parse the file1 to get the start and end of the clusters as well as their size. Remember, recombination make it a mosaic, so the clusters are "mixed".
```{bash}
awk '
function print_row() {
    if ( feature != "" )
       print feature, start, end, (end - start)
}

BEGIN {
    FS=OFS="\t";
    print "Cluster", "Start", "End", "Size";
}
NR == 1 { next } # Skip the first line
$1 != feature {
    print_row();
    feature = $1;
    start = $2;
    end = $2;
    next;
}
{
    end = $2;
}
END {
    print_row();
}
' output/ldna/pop/chr1/clusters/sushi_AUT_chr1.txt > output/ldna/pop/chr1/clusters/sushi2_AUT_chr1.txt;
head output/ldna/pop/chr1/clusters/sushi2_AUT_chr1.txt
```

Get SNP count
```{r aut-chr1-count-cluster-segments}
# Define the path
input_path <- here("output", "ldna", "pop", "chr1", "clusters", "sushi2_AUT_chr1.txt")

# Read the data into R
clusters_data <- read_table(input_path, col_names = TRUE, col_types = NULL)

# Arrange by position
clusters_data <- clusters_data |>
  arrange(Start)

# Count segments for each cluster - if you want only the counts
segment_counts <- clusters_data %>%
  group_by(Cluster) %>%
  summarize(nSegments = n(), .groups = 'drop')

# View the result
head(segment_counts)
```

Plot it
```{r aut-chr1-plot-clusters-basic, fig.height=5, fig.width=8}
# Define the path
input_path <-
  here("output",
       "ldna",
       "pop",
       "chr1",
       "clusters",
       "sushi2_AUT_chr1.txt")

# Read the data into R
clusters_data <- read.table(input_path, header = TRUE, sep = "\t") |>
  arrange(Start)

# Arrange by position
clusters_data <- clusters_data |>
  arrange(Start)

# Filter out rows with Size equal to 0
clusters_data <- subset(clusters_data, Size > 100000) # show only bigger than 100kb


# Calculate the maximum size for scaling
max_size <- max(clusters_data$Size)

# Create Start_Mb and End_Mb within clusters_data
clusters_data$Start_Mb <- clusters_data$Start / 1e6
clusters_data$End_Mb <- clusters_data$End / 1e6

# Create the plot
ggplot(clusters_data, aes(xmin = Start_Mb, xmax = End_Mb, ymin = 0, ymax = Size / 1e6)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = label_number(unit = "M"), breaks = pretty_breaks(n = 10)) +
  scale_y_continuous(labels = label_number(unit = "M"), breaks = pretty_breaks(n = 10)) +
  labs(x = "Chromosome 1 (Mb)", y = "Cluster Size (Mb)", fill = "Cluster ID") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    panel.grid.minor.x = element_blank(),
    legend.position = "top",
    legend.box = "horizontal"
  ) +
  guides(fill = guide_legend(ncol = 7, title.position = "top", title.hjust = 0.5))


# Use ggsave to save the plot as a PDF
# ggsave(
#   filename = here("output", "ldna", "pop", "chr1", "clusters", "AUT_chr1.pdf"),
#   device = "pdf",
#   width = 8,
#   height = 5,
#   units = "in"
# )
```


We can annotate the number of segments in the legend.

```{r aut-chr1-plot-clusters-annotated, fig.height=5, fig.width=10}
# Merge SNP counts into clusters_data while specifying suffixes
annotated_data <- merge(clusters_data, segment_counts, by = "Cluster")

# Create a new column with Cluster names and SNP counts
annotated_data$Cluster_with_SNPs <- with(annotated_data, paste(Cluster, " (", nSegments, ")", sep=""))

# Now the plot uses the new Cluster_with_SNPs for the fill legend
ggplot(annotated_data, aes(xmin = Start_Mb, xmax = End_Mb, ymin = 0, ymax = Size / 1e6)) +
  geom_rect(aes(fill = as.factor(Cluster_with_SNPs)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = scales::label_number(unit = "M"), breaks = scales::pretty_breaks(n = 10)) +
  scale_y_continuous(labels = scales::label_number(unit = "M"), breaks = scales::pretty_breaks(n = 10)) +
  labs(x = "Chromosome 1 (Mb)", y = "Cluster Size (Mb)", fill = "Cluster ID") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    panel.grid.minor.x = element_blank(),
    legend.position = "top",
    legend.box = "horizontal"
  ) +
  guides(fill = guide_legend(nrow = 6, title.position = "top", title.hjust = 0.5))

# Use ggsave to save the plot as a PDF
ggsave(
  filename = here("output", "ldna", "pop", "chr1", "clusters", "AUT_chr1.pdf"),
  device = "pdf",
  width = 10,
  height = 5,
  units = "in"
)
```

Save plotting data, we can arrange the column order first
```{r aut-chr1-save-plot-data}
plot_data <- annotated_data |>
  dplyr::select(
    Cluster, nSegments, Start, End, Size, Start_Mb, End_Mb
  )

# Create a new column with Chromosome
plot_data <- data.frame(Chromosome = rep(1, nrow(plot_data)), plot_data)

# Split the 'Cluster' column at "_", and create the 'r2' column with the second part
plot_data$r2 <- sapply(strsplit(as.character(plot_data$Cluster), "_"), `[`, 2)

# Move the 'r2' column to be the second column, right after 'Chromosome'
plot_data <- plot_data[, c(1:2, ncol(plot_data), 3:(ncol(plot_data)-1))]


# Arrange
plot_data <- plot_data |>
  dplyr::arrange(Start)

# Check it
head(plot_data)

# Save it
saveRDS(plot_data, file = here("output", "ldna", "pop", "chr1", "AUT_plot.rds"))
```
We can also import the summary file to create a object with SNP counts per segment
```{r aut-chr1-build-cluster-seg-snp}
# Define the path
input_path <-
  here("output",
       "ldna",
       "pop",
       "chr1",
       "summary_AUT.txt")

# Read the data into R
clusters_snps <- read.table(input_path, header = TRUE, sep = "\t")

# We subset and rename the column 1
clusters_snps <- clusters_snps |>
  dplyr::select(
    Name, nLoci
  ) |>
  dplyr::rename(
    Cluster = Name,
    nSNPs = nLoci
  )

# Merge
cluster_seg_snp <- inner_join(segment_counts, clusters_snps, by = "Cluster")
cluster_seg_snp <- inner_join(clusters_data, cluster_seg_snp, by = "Cluster")

# Split the 'Cluster' column at "_", and create the 'r2' column with the second part
cluster_seg_snp$r2 <- sapply(strsplit(as.character(cluster_seg_snp$Cluster), "_"), `[`, 2)

# Add chromosome number and population name
cluster_seg_snp <- data.frame(Chromosome = rep(1, nrow(cluster_seg_snp)), cluster_seg_snp)
cluster_seg_snp <- data.frame(Population = rep("AUT", nrow(cluster_seg_snp)), cluster_seg_snp)


# Reorder the columns
cluster_seg_snp <- cluster_seg_snp |>
  dplyr::select(
    Chromosome, Cluster, r2, Start, End, nSegments, nSNPs
  ) |>
  dplyr::arrange(
    Chromosome, Start
  )

head(cluster_seg_snp)


# Save the data for plotting later
saveRDS(cluster_seg_snp, file = here("output", "ldna", "pop", "chr1", "AUT_plot2.rds"))
```


## 4. LDna chromosome 2

Because MAN has only 10 mosquitoes we will not estimate LD for it

### 4.1 NEW (chromosome 2)
Clean env and memory
```{r clean-env-new-chr2}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```

Import the data (make sure you have enough memory)
```{r new-chr2-load-ld-matrix, eval=TRUE}
ld1 <-
  read.delim(
    here("output", "ldna", "pop", "chr2", "NEW.chr2.txt"),
    sep = "\t",
    header = T,
    row.names = 1,
    stringsAsFactors = F,
    check.names = FALSE
  )
```

Check the LD matrix we built using bash
```{r new-chr2-check-ld-matrix, eval=TRUE}
head(ld1)
```

The row and column names are correct. Now we can convert the data frame to matrix

```{r new-chr2-convert-to-matrix, eval=TRUE}
# Create the matrix for ldna
ld2<- as.matrix(sapply(ld1, as.numeric))

# Get names
names<- rownames(ld1)
#names<-gsub("-", ".", names)

# Add names
row.names(ld2)<-names

# Clean non-numeric values
ld2[is.nan(ld2)] <- NA
ld2[is.infinite(ld2)] <- NA

# Remove diagonal values = 1
diag(ld2)=NA

# Make sure it is only the lower triangle matrix
ld2[!lower.tri(ld2)] <- NA
```

Create a LDna object
```{r new-chr2-run-ldna-raw, eval=TRUE}
# create the LDna object (change the number of cores as needed)
diag(ld2) <- NA
ld2[!lower.tri(ld2)] <- NA
ldna <- LDnaRaw(ld2, mc.cores=2, method = "single")
```

Save ldna
```{r new-chr2-save-ldna, eval=TRUE}
saveRDS(ldna, file = here("output", "ldna", "pop", "chr2", "NEW.rds"))
saveRDS(ld2, file = here("output", "ldna", "pop", "chr2", "NEW_ld2.rds"))
```

Import the data
```{r new-chr2-read-ldna}
ldna <- readRDS(here("output", "ldna", "pop", "chr2", "NEW.rds"))
ld2 <- readRDS(here("output", "ldna", "pop", "chr2", "NEW_ld2.rds"))
```


Check the object
```{r new-chr2-check-ldna-str}
str(ldna)
```

We can use the same parameters that we used for the other population
```{r new-chr2-set-edge-params}
edges_max <- 100
edges_min <- 20
```

We can test the number of edges in increments of 10 and use the default phi=2.
```{r new-chr2-extract-branches-loop}
pdf(file = here("output", "ldna", "pop", "chr2", paste(edges_min, "2.NEW_loop.edges.pdf", sep = "_")), width = 20, height = 12)
op <- par(mfcol=c(1,4))

# Adjust the step size here
step_size <- 10  # for example, try a smaller step like 10

for (edges_estimate in seq(edges_min, edges_max, by = step_size)) {
  clusters_count_prev <- 100
  clusters_count <- 50

  # The 'if' condition seems redundant here since clusters_count_prev is always set to 100 before the check
  if (clusters_count_prev > clusters_count) {
    print(edges_estimate) # This prints the current edges_estimate, showing that the loop is working
    clusters_count_prev <- clusters_count

    # Extract clusters
    clusters <- extractBranches(ldna, min.edges=edges_estimate)

    # Summarize the clusters
    summary <- summaryLDna(ldna, clusters, ld2)

    # Write the summary to a file
    write.table(summary, file = here("output", "ldna", "pop", "chr2", "summary_NEW.txt"), row.names=FALSE, sep="\t", quote=FALSE)

    # Update clusters_count with the number of clusters just written
    clusters_count <- nrow(read.delim(here("output", "ldna", "pop", "chr2", "summary_NEW.txt")))
  }
}

# Reset graphical parameters and close all open graphic devices
par(op)
while (!is.null(dev.list())) dev.off()
```

Get cluster information
```{r new-chr2-get-cluster-info}
# to get the list of names of the clusters
# Find the length of the longest vector
max_length <- max(sapply(clusters, length))

# Function to pad vectors with NAs to make them the same length
pad_vector_to_max_length <- function(vec, max_length) {
  c(vec, rep(NA, max_length - length(vec)))
}

# Apply the padding function to each vector in the list and combine into a data frame
do1 <- as.data.frame(do.call(cbind, lapply(clusters, pad_vector_to_max_length, max_length)))

# to get list of snps for each cluster
do <- do.call(cbind, unname(lapply(clusters, `length<-`, max(lengths(clusters)))))
colnames(do) <- colnames(do1)
```

Save it
```{r new-chr2-build-cluster-df}
max_length <- max(sapply(clusters, length))
# Initialize an empty list
cluster_df_list <- list()

# Loop through each cluster and pad with NAs
for (cluster_name in names(clusters)) {
  cluster_length <- length(clusters[[cluster_name]])
  padded_cluster <- c(clusters[[cluster_name]], rep(NA, max_length - cluster_length))
  cluster_df_list[[cluster_name]] <- padded_cluster
}

# Combine the lists into a data frame
df <- data.frame(cluster_df_list)
# Remove 'X' from column names
names(df) <- make.names(names(df), unique = FALSE)
names(df) <- sub("X", "", names(df), fixed = TRUE)

# Save it
saveRDS(df, file = here("output", "ldna", "pop", "chr2", "AUT_clusters_snps.rds"))
head(df)
```
We can save it in a different format as well
```{r new-chr2-clusters-long-format}
# Assuming your data frame is named df
long_df <- gather(df, Cluster, SNP, everything())

# View the first few rows of the transformed data
head(long_df)

# Save it
saveRDS(long_df, file = here("output", "ldna", "pop", "chr2", "AUT_clusters_snps1.rds"))
```

Make long format
```{r new-chr2-melt-clusters}
new_chr2 <- melt(do, na.rm = T, value.name = "value")
head(new_chr2)
```


Update names
```{r new-chr2-update-colnames}
colnames(new_chr2)<- c("v1", "cluster", "SNP")
head(new_chr2)
```



Import the bim file to get the SNP positions
```{r new-chr2-import-bim-positions}
# Import the data
new_snps_chr2 <- import_bim(here("output", "ldna", "pop", "NEW.bim")) |>
  dplyr::select(
    SNP, Scaffold, Position
  ) |>
  dplyr::rename(
    Chromosome = Scaffold
  )

# Check it
head(new_snps_chr2)
```

Merge the cluster data and the SNP data
```{r new-chr2-merge-clusters-positions}
# merge dataframes
merged<- merge(new_chr2, new_snps_chr2, by="SNP", all.x=TRUE)
merged<-na.omit(merged)
head(merged)
```

Select the columns we need
```{r new-chr2-select-cluster-cols}
# subset
merged <- subset(merged, select = c(cluster, SNP, Position))
head(merged)
```

Sort by distance
```{r new-chr2-sort-by-position}
merged<- merged[order(merged$Position),,drop=FALSE]
head(merged)
```


```{r new-chr2-prep-sushi-data}
# for parsing later
sushi1<- subset(merged, select = c(cluster, Position))
head(sushi1)
```

Save the data
```{r new-chr2-save-cluster-files, warning=FALSE}
# Create directory
new_directory_path <- here("output", "ldna", "pop", "chr2", "clusters")
dir.create(new_directory_path, recursive = TRUE)

# save file
write.table(merged, file = here("output", "ldna", "pop", "chr2", "SNPs_clusters_NEW_chr2.txt"), row.names = F, sep = "\t", quote = F)
write.table(sushi1, file = here("output", "ldna", "pop", "chr2", "clusters", "sushi_NEW_chr2.txt"), row.names = F, sep = "\t", quote = F)
```

Parse the file1 to get the start and end of the clusters as well as their size. Remember, recombination make it a mosaic, so the clusters are "mixed".
```{bash}
awk '
function print_row() {
    if ( feature != "" )
       print feature, start, end, (end - start)
}

BEGIN {
    FS=OFS="\t";
    print "Cluster", "Start", "End", "Size";
}
NR == 1 { next } # Skip the first line
$1 != feature {
    print_row();
    feature = $1;
    start = $2;
    end = $2;
    next;
}
{
    end = $2;
}
END {
    print_row();
}
' output/ldna/pop/chr2/clusters/sushi_NEW_chr2.txt > output/ldna/pop/chr2/clusters/sushi2_NEW_chr2.txt;
head output/ldna/pop/chr2/clusters/sushi2_NEW_chr2.txt
```

```{r new-chr2-count-cluster-segments}
# Define the path
input_path <- here("output", "ldna", "pop", "chr2", "clusters", "sushi2_NEW_chr2.txt")

# Read the data into R
clusters_data <- read_table(input_path, col_names = TRUE, col_types = NULL)

# Arrange by position
clusters_data <- clusters_data |>
  arrange(Start)


# Count segments for each cluster - if you want only the counts
segment_counts <- clusters_data %>%
  group_by(Cluster) %>%
  summarize(nSegments = n(), .groups = 'drop')

# View the result
head(segment_counts)
```


Plot it
```{r new-chr2-plot-clusters-basic, fig.height=5, fig.width=8}
# Define the path
input_path <-
  here("output",
       "ldna",
       "pop",
       "chr2",
       "clusters",
       "sushi2_NEW_chr2.txt")

# Read the data into R
clusters_data <- read.table(input_path, header = TRUE, sep = "\t") |>
  arrange(Start)

# Filter out rows with Size equal to 0
clusters_data <- subset(clusters_data, Size > 100000) # show only bigger than 100kb


# Calculate the maximum size for scaling
max_size <- max(clusters_data$Size)

# Create Start_Mb and End_Mb within clusters_data
clusters_data$Start_Mb <- clusters_data$Start / 1e6
clusters_data$End_Mb <- clusters_data$End / 1e6

# Create the plot
ggplot(clusters_data, aes(xmin = Start_Mb, xmax = End_Mb, ymin = 0, ymax = Size / 1e6)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = label_number(unit = "M"), breaks = pretty_breaks(n = 10)) +
  scale_y_continuous(labels = label_number(unit = "M"), breaks = pretty_breaks(n = 10)) +
  labs(x = "Chromosome 2 (Mb)", y = "Cluster Size (Mb)", fill = "Cluster ID") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    panel.grid.minor.x = element_blank(),
    legend.position = "top",
    legend.box = "horizontal"
  ) +
  guides(fill = guide_legend(nrow = 3, title.position = "top", title.hjust = 0.5))


# Use ggsave to save the plot as a PDF
# ggsave(
#   filename = here("output", "ldna", "pop", "chr2", "clusters", "NEW_chr2.pdf"),
#   device = "pdf",
#   width = 8,
#   height = 5,
#   units = "in"
# )
```


We can annotate the number of segments in the legend.

```{r new-chr2-plot-clusters-annotated, fig.height=5, fig.width=8}
# Merge SNP counts into clusters_data while specifying suffixes
annotated_data <- merge(clusters_data, segment_counts, by = "Cluster")

# Create a new column with Cluster names and SNP counts
annotated_data$Cluster_with_SNPs <- with(annotated_data, paste(Cluster, " (", nSegments, ")", sep=""))

# Now the plot uses the new Cluster_with_SNPs for the fill legend
ggplot(annotated_data, aes(xmin = Start_Mb, xmax = End_Mb, ymin = 0, ymax = Size / 1e6)) +
  geom_rect(aes(fill = as.factor(Cluster_with_SNPs)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = scales::label_number(unit = "M"), breaks = scales::pretty_breaks(n = 10)) +
  scale_y_continuous(labels = scales::label_number(unit = "M"), breaks = scales::pretty_breaks(n = 10)) +
  labs(x = "Chromosome 2 (Mb)", y = "Cluster Size (Mb)", fill = "Cluster ID") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    panel.grid.minor.x = element_blank(),
    legend.position = "top",
    legend.box = "horizontal"
  ) +
  guides(fill = guide_legend(nrow = 3, title.position = "top", title.hjust = 0.5))

# Use ggsave to save the plot as a PDF
ggsave(
  filename = here("output", "ldna", "pop", "chr2", "clusters", "NEW_chr2.pdf"),
  device = "pdf",
  width = 8,
  height = 5,
  units = "in"
)
```
Check how many SNPs per cluster
```{r new-chr2-check-snp-count}
# Count non-NA values in each column
non_na_counts <- colSums(!is.na(df))

# View the counts
non_na_counts
```
Check if it matches with the output from LDna
```{r new-chr2-check-summary}
# Replace 'path_to_file.txt' with the actual path to your text file
snp_data <- read.table(here("output", "ldna", "pop", "chr2", "summary_NEW.txt"), header = TRUE, sep = "\t")

# View the first few rows of the data
head(snp_data)
```
It does match.

We can save the data to plot all chromosomes together using a facet plot
```{r new-chr2-check-annotated-data}
head(annotated_data)
```

Save plotting data, we can arrange the column order first
```{r new-chr2-save-plot-data}
plot_data <- annotated_data |>
  dplyr::select(
    Cluster, nSegments, Start, End, Size, Start_Mb, End_Mb
  )

# Create a new column with Chromosome
plot_data <- data.frame(Chromosome = rep(1, nrow(plot_data)), plot_data)

# Split the 'Cluster' column at "_", and create the 'r2' column with the second part
plot_data$r2 <- sapply(strsplit(as.character(plot_data$Cluster), "_"), `[`, 2)

# Move the 'r2' column to be the second column, right after 'Chromosome'
plot_data <- plot_data[, c(1:2, ncol(plot_data), 3:(ncol(plot_data)-1))]


# Arrange
plot_data <- plot_data |>
  dplyr::arrange(Start)

# Check it
head(plot_data)

# Save it
saveRDS(plot_data, file = here("output", "ldna", "pop", "chr2", "NEW_plot.rds"))
```

We can also import the summary file to create a object with SNP counts per segment
```{r new-chr2-build-cluster-seg-snp}
# Define the path
input_path <-
  here("output",
       "ldna",
       "pop",
       "chr2",
       "summary_NEW.txt")

# Read the data into R
clusters_snps <- read.table(input_path, header = TRUE, sep = "\t")

# We subset and rename the column 1
clusters_snps <- clusters_snps |>
  dplyr::select(
    Name, nLoci
  ) |>
  dplyr::rename(
    Cluster = Name,
    nSNPs = nLoci
  )

# Merge
cluster_seg_snp <- inner_join(segment_counts, clusters_snps, by = "Cluster")
cluster_seg_snp <- inner_join(clusters_data, cluster_seg_snp, by = "Cluster")

# Split the 'Cluster' column at "_", and create the 'r2' column with the second part
cluster_seg_snp$r2 <- sapply(strsplit(as.character(cluster_seg_snp$Cluster), "_"), `[`, 2)

# Add chromosome number and population name
cluster_seg_snp <- data.frame(Chromosome = rep(2, nrow(cluster_seg_snp)), cluster_seg_snp)
cluster_seg_snp <- data.frame(Population = rep("NEW", nrow(cluster_seg_snp)), cluster_seg_snp)


# Reorder the columns
cluster_seg_snp <- cluster_seg_snp |>
  dplyr::select(
    Chromosome, Cluster, r2, Start, End, nSegments, nSNPs
  ) |>
  dplyr::arrange(
    Chromosome, Start
  )

head(cluster_seg_snp)


# Save the data for plotting later
saveRDS(cluster_seg_snp, file = here("output", "ldna", "pop", "chr2", "NEW_plot2.rds"))
```


### 4.2 AUT (chromosome 2)
Clean env and memory
```{r clean-env-aut-chr2, eval=TRUE}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```

Import the data
```{r aut-chr2-load-ld-matrix, eval=TRUE}
ld1 <-
  read.delim(
    here("output", "ldna", "pop", "chr2", "AUT.chr2.txt"),
    sep = "\t",
    header = T,
    row.names = 1,
    stringsAsFactors = F,
    check.names = FALSE
  )
```

Check the LD matrix we built using bash
```{r aut-chr2-check-ld-matrix, eval=TRUE}
head(ld1)
```

The row and column names are correct. Now we can convert the data frame to matrix

```{r aut-chr2-convert-to-matrix, eval=TRUE}
# Create the matrix for ldna
ld2<- as.matrix(sapply(ld1, as.numeric))

# Get names
names<- rownames(ld1)
#names<-gsub("-", ".", names)

# Add names
row.names(ld2)<-names

# Clean non-numeric values
ld2[is.nan(ld2)] <- NA
ld2[is.infinite(ld2)] <- NA

# Remove diagonal values = 1
diag(ld2)=NA

# Make sure it is only the lower triangle matrix
ld2[!lower.tri(ld2)] <- NA
```

Create a LDna object
```{r aut-chr2-run-ldna-raw, eval=TRUE}
# create the LDna object (change the number of cores as needed)
diag(ld2) <- NA
ld2[!lower.tri(ld2)] <- NA
ldna <- LDnaRaw(ld2, mc.cores=2, method = "single")
```

Check the object
```{r aut-chr2-check-ldna-str, eval=TRUE}
str(ldna)
```

Save it
```{r aut-chr2-save-ldna, eval=TRUE}
saveRDS(ldna, file = here("output", "ldna", "pop", "chr2", "AUT.rds"))
saveRDS(ld2, file = here("output", "ldna", "pop", "chr2", "AUT_ld2.rds"))
```

Load data
```{r aut-chr2-read-ldna}
ldna <- readRDS(here("output", "ldna", "pop", "chr2", "AUT.rds"))
ld2 <- readRDS(here("output", "ldna", "pop", "chr2", "AUT_ld2.rds"))
```

We can use the same parameters that we used for the other population
```{r aut-chr2-set-edge-params}
# snp_count <- ncol(ld2)
edges_max <- 100
edges_min <- 20
```

We can test the number of edges in increments of 10 and use the default phi=2
```{r aut-chr2-extract-branches-loop}
pdf(file = here("output", "ldna", "pop", "chr2", paste(edges_min, "2.AUT_loop.edges.pdf", sep = "_")), width = 20, height = 12)
op <- par(mfcol=c(1,4))

# Adjust the step size here
step_size <- 10  # for example, try a smaller step like 10

for (edges_estimate in seq(edges_min, edges_max, by = step_size)) {
  clusters_count_prev <- 100
  clusters_count <- 50

  # The 'if' condition seems redundant here since clusters_count_prev is always set to 100 before the check
  if (clusters_count_prev > clusters_count) {
    print(edges_estimate) # This prints the current edges_estimate, showing that the loop is working
    clusters_count_prev <- clusters_count

    # Extract clusters
    clusters <- extractBranches(ldna, min.edges=edges_estimate)

    # Summarize the clusters
    summary <- summaryLDna(ldna, clusters, ld2)

    # Write the summary to a file
    write.table(summary, file = here("output", "ldna", "pop", "chr2", "summary_AUT.txt"), row.names=FALSE, sep="\t", quote=FALSE)

    # Update clusters_count with the number of clusters just written
    clusters_count <- nrow(read.delim(here("output", "ldna", "pop", "chr2", "summary_AUT.txt")))
  }
}

# Reset graphical parameters and close all open graphic devices
par(op)
while (!is.null(dev.list())) dev.off()
```

Get cluster information
```{r aut-chr2-get-cluster-info}
# to get the list of names of the clusters
# Find the length of the longest vector
max_length <- max(sapply(clusters, length))

# Function to pad vectors with NAs to make them the same length
pad_vector_to_max_length <- function(vec, max_length) {
  c(vec, rep(NA, max_length - length(vec)))
}

# Apply the padding function to each vector in the list and combine into a data frame
do1 <- as.data.frame(do.call(cbind, lapply(clusters, pad_vector_to_max_length, max_length)))

# to get list of snps for each cluster
do <- do.call(cbind, unname(lapply(clusters, `length<-`, max(lengths(clusters)))))
colnames(do) <- colnames(do1)
```

```{r aut-chr2-build-cluster-df}
max_length <- max(sapply(clusters, length))
# Initialize an empty list
cluster_df_list <- list()

# Loop through each cluster and pad with NAs
for (cluster_name in names(clusters)) {
  cluster_length <- length(clusters[[cluster_name]])
  padded_cluster <- c(clusters[[cluster_name]], rep(NA, max_length - cluster_length))
  cluster_df_list[[cluster_name]] <- padded_cluster
}

# Combine the lists into a data frame
df <- data.frame(cluster_df_list)
# Remove 'X' from column names
names(df) <- make.names(names(df), unique = FALSE)
names(df) <- sub("X", "", names(df), fixed = TRUE)

# Save it
saveRDS(df, file = here("output", "ldna", "pop", "chr2", "AUT_clusters_snps.rds"))
head(df)
```
We can save it in a different format as well
```{r aut-chr2-clusters-long-format}
# Assuming your data frame is named df
long_df <- gather(df, Cluster, SNP, everything())

# View the first few rows of the transformed data
head(long_df)

# Save it
saveRDS(long_df, file = here("output", "ldna", "pop", "chr2", "AUT_clusters_snps1.rds"))
```

Make long format
```{r aut-chr2-melt-clusters}
aut_chr2 <- melt(do, na.rm = T, value.name = "value")
head(aut_chr2)
```


Update names
```{r aut-chr2-update-colnames}
colnames(aut_chr2)<- c("v1", "cluster", "SNP")
head(aut_chr2)
```


Import the bim file to get the SNP positions
```{r aut-chr2-import-bim-positions}
# Import the data
aut_snps_chr <- import_bim(here("output", "ldna", "pop", "AUT.bim")) |>
  dplyr::select(
    SNP, Scaffold, Position
  ) |>
  dplyr::rename(
    Chromosome = Scaffold
  )

# Check it
head(aut_snps_chr)
```

Merge the cluster data and the SNP data
```{r aut-chr2-merge-clusters-positions}
# merge dataframes
merged<- merge(aut_chr2, aut_snps_chr, by="SNP", all.x=TRUE)
merged<-na.omit(merged)
head(merged)
```

Select the columns we need
```{r aut-chr2-select-cluster-cols}
# subset
merged <- subset(merged, select = c(cluster, SNP, Position))
head(merged)
```

Sort by distance
```{r aut-chr2-sort-by-position}
merged<- merged[order(merged$Position),,drop=FALSE]
head(merged)
```


```{r aut-chr2-prep-sushi-data}
# for parsing later
sushi1<- subset(merged, select = c(cluster, Position))
head(sushi1)
```

Save the data
```{r aut-chr2-save-cluster-files, warning=FALSE}
# Create directory
new_directory_path <- here("output", "ldna", "pop", "chr2", "clusters")
dir.create(new_directory_path, recursive = TRUE)

# save file
write.table(merged, file = here("output", "ldna", "pop", "chr2", "SNPs_clusters_AUT_chr2.txt"), row.names = F, sep = "\t", quote = F)
write.table(sushi1, file = here("output", "ldna", "pop", "chr2", "clusters", "sushi_AUT_chr2.txt"), row.names = F, sep = "\t", quote = F)
```

Parse the file1 to get the start and end of the clusters as well as their size. Remember, recombination make it a mosaic, so the clusters are "mixed".
```{bash}
awk '
function print_row() {
    if ( feature != "" )
       print feature, start, end, (end - start)
}

BEGIN {
    FS=OFS="\t";
    print "Cluster", "Start", "End", "Size";
}
NR == 1 { next } # Skip the first line
$1 != feature {
    print_row();
    feature = $1;
    start = $2;
    end = $2;
    next;
}
{
    end = $2;
}
END {
    print_row();
}
' output/ldna/pop/chr2/clusters/sushi_AUT_chr2.txt > output/ldna/pop/chr2/clusters/sushi2_AUT_chr2.txt;
head output/ldna/pop/chr2/clusters/sushi2_AUT_chr2.txt
```

Get SNP count
```{r aut-chr2-count-cluster-segments}
# Define the path
input_path <- here("output", "ldna", "pop", "chr2", "clusters", "sushi2_AUT_chr2.txt")

# Read the data into R
clusters_data <- read_table(input_path, col_names = TRUE, col_types = NULL)

# Arrange by position
clusters_data <- clusters_data |>
  arrange(Start)

# Count segments for each cluster - if you want only the counts
segment_counts <- clusters_data %>%
  group_by(Cluster) %>%
  summarize(nSegments = n(), .groups = 'drop')

# View the result
head(segment_counts)
```

Plot it
```{r aut-chr2-plot-clusters-basic, fig.height=5, fig.width=8}
# Define the path
input_path <-
  here("output",
       "ldna",
       "pop",
       "chr2",
       "clusters",
       "sushi2_AUT_chr2.txt")

# Read the data into R
clusters_data <- read.table(input_path, header = TRUE, sep = "\t") |>
  arrange(Start)

# Arrange by position
clusters_data <- clusters_data |>
  arrange(Start)

# Filter out rows with Size equal to 0
clusters_data <- subset(clusters_data, Size > 100000) # show only bigger than 100kb


# Calculate the maximum size for scaling
max_size <- max(clusters_data$Size)

# Create Start_Mb and End_Mb within clusters_data
clusters_data$Start_Mb <- clusters_data$Start / 1e6
clusters_data$End_Mb <- clusters_data$End / 1e6

# Create the plot
ggplot(clusters_data, aes(xmin = Start_Mb, xmax = End_Mb, ymin = 0, ymax = Size / 1e6)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = label_number(unit = "M"), breaks = pretty_breaks(n = 10)) +
  scale_y_continuous(labels = label_number(unit = "M"), breaks = pretty_breaks(n = 10)) +
  labs(x = "Chromosome 2 (Mb)", y = "Cluster Size (Mb)", fill = "Cluster ID") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    panel.grid.minor.x = element_blank(),
    legend.position = "top",
    legend.box = "horizontal"
  ) +
  guides(fill = guide_legend(ncol = 7, title.position = "top", title.hjust = 0.5))


# Use ggsave to save the plot as a PDF
ggsave(
  filename = here("output", "ldna", "pop", "chr2", "clusters", "AUT_chr2.pdf"),
  device = "pdf",
  width = 8,
  height = 5,
  units = "in"
)
```


We can annotate the number of segments in the legend.

```{r aut-chr2-plot-clusters-annotated, fig.height=5, fig.width=10}
# Merge SNP counts into clusters_data while specifying suffixes
annotated_data <- merge(clusters_data, segment_counts, by = "Cluster")

# Create a new column with Cluster names and SNP counts
annotated_data$Cluster_with_SNPs <- with(annotated_data, paste(Cluster, " (", nSegments, ")", sep=""))

# Now the plot uses the new Cluster_with_SNPs for the fill legend
ggplot(annotated_data, aes(xmin = Start_Mb, xmax = End_Mb, ymin = 0, ymax = Size / 1e6)) +
  geom_rect(aes(fill = as.factor(Cluster_with_SNPs)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = scales::label_number(unit = "M"), breaks = scales::pretty_breaks(n = 10)) +
  scale_y_continuous(labels = scales::label_number(unit = "M"), breaks = scales::pretty_breaks(n = 10)) +
  labs(x = "Chromosome 2 (Mb)", y = "Cluster Size (Mb)", fill = "Cluster ID") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    panel.grid.minor.x = element_blank(),
    legend.position = "top",
    legend.box = "horizontal"
  ) +
  guides(fill = guide_legend(nrow = 11, title.position = "top", title.hjust = 0.5))

# Use ggsave to save the plot as a PDF
ggsave(
  filename = here("output", "ldna", "pop", "chr2", "clusters", "AUT_chr2b.pdf"),
  device = "pdf",
  width = 10,
  height = 5,
  units = "in"
)
```

Save plotting data, we can arrange the column order first
```{r aut-chr2-save-plot-data}
plot_data <- annotated_data |>
  dplyr::select(
    Cluster, nSegments, Start, End, Size, Start_Mb, End_Mb
  )

# Create a new column with Chromosome
plot_data <- data.frame(Chromosome = rep(1, nrow(plot_data)), plot_data)

# Split the 'Cluster' column at "_", and create the 'r2' column with the second part
plot_data$r2 <- sapply(strsplit(as.character(plot_data$Cluster), "_"), `[`, 2)

# Move the 'r2' column to be the second column, right after 'Chromosome'
plot_data <- plot_data[, c(1:2, ncol(plot_data), 3:(ncol(plot_data)-1))]


# Arrange
plot_data <- plot_data |>
  dplyr::arrange(Start)

# Check it
head(plot_data)

# Save it
saveRDS(plot_data, file = here("output", "ldna", "pop", "chr2", "AUT_plot.rds"))
```

We can also import the summary file to create a object with SNP counts per segment
```{r aut-chr2-build-cluster-seg-snp}
# Define the path
input_path <-
  here("output",
       "ldna",
       "pop",
       "chr2",
       "summary_AUT.txt")

# Read the data into R
clusters_snps <- read.table(input_path, header = TRUE, sep = "\t")

# We subset and rename the column 1
clusters_snps <- clusters_snps |>
  dplyr::select(
    Name, nLoci
  ) |>
  dplyr::rename(
    Cluster = Name,
    nSNPs = nLoci
  )

# Merge
cluster_seg_snp <- inner_join(segment_counts, clusters_snps, by = "Cluster")
cluster_seg_snp <- inner_join(clusters_data, cluster_seg_snp, by = "Cluster")

# Split the 'Cluster' column at "_", and create the 'r2' column with the second part
cluster_seg_snp$r2 <- sapply(strsplit(as.character(cluster_seg_snp$Cluster), "_"), `[`, 2)

# Add chromosome number and population name
cluster_seg_snp <- data.frame(Chromosome = rep(2, nrow(cluster_seg_snp)), cluster_seg_snp)
cluster_seg_snp <- data.frame(Population = rep("AUT", nrow(cluster_seg_snp)), cluster_seg_snp)


# Reorder the columns
cluster_seg_snp <- cluster_seg_snp |>
  dplyr::select(
    Chromosome, Cluster, r2, Start, End, nSegments, nSNPs
  ) |>
  dplyr::arrange(
    Chromosome, Start
  )

head(cluster_seg_snp)


# Save the data for plotting later
saveRDS(cluster_seg_snp, file = here("output", "ldna", "pop", "chr2", "AUT_plot2.rds"))
```


## 5. LDna chromosome 3

Because MAN has only 10 mosquitoes we will not estimate LD for it

### 5.1 NEW (chromosome 3)
Clean env and memory
```{r clean-env-new-chr3}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```

Import the data (make sure you have enough memory)
```{r new-chr3-load-ld-matrix, eval=TRUE}
ld1 <-
  read.delim(
    here("output", "ldna", "pop", "chr3", "NEW.chr3.txt"),
    sep = "\t",
    header = T,
    row.names = 1,
    stringsAsFactors = F,
    check.names = FALSE
  )
```

Check the LD matrix we built using bash
```{r new-chr3-check-ld-matrix, eval=TRUE}
head(ld1)
```

The row and column names are correct. Now we can convert the data frame to matrix

```{r new-chr3-convert-to-matrix, eval=TRUE}
# Create the matrix for ldna
ld2<- as.matrix(sapply(ld1, as.numeric))

# Get names
names<- rownames(ld1)
#names<-gsub("-", ".", names)

# Add names
row.names(ld2)<-names

# Clean non-numeric values
ld2[is.nan(ld2)] <- NA
ld2[is.infinite(ld2)] <- NA

# Remove diagonal values = 1
diag(ld2)=NA

# Make sure it is only the lower triangle matrix
ld2[!lower.tri(ld2)] <- NA
```

Create a LDna object
```{r new-chr3-run-ldna-raw, eval=TRUE}
# create the LDna object (change the number of cores as needed)
diag(ld2) <- NA
ld2[!lower.tri(ld2)] <- NA
ldna <- LDnaRaw(ld2, mc.cores=2, method = "single")
```

Save ldna
```{r new-chr3-save-ldna, eval=TRUE}
saveRDS(ldna, file = here("output", "ldna", "pop", "chr3", "NEW.rds"))
saveRDS(ld2, file = here("output", "ldna", "pop", "chr3", "NEW_ld2.rds"))
```

```{r new-chr3-read-ldna}
ldna <- readRDS(here("output", "ldna", "pop", "chr3", "NEW.rds"))
ld2 <- readRDS(here("output", "ldna", "pop", "chr3", "NEW_ld2.rds"))
```


Check the object
```{r new-chr3-check-ldna-str}
str(ldna)
```

We can use the same parameters that we used for the other population
```{r new-chr3-set-edge-params}
# snp_count <- ncol(ld2)
edges_max <- 100
edges_min <- 20
```

We can test the number of edges in increments of 10 and use the default phi=2.
```{r new-chr3-extract-branches-loop}
pdf(file = here("output", "ldna", "pop", "chr3", paste(edges_min, "3.NEW_loop.edges.pdf", sep = "_")), width = 20, height = 12)
op <- par(mfcol=c(1,4))

# Adjust the step size here
step_size <- 10  # for example, try a smaller step like 10

for (edges_estimate in seq(edges_min, edges_max, by = step_size)) {
  clusters_count_prev <- 100
  clusters_count <- 50

  # The 'if' condition seems redundant here since clusters_count_prev is always set to 100 before the check
  if (clusters_count_prev > clusters_count) {
    print(edges_estimate) # This prints the current edges_estimate, showing that the loop is working
    clusters_count_prev <- clusters_count

    # Extract clusters
    clusters <- extractBranches(ldna, min.edges=edges_estimate)

    # Summarize the clusters
    summary <- summaryLDna(ldna, clusters, ld2)

    # Write the summary to a file
    write.table(summary, file = here("output", "ldna", "pop", "chr3", "summary_NEW.txt"), row.names=FALSE, sep="\t", quote=FALSE)

    # Update clusters_count with the number of clusters just written
    clusters_count <- nrow(read.delim(here("output", "ldna", "pop", "chr3", "summary_NEW.txt")))
  }
}

# Reset graphical parameters and close all open graphic devices
par(op)
while (!is.null(dev.list())) dev.off()
```

Get cluster information
```{r new-chr3-get-cluster-info}
# to get the list of names of the clusters
# Find the length of the longest vector
max_length <- max(sapply(clusters, length))

# Function to pad vectors with NAs to make them the same length
pad_vector_to_max_length <- function(vec, max_length) {
  c(vec, rep(NA, max_length - length(vec)))
}

# Apply the padding function to each vector in the list and combine into a data frame
do1 <- as.data.frame(do.call(cbind, lapply(clusters, pad_vector_to_max_length, max_length)))

# to get list of snps for each cluster
do <- do.call(cbind, unname(lapply(clusters, `length<-`, max(lengths(clusters)))))
colnames(do) <- colnames(do1)
```

Save it
```{r new-chr3-build-cluster-df}
max_length <- max(sapply(clusters, length))
# Initialize an empty list
cluster_df_list <- list()

# Loop through each cluster and pad with NAs
for (cluster_name in names(clusters)) {
  cluster_length <- length(clusters[[cluster_name]])
  padded_cluster <- c(clusters[[cluster_name]], rep(NA, max_length - cluster_length))
  cluster_df_list[[cluster_name]] <- padded_cluster
}

# Combine the lists into a data frame
df <- data.frame(cluster_df_list)
# Remove 'X' from column names
names(df) <- make.names(names(df), unique = FALSE)
names(df) <- sub("X", "", names(df), fixed = TRUE)

# Save it
saveRDS(df, file = here("output", "ldna", "pop", "chr3", "NEW_clusters_snps.rds"))
head(df)
```
We can save it in a different format as well
```{r new-chr3-clusters-long-format}
# Assuming your data frame is named df
long_df <- gather(df, Cluster, SNP, everything())

# View the first few rows of the transformed data
head(long_df)

# Save it
saveRDS(long_df, file = here("output", "ldna", "pop", "chr3", "NEW_clusters_snps1.rds"))
```


Make long format
```{r new-chr3-melt-clusters}
new_chr3 <- melt(do, na.rm = T, value.name = "value")
head(new_chr3)
```


Update names
```{r new-chr3-update-colnames}
colnames(new_chr3)<- c("v1", "cluster", "SNP")
head(new_chr3)
```



Import the bim file to get the SNP positions
```{r new-chr3-import-bim-positions}
# Import the data
new_snps_chr <- import_bim(here("output", "ldna", "pop", "NEW.bim")) |>
  dplyr::select(
    SNP, Scaffold, Position
  ) |>
  dplyr::rename(
    Chromosome = Scaffold
  )

# Check it
head(new_snps_chr)
```

Merge the cluster data and the SNP data
```{r new-chr3-merge-clusters-positions}
# merge dataframes
merged<- merge(new_chr3, new_snps_chr, by="SNP", all.x=TRUE)
merged<-na.omit(merged)
head(merged)
```

Select the columns we need
```{r new-chr3-select-cluster-cols}
# subset
merged <- subset(merged, select = c(cluster, SNP, Position))
head(merged)
```

Sort by distance
```{r new-chr3-sort-by-position}
merged<- merged[order(merged$Position),,drop=FALSE]
head(merged)
```


```{r new-chr3-prep-sushi-data}
# for parsing later
sushi1<- subset(merged, select = c(cluster, Position))
head(sushi1)
```

Save the data
```{r new-chr3-save-cluster-files, warning=FALSE}
# Create directory
new_directory_path <- here("output", "ldna", "pop", "chr3", "clusters")
dir.create(new_directory_path, recursive = TRUE)

# save file
write.table(merged, file = here("output", "ldna", "pop", "chr3", "SNPs_clusters_NEW_chr3.txt"), row.names = F, sep = "\t", quote = F)
write.table(sushi1, file = here("output", "ldna", "pop", "chr3", "clusters", "sushi_NEW_chr3.txt"), row.names = F, sep = "\t", quote = F)
```

Parse the file1 to get the start and end of the clusters as well as their size. Remember, recombination make it a mosaic, so the clusters are "mixed".
```{bash}
awk '
function print_row() {
    if ( feature != "" )
       print feature, start, end, (end - start)
}

BEGIN {
    FS=OFS="\t";
    print "Cluster", "Start", "End", "Size";
}
NR == 1 { next } # Skip the first line
$1 != feature {
    print_row();
    feature = $1;
    start = $2;
    end = $2;
    next;
}
{
    end = $2;
}
END {
    print_row();
}
' output/ldna/pop/chr3/clusters/sushi_NEW_chr3.txt > output/ldna/pop/chr3/clusters/sushi2_NEW_chr3.txt;
head output/ldna/pop/chr3/clusters/sushi2_NEW_chr3.txt
```

```{r new-chr3-count-cluster-segments}
# Define the path
input_path <- here("output", "ldna", "pop", "chr3", "clusters", "sushi2_NEW_chr3.txt")

# Read the data into R
clusters_data <- read_table(input_path, col_names = TRUE, col_types = NULL)

# Arrange by position
clusters_data <- clusters_data |>
  arrange(Start)

# Count segments for each cluster - if you want only the counts
segment_counts <- clusters_data %>%
  group_by(Cluster) %>%
  summarize(nSegments = n(), .groups = 'drop')


# View the result
head(segment_counts)
```


Plot it
```{r new-chr3-plot-clusters-basic, fig.height=5, fig.width=8}
# Define the path
input_path <-
  here("output",
       "ldna",
       "pop",
       "chr3",
       "clusters",
       "sushi2_NEW_chr3.txt")

# Read the data into R
clusters_data <- read.table(input_path, header = TRUE, sep = "\t") |>
  arrange(Start)

# Filter out rows with Size equal to 0
clusters_data <- subset(clusters_data, Size > 100000) # show only bigger than 100kb


# Calculate the maximum size for scaling
max_size <- max(clusters_data$Size)

# Create Start_Mb and End_Mb within clusters_data
clusters_data$Start_Mb <- clusters_data$Start / 1e6
clusters_data$End_Mb <- clusters_data$End / 1e6

# Create the plot
ggplot(clusters_data, aes(xmin = Start_Mb, xmax = End_Mb, ymin = 0, ymax = Size / 1e6)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = label_number(unit = "M"), breaks = pretty_breaks(n = 10)) +
  scale_y_continuous(labels = label_number(unit = "M"), breaks = pretty_breaks(n = 10)) +
  labs(x = "Chromosome 3 (Mb)", y = "Cluster Size (Mb)", fill = "Cluster ID") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    panel.grid.minor.x = element_blank(),
    legend.position = "top",
    legend.box = "horizontal"
  ) +
  guides(fill = guide_legend(nrow = 3, title.position = "top", title.hjust = 0.5))


# Use ggsave to save the plot as a PDF
ggsave(
  filename = here("output", "ldna", "pop", "chr3", "clusters", "NEW_chr3.pdf"),
  device = "pdf",
  width = 8,
  height = 5,
  units = "in"
)
```


We can annotate the number of segments in the legend.

```{r new-chr3-plot-clusters-annotated, fig.height=5, fig.width=8}
# Merge SNP counts into clusters_data while specifying suffixes
annotated_data <- merge(clusters_data, segment_counts, by = "Cluster")

# Create a new column with Cluster names and SNP counts
annotated_data$Cluster_with_SNPs <- with(annotated_data, paste(Cluster, " (", nSegments, ")", sep=""))

# Now the plot uses the new Cluster_with_SNPs for the fill legend
ggplot(annotated_data, aes(xmin = Start_Mb, xmax = End_Mb, ymin = 0, ymax = Size / 1e6)) +
  geom_rect(aes(fill = as.factor(Cluster_with_SNPs)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = scales::label_number(unit = "M"), breaks = scales::pretty_breaks(n = 10)) +
  scale_y_continuous(labels = scales::label_number(unit = "M"), breaks = scales::pretty_breaks(n = 10)) +
  labs(x = "Chromosome 3 (Mb)", y = "Cluster Size (Mb)", fill = "Cluster ID") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    panel.grid.minor.x = element_blank(),
    legend.position = "top",
    legend.box = "horizontal"
  ) +
  guides(fill = guide_legend(nrow = 3, title.position = "top", title.hjust = 0.5))

# Use ggsave to save the plot as a PDF
ggsave(
  filename = here("output", "ldna", "pop", "chr3", "clusters", "NEW_chr3b.pdf"),
  device = "pdf",
  width = 8,
  height = 5,
  units = "in"
)
```



We can save the data to plot all chromosomes togeter using a facet plot
```{r new-chr3-check-annotated-data}
head(annotated_data)
```

Save plotting data, we can arrange the column order first
```{r new-chr3-save-plot-data}
plot_data <- annotated_data |>
  dplyr::select(
    Cluster, nSegments, Start, End, Size, Start_Mb, End_Mb
  )

# Create a new column with Chromosome
plot_data <- data.frame(Chromosome = rep(1, nrow(plot_data)), plot_data)

# Split the 'Cluster' column at "_", and create the 'r2' column with the second part
plot_data$r2 <- sapply(strsplit(as.character(plot_data$Cluster), "_"), `[`, 2)

# Move the 'r2' column to be the second column, right after 'Chromosome'
plot_data <- plot_data[, c(1:2, ncol(plot_data), 3:(ncol(plot_data)-1))]


# Arrange
plot_data <- plot_data |>
  dplyr::arrange(Start)

# Check it
head(plot_data)

# Save it
saveRDS(plot_data, file = here("output", "ldna", "pop", "chr3", "NEW_plot.rds"))
```

We can also import the summary file to create a object with SNP counts per segment
```{r new-chr3-build-cluster-seg-snp}
# Define the path
input_path <-
  here("output",
       "ldna",
       "pop",
       "chr3",
       "summary_NEW.txt")

# Read the data into R
clusters_snps <- read.table(input_path, header = TRUE, sep = "\t")

# We subset and rename the column 1
clusters_snps <- clusters_snps |>
  dplyr::select(
    Name, nLoci
  ) |>
  dplyr::rename(
    Cluster = Name,
    nSNPs = nLoci
  )

# Merge
cluster_seg_snp <- inner_join(segment_counts, clusters_snps, by = "Cluster")
cluster_seg_snp <- inner_join(clusters_data, cluster_seg_snp, by = "Cluster")

# Split the 'Cluster' column at "_", and create the 'r2' column with the second part
cluster_seg_snp$r2 <- sapply(strsplit(as.character(cluster_seg_snp$Cluster), "_"), `[`, 2)

# Add chromosome number and population name
cluster_seg_snp <- data.frame(Chromosome = rep(3, nrow(cluster_seg_snp)), cluster_seg_snp)
cluster_seg_snp <- data.frame(Population = rep("NEW", nrow(cluster_seg_snp)), cluster_seg_snp)


# Reorder the columns
cluster_seg_snp <- cluster_seg_snp |>
  dplyr::select(
    Chromosome, Cluster, r2, Start, End, nSegments, nSNPs
  ) |>
  dplyr::arrange(
    Chromosome, Start
  )

head(cluster_seg_snp)


# Save the data for plotting later
saveRDS(cluster_seg_snp, file = here("output", "ldna", "pop", "chr3", "NEW_plot2.rds"))
```


### 5.2 AUT (chromosome 3)
Clean env and memory
```{r clean-env-aut-chr3}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```

Import the data
```{r aut-chr3-load-ld-matrix, eval=TRUE}
ld1 <-
  read.delim(
    here("output", "ldna", "pop", "chr3", "AUT.chr3.txt"),
    sep = "\t",
    header = T,
    row.names = 1,
    stringsAsFactors = F,
    check.names = FALSE
  )
```

Check the LD matrix we built using bash
```{r aut-chr3-check-ld-matrix, eval=TRUE}
head(ld1)
```

The row and column names are correct. Now we can convert the data frame to matrix

```{r aut-chr3-convert-to-matrix, eval=TRUE}
# Create the matrix for ldna
ld2<- as.matrix(sapply(ld1, as.numeric))

# Get names
names<- rownames(ld1)
#names<-gsub("-", ".", names)

# Add names
row.names(ld2)<-names

# Clean non-numeric values
ld2[is.nan(ld2)] <- NA
ld2[is.infinite(ld2)] <- NA

# Remove diagonal values = 1
diag(ld2)=NA

# Make sure it is only the lower triangle matrix
ld2[!lower.tri(ld2)] <- NA
```

Create a LDna object
```{r aut-chr3-run-ldna-raw, eval=TRUE}
# create the LDna object (reduce cores to limit memory during full render)
diag(ld2) <- NA
ld2[!lower.tri(ld2)] <- NA
gc()
ldna <- LDnaRaw(ld2, mc.cores=2, method = "single")
```

Check the object
```{r aut-chr3-check-ldna-str, eval=TRUE}
str(ldna)
```

Save it
```{r aut-chr3-save-ldna, eval=TRUE}
saveRDS(ldna, file = here("output", "ldna", "pop", "chr3", "AUT.rds"))
saveRDS(ld2, file = here("output", "ldna", "pop", "chr3", "AUT_ld2.rds"))
```

Load the data
```{r aut-chr3-read-ldna}
ldna <- readRDS(here("output", "ldna", "pop", "chr3", "AUT.rds"))
ld2 <- readRDS(here("output", "ldna", "pop", "chr3", "AUT_ld2.rds"))
```

We can use the same parameters that we used for the other population
```{r aut-chr3-set-edge-params}
# snp_count <- ncol(ld2)
edges_max <- 100
edges_min <- 20
```

We can test the number of edges in increments of 10 and use the default phi=2
```{r aut-chr3-extract-branches-loop}
pdf(file = here("output", "ldna", "pop", "chr3", paste(edges_min, "3.AUT_loop.edges.pdf", sep = "_")), width = 20, height = 12)
op <- par(mfcol=c(1,4))

# Adjust the step size here
step_size <- 10  # for example, try a smaller step like 10

for (edges_estimate in seq(edges_min, edges_max, by = step_size)) {
  clusters_count_prev <- 100
  clusters_count <- 50

  # The 'if' condition seems redundant here since clusters_count_prev is always set to 100 before the check
  if (clusters_count_prev > clusters_count) {
    print(edges_estimate) # This prints the current edges_estimate, showing that the loop is working
    clusters_count_prev <- clusters_count

    # Extract clusters
    clusters <- extractBranches(ldna, min.edges=edges_estimate)

    # Summarize the clusters
    summary <- summaryLDna(ldna, clusters, ld2)

    # Write the summary to a file
    write.table(summary, file = here("output", "ldna", "pop", "chr3", "summary_AUT.txt"), row.names=FALSE, sep="\t", quote=FALSE)

    # Update clusters_count with the number of clusters just written
    clusters_count <- nrow(read.delim(here("output", "ldna", "pop", "chr3", "summary_AUT.txt")))
  }
}

# Reset graphical parameters and close all open graphic devices
par(op)
while (!is.null(dev.list())) dev.off()
```

Get cluster information
```{r aut-chr3-get-cluster-info}
# to get the list of names of the clusters
# Find the length of the longest vector
max_length <- max(sapply(clusters, length))

# Function to pad vectors with NAs to make them the same length
pad_vector_to_max_length <- function(vec, max_length) {
  c(vec, rep(NA, max_length - length(vec)))
}

# Apply the padding function to each vector in the list and combine into a data frame
do1 <- as.data.frame(do.call(cbind, lapply(clusters, pad_vector_to_max_length, max_length)))

# to get list of snps for each cluster
do <- do.call(cbind, unname(lapply(clusters, `length<-`, max(lengths(clusters)))))
colnames(do) <- colnames(do1)
```

Save it
```{r aut-chr3-build-cluster-df}
max_length <- max(sapply(clusters, length))
# Initialize an empty list
cluster_df_list <- list()

# Loop through each cluster and pad with NAs
for (cluster_name in names(clusters)) {
  cluster_length <- length(clusters[[cluster_name]])
  padded_cluster <- c(clusters[[cluster_name]], rep(NA, max_length - cluster_length))
  cluster_df_list[[cluster_name]] <- padded_cluster
}

# Combine the lists into a data frame
df <- data.frame(cluster_df_list)
# Remove 'X' from column names
names(df) <- make.names(names(df), unique = FALSE)
names(df) <- sub("X", "", names(df), fixed = TRUE)

# Save it
saveRDS(df, file = here("output", "ldna", "pop", "chr3", "AUT_clusters_snps.rds"))
head(df)
```
We can save it in a different format as well
```{r aut-chr3-clusters-long-format}
# Assuming your data frame is named df
long_df <- gather(df, Cluster, SNP, everything())

# View the first few rows of the transformed data
head(long_df)

# Save it
saveRDS(long_df, file = here("output", "ldna", "pop", "chr3", "AUT_clusters_snps1.rds"))
```

Make long format
```{r aut-chr3-melt-clusters}
aut_chr2 <- melt(do, na.rm = T, value.name = "value")
head(aut_chr2)
```


Update names
```{r aut-chr3-update-colnames}
colnames(aut_chr2)<- c("v1", "cluster", "SNP")
head(aut_chr2)
```


Import the bim file to get the SNP positions
```{r aut-chr3-import-bim-positions}
# Import the data
aut_snps_chr <- import_bim(here("output", "ldna", "pop", "AUT.bim")) |>
  dplyr::select(
    SNP, Scaffold, Position
  ) |>
  dplyr::rename(
    Chromosome = Scaffold
  )

# Check it
head(aut_snps_chr)
```

Merge the cluster data and the SNP data
```{r aut-chr3-merge-clusters-positions}
# merge dataframes
merged<- merge(aut_chr2, aut_snps_chr, by="SNP", all.x=TRUE)
merged<-na.omit(merged)
head(merged)
```

Select the columns we need
```{r aut-chr3-select-cluster-cols}
# subset
merged <- subset(merged, select = c(cluster, SNP, Position))
head(merged)
```

Sort by distance
```{r aut-chr3-sort-by-position}
merged<- merged[order(merged$Position),,drop=FALSE]
head(merged)
```


```{r aut-chr3-prep-sushi-data}
# for parsing later
sushi1<- subset(merged, select = c(cluster, Position))
head(sushi1)
```

Save the data
```{r aut-chr3-save-cluster-files, warning=FALSE}
# Create directory
new_directory_path <- here("output", "ldna", "pop", "chr3", "clusters")
dir.create(new_directory_path, recursive = TRUE)

# save file
write.table(merged, file = here("output", "ldna", "pop", "chr3", "SNPs_clusters_AUT_chr3.txt"), row.names = F, sep = "\t", quote = F)
write.table(sushi1, file = here("output", "ldna", "pop", "chr3", "clusters", "sushi_AUT_chr3.txt"), row.names = F, sep = "\t", quote = F)
```

Parse the file1 to get the start and end of the clusters as well as their size. Remember, recombination make it a mosaic, so the clusters are "mixed".
```{bash}
awk '
function print_row() {
    if ( feature != "" )
       print feature, start, end, (end - start)
}

BEGIN {
    FS=OFS="\t";
    print "Cluster", "Start", "End", "Size";
}
NR == 1 { next } # Skip the first line
$1 != feature {
    print_row();
    feature = $1;
    start = $2;
    end = $2;
    next;
}
{
    end = $2;
}
END {
    print_row();
}
' output/ldna/pop/chr3/clusters/sushi_AUT_chr3.txt > output/ldna/pop/chr3/clusters/sushi2_AUT_chr3.txt;
head output/ldna/pop/chr3/clusters/sushi2_AUT_chr3.txt
```

Get SNP count
```{r aut-chr3-count-cluster-segments}
# Define the path
input_path <- here("output", "ldna", "pop", "chr3", "clusters", "sushi2_AUT_chr3.txt")

# Read the data into R
clusters_data <- read_table(input_path, col_names = TRUE, col_types = NULL)

# Arrange by position
clusters_data <- clusters_data |>
  arrange(Start)

# Count segments for each cluster - if you want only the counts
segment_counts <- clusters_data %>%
  group_by(Cluster) %>%
  summarize(nSegments = n(), .groups = 'drop')

# View the result
head(segment_counts)
```

Plot it
```{r aut-chr3-plot-clusters-basic, fig.height=5, fig.width=8}
# Define the path
input_path <-
  here("output",
       "ldna",
       "pop",
       "chr3",
       "clusters",
       "sushi2_AUT_chr3.txt")

# Read the data into R
clusters_data <- read.table(input_path, header = TRUE, sep = "\t") |>
  arrange(Start)

# Arrange by position
clusters_data <- clusters_data |>
  arrange(Start)

# Filter out rows with Size equal to 0
clusters_data <- subset(clusters_data, Size > 100000) # show only bigger than 100kb


# Calculate the maximum size for scaling
max_size <- max(clusters_data$Size)

# Create Start_Mb and End_Mb within clusters_data
clusters_data$Start_Mb <- clusters_data$Start / 1e6
clusters_data$End_Mb <- clusters_data$End / 1e6

# Create the plot
ggplot(clusters_data, aes(xmin = Start_Mb, xmax = End_Mb, ymin = 0, ymax = Size / 1e6)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = label_number(unit = "M"), breaks = pretty_breaks(n = 10)) +
  scale_y_continuous(labels = label_number(unit = "M"), breaks = pretty_breaks(n = 10)) +
  labs(x = "Chromosome 3 (Mb)", y = "Cluster Size (Mb)", fill = "Cluster ID") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    panel.grid.minor.x = element_blank(),
    legend.position = "top",
    legend.box = "horizontal"
  ) +
  guides(fill = guide_legend(ncol = 7, title.position = "top", title.hjust = 0.5))


# Use ggsave to save the plot as a PDF
ggsave(
  filename = here("output", "ldna", "pop", "chr3", "clusters", "AUT_chr3.pdf"),
  device = "pdf",
  width = 8,
  height = 5,
  units = "in"
)
```


We can annotate the number of segments in the legend.

```{r aut-chr3-plot-clusters-annotated, fig.height=5, fig.width=10}
# Merge SNP counts into clusters_data while specifying suffixes
annotated_data <- merge(clusters_data, segment_counts, by = "Cluster")

# Create a new column with Cluster names and SNP counts
annotated_data$Cluster_with_SNPs <- with(annotated_data, paste(Cluster, " (", nSegments, ")", sep=""))

# Now the plot uses the new Cluster_with_SNPs for the fill legend
ggplot(annotated_data, aes(xmin = Start_Mb, xmax = End_Mb, ymin = 0, ymax = Size / 1e6)) +
  geom_rect(aes(fill = as.factor(Cluster_with_SNPs)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = scales::label_number(unit = "M"), breaks = scales::pretty_breaks(n = 10)) +
  scale_y_continuous(labels = scales::label_number(unit = "M"), breaks = scales::pretty_breaks(n = 10)) +
  labs(x = "Chromosome 3 (Mb)", y = "Cluster Size (Mb)", fill = "Cluster ID") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    panel.grid.minor.x = element_blank(),
    legend.position = "top",
    legend.box = "horizontal"
  ) +
  guides(fill = guide_legend(nrow = 10, title.position = "top", title.hjust = 0.5))

# Use ggsave to save the plot as a PDF
ggsave(
  filename = here("output", "ldna", "pop", "chr3", "clusters", "AUT_chr3b.pdf"),
  device = "pdf",
  width = 10,
  height = 5,
  units = "in"
)
```

Save plotting data, we can arrange the column order first
```{r aut-chr3-save-plot-data}
plot_data <- annotated_data |>
  dplyr::select(
    Cluster, nSegments, Start, End, Size, Start_Mb, End_Mb
  )

# Create a new column with Chromosome
plot_data <- data.frame(Chromosome = rep(1, nrow(plot_data)), plot_data)

# Split the 'Cluster' column at "_", and create the 'r2' column with the second part
plot_data$r2 <- sapply(strsplit(as.character(plot_data$Cluster), "_"), `[`, 2)

# Move the 'r2' column to be the second column, right after 'Chromosome'
plot_data <- plot_data[, c(1:2, ncol(plot_data), 3:(ncol(plot_data)-1))]


# Arrange
plot_data <- plot_data |>
  dplyr::arrange(Start)

# Check it
head(plot_data)

# Save it
saveRDS(plot_data, file = here("output", "ldna", "pop", "chr3", "AUT_plot.rds"))
```

We can also import the summary file to create a object with SNP counts per segment
```{r aut-chr3-build-cluster-seg-snp}
# Define the path
input_path <-
  here("output",
       "ldna",
       "pop",
       "chr3",
       "summary_AUT.txt")

# Read the data into R
clusters_snps <- read.table(input_path, header = TRUE, sep = "\t")

# We subset and rename the column 1
clusters_snps <- clusters_snps |>
  dplyr::select(
    Name, nLoci
  ) |>
  dplyr::rename(
    Cluster = Name,
    nSNPs = nLoci
  )

# Merge
cluster_seg_snp <- inner_join(segment_counts, clusters_snps, by = "Cluster")
cluster_seg_snp <- inner_join(clusters_data, cluster_seg_snp, by = "Cluster")

# Split the 'Cluster' column at "_", and create the 'r2' column with the second part
cluster_seg_snp$r2 <- sapply(strsplit(as.character(cluster_seg_snp$Cluster), "_"), `[`, 2)

# Add chromosome number and population name
cluster_seg_snp <- data.frame(Chromosome = rep(3, nrow(cluster_seg_snp)), cluster_seg_snp)
cluster_seg_snp <- data.frame(Population = rep("AUT", nrow(cluster_seg_snp)), cluster_seg_snp)


# Reorder the columns
cluster_seg_snp <- cluster_seg_snp |>
  dplyr::select(
    Chromosome, Cluster, r2, Start, End, nSegments, nSNPs
  ) |>
  dplyr::arrange(
    Chromosome, Start
  )

head(cluster_seg_snp)


# Save the data for plotting later
saveRDS(cluster_seg_snp, file = here("output", "ldna", "pop", "chr3", "AUT_plot2.rds"))
```


## 6. Combine the chromosomal data

Clean env and memory
```{r clean-env-combine}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```

Read the plotting data for NEW and AUT
```{r combine-read-plot-data}
# NEW
NEW1 <- readRDS(here("output", "ldna", "pop", "chr1", "NEW_plot2.rds")) |> mutate(Population = "NEW")
NEW2 <- readRDS(here("output", "ldna", "pop", "chr2", "NEW_plot2.rds")) |> mutate(Population = "NEW")
NEW3 <- readRDS(here("output", "ldna", "pop", "chr3", "NEW_plot2.rds")) |> mutate(Population = "NEW")

# AUT
AUT1 <- readRDS(here("output", "ldna", "pop", "chr1", "AUT_plot2.rds")) |> mutate(Population = "AUT")
AUT2 <- readRDS(here("output", "ldna", "pop", "chr2", "AUT_plot2.rds")) |> mutate(Population = "AUT")
AUT3 <- readRDS(here("output", "ldna", "pop", "chr3", "AUT_plot2.rds")) |> mutate(Population = "AUT")
```

Bind the objects

```{r combine-bind-all}
albo <- rbind(NEW1, NEW2, NEW3, AUT1, AUT2, AUT3)

head(albo)
```

We can create a table
```{r combine-build-table}
table1 <- albo |>
  mutate(Size = End - Start) |>
  dplyr::select(
    Population, Chromosome, Cluster, r2, nSegments, nSNPs, Start, End, Size
  ) |>
  arrange(Population, Chromosome, Start)
head(table1)
```

We can focus on LD blocks equal or bigger than 1Mb
```{r combine-filter-1mb}
table1 <- table1 %>%
  dplyr::filter(Size >= 1000000) %>%
  dplyr::mutate(`Size (Mb)` = round(Size / 1000000, 2)) # %>%
  # dplyr::select(Size)
head(table1)
```

Create table
```{r combine-display-table}
knitr::kable(table1)
# removed flextable code - not in container
# removed flextable code - not in container
# removed flextable code - not in container
# removed flextable code - not in container
# removed flextable code - not in container
# removed flextable code - not in container
# removed flextable code - not in container
# removed flextable code - not in container
# removed flextable code - not in container
# removed flextable code - not in container
# removed flextable code - not in container
# removed flextable code - not in container
# removed flextable code - not in container
# Save the Word document
print(doc, target = here("output", "ldna", "clusters_AUT_NEW_1Mb.docx"))
```

We can create a facet plot now for all chromosomes

```{r combine-facet-plot-fixed, fig.height=5, fig.width=10}
# To plot only clusters equal or bigger than 1Mb
albo2 <- table1 |> dplyr::filter(Size >= 1000000)

# Function to format numbers as Mb
label_mb <- function(x) {
  sprintf("%.0fMb", x / 1e6)
}

# Plot it
ggplot(albo2, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    strip.background = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position = "none",
    panel.spacing.x = unit(1, "lines")  # Adjust the unit and number to increase space as needed
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed") +
  guides(fill = "none")

# Use ggsave to save the plot as a PDF
# ggsave(
#   filename = here("output", "ldna", "AUT_NEW_fixed.pdf"),
#   device = "pdf",
#   width = 10,
#   height = 5,
#   units = "in"
# )
```

Make each plot with different scales
```{r combine-facet-plot-free, fig.height=5, fig.width=8}
# Plot it
ggplot(albo2, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    panel.grid.minor.x = element_blank(),
    # strip.background = element_rect(fill = "#e8e8e8", colour = NA),
    legend.position = "none",
    panel.spacing.x = unit(1, "lines")
  ) +
  facet_wrap(~ Population + Chromosome, scales = "free_y", ncol = 3) +  # Free y scale, and each chromosome gets its own row
  guides(fill = "none")

# Use ggsave to save the plot as a PDF
# ggsave(
#   filename = here("output", "ldna", "AUT_NEW_free.pdf"),
#   device = "pdf",
#   width = 8,
#   height = 5,
#   units = "in"
# )
```

We can also remove the large clusters and plot with fixed scale to look at the cluster between 1 and 10Mb
```{r combine-facet-plot-1-10mb, fig.height=5, fig.width=8}
# To plot only clusters equal or bigger than 1Mb but less that 10Mb
albo3 <- albo2 |> dplyr::filter(Size >= 1000000, Size <= 10000000)


# Plot it
ggplot(albo3, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    strip.background = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position = "none",
    panel.spacing.x = unit(1, "lines")  # Adjust the unit and number to increase space as needed
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed") +
  guides(fill = "none")

# Use ggsave to save the plot as a PDF
# ggsave(
#   filename = here("output", "ldna", "AUT_NEW_fixed_1_to_10Mb.pdf"),
#   device = "pdf",
#   width = 8,
#   height = 5,
#   units = "in"
# )
```

It seems that the linkage patters are quite different from AUT to the wild population. I am not sure why we are seeing clusters at such long distance the wild population for chromosome 2 and 3

We could do the analysis by sex, but we only have 13 females and 15 males in AUT. Perhaps we could use the same number of samples for each. However, the sample size will be too small still.

## 7. Plot SNPs on DE genes with LD blocks
```{r load-snps-scan}
snps_scan <-
  read.table(
    # here("output", "pcadapt", "man_aut_common_SNPs_pcadapt_outflank.txt"),
    here("output", "snpeff", "SNPs_79_DE.txt"),
    stringsAsFactors = FALSE
  ) |>
  dplyr::rename(
    SNP = V1
  )

head(snps_scan)
```

Import bim file
```{r import-bim-for-snps}
# Import the data
snps <- import_bim(here("output", "ldna", "files", "file1.bim"))

# Check it
head(snps)
```

Merge the objects
```{r merge-snps-scan-bim}
merged_snps <- inner_join(snps_scan, snps, by = "SNP") |>
  dplyr::rename(
    Chromosome = Scaffold
  )
head(merged_snps)
```

We can add them as lines in the plot
```{r plot-ld-blocks-with-snps, fig.height=5, fig.width=8}
# First, create a new data frame with 1Mb windows and count SNPs in each window
merged_snps$Window <- floor(merged_snps$Position / 20e6) * 20e6 # 20mb
snp_counts <- merged_snps %>%
  group_by(Chromosome, Window) %>%
  summarize(SNP_count = n(), .groups = 'drop')

# Now, add this SNP information to the plot
snp_plot <- ggplot() +
  geom_rect(data = albo2, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size, fill = as.factor(Cluster)),
            color = "black", linewidth = 0.2) +
  geom_vline(data = snp_counts, aes(xintercept = Window + 0.5e6), color = "pink", linetype = "dotted", linewidth = 0.5) +
  geom_text(data = snp_counts, aes(x = Window + 0.5e6, y = 8e6, label = SNP_count), vjust = -0.5, color = "blue", size = 2) +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position (Mb)", y = "Cluster Size (Mb)") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotdash"),
    strip.background = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position = "none",
    panel.spacing.x = unit(1, "lines")
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed")

# Display the plot
print(snp_plot)

#
# # Use ggsave to save the plot as a PDF
# ggsave(
#   filename = here("output", "ldna", "linkage_snps_genes_de.pdf"),
#   device = "pdf",
#   width = 8,
#   height = 5,
#   units = "in"
# )
```

Find the non-overlapping clusters
```{r find-non-overlapping-clusters}
# Split the data into two subsets based on Population
albo_NEW <- subset(albo2, Population == "NEW")
albo_AUT <- subset(albo2, Population == "AUT")

# Function to find non-overlapping clusters
find_non_overlapping_clusters <- function(new_clusters, aut_clusters) {
  non_overlapping <- vector("list", length(new_clusters$Chromosome))
  names(non_overlapping) <- unique(new_clusters$Chromosome)

  for (chr in unique(new_clusters$Chromosome)) {
    new_chrom_clusters <- new_clusters[new_clusters$Chromosome == chr, ]
    aut_chrom_clusters <- aut_clusters[aut_clusters$Chromosome == chr, ]

    non_overlap <- vector("list", nrow(new_chrom_clusters))

    for (i in 1:nrow(new_chrom_clusters)) {
      overlaps <- any(aut_chrom_clusters$Start <= new_chrom_clusters$End[i] & aut_chrom_clusters$End >= new_chrom_clusters$Start[i])
      if (!overlaps) {
        non_overlap[[i]] <- new_chrom_clusters[i, ]
      }
    }

    non_overlapping[[chr]] <- do.call(rbind, non_overlap)
  }

  do.call(rbind, non_overlapping)
}

# Find non-overlapping clusters
non_overlapping_clusters <- find_non_overlapping_clusters(albo_NEW, albo_AUT)

# View the non-overlapping clusters
non_overlapping_clusters
```
Sanity check: Second way to get it
```{r find-non-overlapping-segments}
# Function to find non-overlapping segments
find_non_overlapping_segments <- function(new_clusters, aut_clusters) {
  non_overlapping <- list()

  for (chr in unique(new_clusters$Chromosome)) {
    new_chrom_segments <- new_clusters[new_clusters$Chromosome == chr, ]
    aut_chrom_segments <- aut_clusters[aut_clusters$Chromosome == chr, ]

    non_overlap <- list()

    for (i in 1:nrow(new_chrom_segments)) {
      overlaps <- FALSE
      for (j in 1:nrow(aut_chrom_segments)) {
        if (new_chrom_segments$Start[i] <= aut_chrom_segments$End[j] && new_chrom_segments$End[i] >= aut_chrom_segments$Start[j]) {
          overlaps <- TRUE
          break
        }
      }

      if (!overlaps) {
        non_overlap <- c(non_overlap, list(new_chrom_segments[i, ]))
      }
    }

    non_overlapping[[chr]] <- do.call(rbind, non_overlap)
  }

  do.call(rbind, non_overlapping)
}

# Find non-overlapping segments
non_overlapping_segments <- find_non_overlapping_segments(albo_NEW, albo_AUT)

# View the non-overlapping segments
non_overlapping_segments
```
```{r find-unique-clusters}
# Function to find unique non-overlapping clusters for each population
find_unique_clusters <- function(population1, population2) {
  unique_clusters <- list()

  for (chr in unique(population1$Chromosome)) {
    pop1_chrom_clusters <- population1[population1$Chromosome == chr, ]
    pop2_chrom_clusters <- population2[population2$Chromosome == chr, ]

    unique_clusters_chr <- list()

    for (i in 1:nrow(pop1_chrom_clusters)) {
      overlaps <- any(
        pop2_chrom_clusters$Start <= pop1_chrom_clusters$End[i] &
        pop2_chrom_clusters$End >= pop1_chrom_clusters$Start[i]
      )

      if (!overlaps) {
        unique_clusters_chr <- c(unique_clusters_chr, list(pop1_chrom_clusters[i, ]))
      }
    }

    if (length(unique_clusters_chr) > 0) {
      unique_clusters[[chr]] <- do.call(rbind, unique_clusters_chr)
    }
  }

  do.call(rbind, unique_clusters)
}

# Find unique non-overlapping clusters for NEW and AUT populations
unique_clusters_NEW <- find_unique_clusters(albo2[albo2$Population == "NEW",], albo2[albo2$Population == "AUT",])
unique_clusters_AUT <- find_unique_clusters(albo2[albo2$Population == "AUT",], albo2[albo2$Population == "NEW",])

# View the unique non-overlapping clusters for each population
unique_clusters_NEW
unique_clusters_AUT
```


Plot it
```{r plot-unique-clusters, fig.height=5, fig.width=10}
# Combine the two data frames
combined_clusters <- rbind(unique_clusters_NEW, unique_clusters_AUT)


# Function to format numbers as Mb
label_mb <- function(x) {
  sprintf("%.0fMb", x / 1e6)
}

# Plot it
ggplot(combined_clusters, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    strip.background = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position = "none",
    panel.spacing.x = unit(1, "lines")  # Adjust the unit and number to increase space as needed
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed") +
  guides(fill = "none")

# Use ggsave to save the plot as a PDF
# ggsave(
#   filename = here("output", "ldna", "AUT_NEW_fixed_non_overlapping.pdf"),
#   device = "pdf",
#   width = 10,
#   height = 5,
#   units = "in"
# )
```

```{r load-snps-de-genes}
snps_DEgenes<- read_delim(here("output", "snpeff", "SNPs_on_DE_genes.txt"), delim = "\t", col_names = TRUE, show_col_types = FALSE)
head(snps_DEgenes)
```

Check if any SNP fall within the clusters

```{r check-snps-in-clusters}
# Initialize an empty data frame to store the results
results <- data.frame(Population = character(), Chromosome = integer(),
                      Cluster = character(), r2 = numeric(),
                      nSegments = integer(), nSNPs = integer(),
                      Start = integer(), End = integer(), Size = integer(),
                      SizeMb = numeric(), SNP = character(), Position_chr = integer(),
                      stringsAsFactors = FALSE)

for (i in 1:nrow(snps_DEgenes)) {
    for (j in 1:nrow(combined_clusters)) {
        if (snps_DEgenes$Chromosome[i] == combined_clusters$Chromosome[j] &&
            snps_DEgenes$Position_chr[i] >= combined_clusters$Start[j] &&
            snps_DEgenes$Position_chr[i] <= combined_clusters$End[j]) {

            # Create a new row as a data frame with the same column names
            new_row <- data.frame(Population = combined_clusters$Population[j],
                                  Chromosome = combined_clusters$Chromosome[j],
                                  Cluster = combined_clusters$Cluster[j],
                                  r2 = combined_clusters$r2[j],
                                  nSegments = combined_clusters$nSegments[j],
                                  nSNPs = combined_clusters$nSNPs[j],
                                  Start = combined_clusters$Start[j],
                                  End = combined_clusters$End[j],
                                  Size = combined_clusters$Size[j],
                                  SizeMb = combined_clusters$Size[j],
                                  SNP = snps_DEgenes$SNP[i],
                                  Position_chr = snps_DEgenes$Position_chr[i],
                                  stringsAsFactors = FALSE)

            # Append the new row to the results data frame
            results <- rbind(results, new_row)
        }
    }
}


head(results)
```

We can count how many SNPs we have per cluster per population
```{r count-snps-per-cluster}
snp_counts <- results %>%
  select(Population, Chromosome, Cluster, nSNPs) %>%
  distinct() %>%
  rename(SNP_Count = nSNPs)


head(snp_counts)
```

Plot the clusters for which we have SNPs on the DE genes
```{r plot-clusters-with-snp-counts, fig.height=5, fig.width=8}
# Compute SNP counts for each cluster
# snp_counts2 <- aggregate(SNP ~ Cluster, data = results, FUN = length)
# colnames(snp_counts2)[2] <- "SNP_Count"


ggplot(results_with_counts, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  geom_text(aes(x = (Start + End) / 2, y = Size + 0.5, label = SNP_Count), size = 3, vjust = -0.2) + # Annotate SNP counts on top
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    strip.background = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position = "none",
    panel.spacing.x = unit(1, "lines")
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed") +
  guides(fill = "none")


```
```{r plot-clusters-with-ids}
# 1. Generate integer IDs for each cluster in plotting order
cluster_ids <- data.frame(
  Cluster     = sort(unique(results$Cluster)),
  Cluster_ID  = seq_along(unique(results$Cluster))
)

# 2. Merge IDs into your results
results_with_ids <- merge(results, cluster_ids, by = "Cluster")

# 3. Plot, labeling by your original 1,2,3,… IDs
ggplot(results_with_ids, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = factor(Cluster_ID)), color = "black", linewidth = 0.2) +
  geom_text(aes(x = (Start + End) / 2,
                y = Size + 0.5,
                label = Cluster_ID),
            size = 3, vjust = -0.2) +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y        = element_text(margin = margin(t = 0, r = 5)),
    axis.ticks.y       = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    strip.background   = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position    = "none",
    panel.spacing.x    = unit(1, "lines")
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed") +
  guides(fill = "none")

```

```{r plot-clusters-with-fill-labels, fig.height=5, fig.width=8}
# 1. Merge integer IDs into your SNP‐counted data
cluster_ids <- data.frame(
  Cluster    = sort(unique(results_with_counts$Cluster)),
  Cluster_ID = seq_along(unique(results_with_counts$Cluster))
)
results_with_counts <- merge(results_with_counts, cluster_ids, by = "Cluster")

# 2. Build the fill label
results_with_counts$Cluster_SNP <- with(
  results_with_counts,
  paste0(Cluster, " (", SNP_Count, " SNPs)")
)

# 3. Plot with fill by SNP count label and text by cluster ID
ggplot(results_with_counts, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = Cluster_SNP), color = "black", linewidth = 0.2) +
  geom_text(
    aes(x = (Start + End) / 2, y = Size + 0.5, label = Cluster_ID),
    size = 3, vjust = -0.2
  ) +
  scale_fill_discrete(name = "Cluster and SNP Count") +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y        = element_text(margin = margin(t = 0, r = 5)),
    axis.ticks.y       = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    strip.background   = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position    = "right",
    panel.spacing.x    = unit(1, "lines")
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed")
```



Save
```{r save-results-with-counts, eval=TRUE}
saveRDS(results_with_counts, file = here("output", "ldna", "results_with_counts.rds"))
```

Read the data
```{r read-results-with-counts}
results_with_counts <- readRDS(file = here("output", "ldna", "results_with_counts.rds"))
```


```{r plot-final-clusters-annotated, fig.height=5, fig.width=12}
# Rebuild the fill label: add ID before Cluster and drop “ SNPs”
results_with_counts$Cluster_SNP <- with(
  results_with_counts,
  paste0(Cluster_ID, ": ", Cluster, " (", SNP_Count, ")")
)

# 1. Reorder Cluster_SNP factor by Cluster_ID so the legend follows numeric order
results_with_counts$Cluster_SNP <- factor(
  results_with_counts$Cluster_SNP,
  levels = results_with_counts %>%
    distinct(Cluster_ID, Cluster_SNP) %>%
    arrange(Cluster_ID) %>%
    pull(Cluster_SNP)
)

# 2. Plot with sorted legend
ggplot(results_with_counts, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = Cluster_SNP), color = "black", linewidth = 0.2) +
  geom_text(
    aes(x = (Start + End) / 2, y = Size + 0.5, label = Cluster_ID),
    size = 3, vjust = -0.2
  ) +
  scale_fill_discrete(name = "Cluster and SNP Count") +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y        = element_text(margin = margin(t = 0, r = 5)),
    axis.ticks.y       = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    strip.background   = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position    = "right",
    panel.spacing.x    = unit(1, "lines")
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed")



# Use ggsave to save the plot as a PDF
ggsave(
  filename = here("output", "ldna", "AUT_NEW_non_overlapping_clusters_with_SNPs_on_DE_genes.pdf"),
  device = "pdf",
  width = 10,
  height = 5,
  units = "in"
)
```


Save it
```{r save-results-excel}
library(writexl)

# 1. Drop duplicate suffix columns
cleaned_results <- results_with_counts %>%
  select(-Cluster_ID.x, -Cluster_ID.y) %>%
  mutate(
    Start        = formatC(Start,        format = "f", digits = 0),
    End          = formatC(End,          format = "f", digits = 0),
    Position_chr = formatC(Position_chr, format = "f", digits = 0)
  )


# 2. Export cleaned results_with_counts to Excel
write_xlsx(
  list(Results = cleaned_results),
  here("output/ldna/results_with_counts.xlsx")
)

```

<!-- Development marker removed -->

Now we can get the SNPs on the DE genes
```{r load-snps-79-de-genes}
snps79_DE_genes<- read_delim(here("output", "snpeff","SNPs_79_DE.txt"), delim = "\t", col_names = FALSE, show_col_types = FALSE)
head(snps79_DE_genes)
```

Or the SNPs from the selection scan
```{r load-snps-selection}
snps_selection <- read_delim(here("output", "pcadapt","outlier_157_SNPs.txt"), delim = "\t", col_names = FALSE, show_col_types = FALSE)
head(snps_selection)
```



Filter
```{r filter-results-de-snps}
# Create a logical vector to filter rows based on SNP IDs
filter_vector <- results_with_counts$SNP %in% snps79_DE_genes$X1
# filter_vector <- results_with_counts$SNP %in% snps_selection$X1  # we do not have any

# Subset results_with_counts to keep only rows where SNP matches
filtered_results <- results_with_counts[filter_vector, ]

head(filtered_results)
# length(filtered_results)
```

We have 16 SNPs
```{r plot-filtered-clusters, fig.height=5, fig.width=8}
# Assign unique numbers to each cluster and create labels
filtered_results <- filtered_results %>%
  mutate(Cluster_Number = as.numeric(as.factor(Cluster))) %>%
  arrange(Cluster_Number) %>%
  mutate(Cluster_Label = paste(Cluster_Number, "_", Cluster, " (", SNP_Count, ")", sep=""))

# Reorder the factor levels of Cluster_Label based on Cluster_Number
filtered_results$Cluster_Label <- factor(filtered_results$Cluster_Label,
                                            levels = unique(filtered_results$Cluster_Label[order(filtered_results$Cluster_Number)]))

# Plotting
ggplot(filtered_results, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = Cluster_Label), color = "black", linewidth = 0.2) +
  geom_text(aes(x = (Start + End) / 2, y = Size + 0.5, label = Cluster_Number), size = 3, vjust = -0.2) +
  scale_fill_discrete(name = "Cluster ID") +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    strip.background = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position = "right",
    panel.spacing.x = unit(1, "lines")
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed")


# Use ggsave to save the plot as a PDF
ggsave(
  filename = here("output", "ldna", "AUT_NEW_overlapping_clusters_with_SNPs_on_DE_genes_intersect.pdf"),
  device = "pdf",
  width = 10,
  height = 5,
  units = "in"
)
```
This are the clusters for which we found the SNP as outlier and the gene the SNP in on is DE



Get the overlapping clusters
```{r find-overlapping-clusters}
# Function to find overlapping clusters
find_overlapping_clusters <- function(population1, population2) {
  overlapping_clusters <- list()

  for (chr in unique(population1$Chromosome)) {
    pop1_chrom_clusters <- population1[population1$Chromosome == chr, ]
    pop2_chrom_clusters <- population2[population2$Chromosome == chr, ]

    overlapping_clusters_chr <- list()

    for (i in 1:nrow(pop1_chrom_clusters)) {
      for (j in 1:nrow(pop2_chrom_clusters)) {
        if (pop1_chrom_clusters$Start[i] <= pop2_chrom_clusters$End[j] &&
            pop1_chrom_clusters$End[i] >= pop2_chrom_clusters$Start[j]) {
          overlapping_clusters_chr <- c(overlapping_clusters_chr, list(pop1_chrom_clusters[i, ]))
          break
        }
      }
    }

    if (length(overlapping_clusters_chr) > 0) {
      overlapping_clusters[[chr]] <- do.call(rbind, overlapping_clusters_chr)
    }
  }

  do.call(rbind, overlapping_clusters)
}

# Subset data for NEW and AUT populations
albo_NEW <- subset(albo2, Population == "NEW")
albo_AUT <- subset(albo2, Population == "AUT")

# Find overlapping clusters between NEW and AUT
overlapping_clusters <- find_overlapping_clusters(albo_NEW, albo_AUT)

# View the overlapping clusters
overlapping_clusters
```

Plot it
```{r plot-overlapping-clusters, fig.height=5, fig.width=10}
# Function to format numbers as Mb
label_mb <- function(x) {
  sprintf("%.0fMb", x / 1e6)
}

# Plot it
ggplot(overlapping_clusters, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    strip.background = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position = "none",
    panel.spacing.x = unit(1, "lines")  # Adjust the unit and number to increase space as needed
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed") +
  guides(fill = "none")

# Use ggsave to save the plot as a PDF
ggsave(
  filename = here("output", "ldna", "AUT_NEW_fixed_overlapping.pdf"),
  device = "pdf",
  width = 10,
  height = 5,
  units = "in"
)
```

## 8. Mapping back SNP clusters to scaffolds

Clean env and memory
```{r clean-env-mapping}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```

I created the chromosomal scale for the linkage networks analysis, now we have to map back the SNPs from clusters of interest to the scaffold scale, then we can check the gene annotation file to find what genes are in the linkage group.

Cluster 14 is on chromosome 2 and only in the AUT line. I create files with the SNP ids, we can import it (on the legend is the cluster 14_3139_0.78 (18))
```{r load-aut-chr2-clusters}
aut_ch2 <- readRDS(here("output", "ldna", "pop", "chr2", "AUT_clusters_snps.rds"))
str(aut_ch2$`3139_0.78`)
# We have 566 SNPs on this cluster
```

Now we can import the bim file with the scaffolds. We can use the backup file we created when we created the chromosomal scale.
```{bash}
head output/quality_control/file7_backup.bim
```

Import it
```{r import-backup-bim}
# Import the data
snps <- import_bim(here("output", "quality_control", "file7_backup.bim"))

# Check it
head(snps)
```


Now we can filter the 566 SNPs that are in the cluster and see what scaffolds there are located

```{r filter-cluster-14-snps}
cluster_14 <- snps[snps$SNP %in% aut_ch2$`3139_0.78`, ]
head(cluster_14) # 316 SNPs

# Save it
saveRDS(
  cluster_14, here(
    "output", "ldna", "cluster_14.rds"
    )
  )
```

Create table
```{r display-cluster-14-table}
knitr::kable(cluster_14)
# flextable removed
# flextable removed
# flextable removed
# flextable removed
# flextable removed
# flextable removed
# flextable removed
# flextable removed
# flextable removed
# flextable removed
# flextable removed
# flextable removed

# Save the Word document
print(doc, target = here("output", "ldna", "clusters_14_aut.docx"))
```

Let's check cluster 6_19_42 on AUT chr1
```{r load-aut-chr1-clusters}
aut_ch1 <- readRDS(here("output", "ldna", "pop", "chr1", "AUT_clusters_snps.rds"))
str(aut_ch1$`1942_0.64`)
# We have 566 SNPs on this cluster
```

```{r filter-cluster-6-snps}
cluster_6 <- snps[snps$SNP %in% aut_ch1$`1942_0.64`, ]
head(cluster_6) # 316 SNPs

# Save it
saveRDS(
  cluster_6, here(
    "output", "ldna", "cluster_6.rds"
    )
  )
```


Create table
```{r display-cluster-6-table}
knitr::kable(cluster_6)
# flextable removed
# flextable removed
# flextable removed
# flextable removed
# flextable removed
# flextable removed
# flextable removed
# flextable removed
# flextable removed
# flextable removed
# flextable removed
# flextable removed

# Save the Word document
print(doc, target = here("output", "ldna", "clusters_6_aut.docx"))
```


```{r get-all-snps-cluster-6-scaffolds}
cluster_6_all <- snps[snps$Scaffold %in% cluster_6$Scaffold, ]
head(cluster_6_all)
```


However we need the other SNPs to know if the blocks are connected or not. Now we know what scaffolds the SNPs are located and we can check it
```{r get-all-snps-cluster-14-scaffolds}
cluster_14_all <- snps[snps$Scaffold %in% cluster_14$Scaffold, ]
head(cluster_14_all)
```

Now we can create a new column and tag the SNPs that are linked
```{r tag-linked-snps}
# Creating the new column 'linked'
cluster_14_all <- cluster_14_all %>%
  mutate(linked = SNP %in% aut_ch2$`3139_0.78`)
head(cluster_14_all)

# Creating the new column 'linked'
cluster_6_all <- cluster_6_all %>%
  mutate(linked = SNP %in% aut_ch1$`1942_0.64`)
head(cluster_6_all)
```

Lets check how many linked SNPs we have and how many aren't
```{r count-linked-snps-cluster-14}
count_linked <- cluster_14_all %>%
  group_by(linked) %>%
  summarise(count = n())
count_linked
```

```{r count-linked-snps-cluster-6}
count_linked <- cluster_6_all %>%
  group_by(linked) %>%
  summarise(count = n())
count_linked
```

Get the 157 outliers
```{r load-outlier-snps-157}
snps_157 <-
  read.table(
    here("output", "pcadapt", "outlier_157_SNPs.txt"),
    stringsAsFactors = FALSE
    )

# Get the 157 SNPs
snps_157b <- cluster_14_all |>
  filter(SNP %in% snps_157$V1)

head(snps_157b)
```


```{r filter-157-snps-cluster-6}
# Get the 157 SNPs
snps_157c <- cluster_6_all |>
  filter(SNP %in% snps_157$V1)

head(snps_157c)
```

Plot it
```{r plot-cluster-14-scaffolds, fig.height=5, fig.width=8}
# Function to format numbers as Mb
label_mb <- function(x) {
  sprintf("%.0fMb", x / 1e6)
}


# Calculate the start and end positions for each stretch of TRUE or FALSE
rect_data <- cluster_14_all %>%
  arrange(Scaffold, Position) %>%
  mutate(change = linked != lag(linked, default = first(linked))) %>%
  group_by(Scaffold) %>%
  mutate(group_id = cumsum(change)) %>%
  group_by(Scaffold, group_id, linked) %>%
  summarize(start = min(Position), end = max(Position), .groups = 'drop') %>%
  ungroup()

# Plotting using geom_rect
ggplot(rect_data, aes(xmin = start, xmax = end, ymin = as.numeric(Scaffold) - 0.1, ymax = as.numeric(Scaffold) + 0.1)) +
  # Add a background layer for each scaffold with gray color
  geom_rect(data = rect_data %>% group_by(Scaffold) %>%
            summarize(start = min(start), end = max(end), .groups = 'drop'),
            aes(xmin = start, xmax = end, ymin = as.numeric(Scaffold) - 0.1, ymax = as.numeric(Scaffold) + 0.1),
            fill = "gray", inherit.aes = FALSE) +
  # Add the TRUE stretches
  geom_rect(aes(fill = linked), data = rect_data %>% filter(linked == TRUE)) +
  scale_fill_manual(values = c("TRUE" = "green")) +
  facet_wrap(~ Scaffold, scales = "fixed", ncol = 1) +
  geom_vline(data = snps_157b, aes(xintercept = Position, color = SNP), linetype = "solid", color = "red") +
  geom_text(data = snps_157b, aes(x = Position, y = as.numeric(Scaffold) + 0.2, label = SNP), inherit.aes = FALSE, angle = 90, vjust = 0, size = 2, check_overlap = TRUE) +
  theme_minimal() +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.spacing = unit(0.1, "lines"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 12, face = "bold", hjust = 0.5, margin = margin(t = 1, b = 5))
  ) +
  scale_x_continuous(labels = label_mb, breaks = scales::pretty_breaks(n = 6)) +
  labs(x = "Position", title = "Continuous Stretches of Linked SNPs by Scaffold - Cluster 14 AUT", fill = "Linked") +
  guides(fill = "none")

# Save Venn diagram to PDF
output_path <- here("output", "ldna", "figures", "cluster_14_aut_scaffolds.pdf")
ggsave(output_path, height = 5, width = 8, units = "in")
```



To make sure the windows appear even if they are small
```{r plot-cluster-14-scaffolds-min-width, fig.height=5, fig.width=8}
# Set a minimum width for the windows
min_width <- 1e4  # for example, 1 million base pairs

# Adjust the rect_data calculation
rect_data <- cluster_14_all %>%
  arrange(Scaffold, Position) %>%
  mutate(change = linked != lag(linked, default = first(linked))) %>%
  group_by(Scaffold) %>%
  mutate(group_id = cumsum(change)) %>%
  group_by(Scaffold, group_id, linked) %>%
  summarize(start = min(Position), end = max(Position), .groups = 'drop') %>%
  ungroup() %>%
  mutate(width = end - start,
         adjusted_end = ifelse(width < min_width, start + min_width, end))

# Plotting using geom_rect with adjusted ends
ggplot(rect_data, aes(xmin = start, xmax = adjusted_end, ymin = as.numeric(Scaffold) - 0.1, ymax = as.numeric(Scaffold) + 0.1)) +
  geom_rect(data = rect_data %>% group_by(Scaffold) %>%
            summarize(start = min(start), end = max(adjusted_end), .groups = 'drop'),
            aes(xmin = start, xmax = end, ymin = as.numeric(Scaffold) - 0.1, ymax = as.numeric(Scaffold) + 0.1),
            fill = "gray", inherit.aes = FALSE) +
  geom_rect(aes(fill = linked), data = rect_data %>% filter(linked == TRUE)) +
  scale_fill_manual(values = c("TRUE" = "green")) +
  facet_wrap(~ Scaffold, scales = "fixed", ncol = 1) +
  theme_minimal() +
  geom_vline(data = snps_157b, aes(xintercept = Position, color = SNP), linetype = "solid", color = "red") +
  geom_text_repel(data = snps_157b, aes(x = Position, y = as.numeric(Scaffold) + 0.2, label = SNP),
                  inherit.aes = FALSE, angle = 45, size = 2,
                  nudge_y = 0.1,   # Adjust this value as needed
                  direction = "y") +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(), panel.spacing = unit(0.1, "lines"),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.text = element_text(size = 12, face = "bold", hjust = 0.5)) +
  scale_x_continuous(labels = label_mb, breaks = scales::pretty_breaks(n = 6)) +
  labs(x = "Position", title = "Continuous Stretches of Linked SNPs by Scaffold - Cluster 14 AUT", fill = "Linked") +
  guides(fill = "none")

# Save Venn diagram to PDF
output_path <- here("output", "ldna", "figures", "cluster_14_aut_scaffolds_with_small_windows.pdf")
ggsave(output_path, height = 5, width = 8, units = "in")
```

Cluster 6
```{r plot-cluster-6-scaffolds, fig.height=5, fig.width=8}
# Function to format numbers as Mb
label_mb <- function(x) {
  sprintf("%.0fMb", x / 1e6)
}


# Calculate the start and end positions for each stretch of TRUE or FALSE
rect_data <- cluster_6_all %>%
  arrange(Scaffold, Position) %>%
  mutate(change = linked != lag(linked, default = first(linked))) %>%
  group_by(Scaffold) %>%
  mutate(group_id = cumsum(change)) %>%
  group_by(Scaffold, group_id, linked) %>%
  summarize(start = min(Position), end = max(Position), .groups = 'drop') %>%
  ungroup()

# Plotting using geom_rect
ggplot(rect_data, aes(xmin = start, xmax = end, ymin = as.numeric(Scaffold) - 0.1, ymax = as.numeric(Scaffold) + 0.1)) +
  # Add a background layer for each scaffold with gray color
  geom_rect(data = rect_data %>% group_by(Scaffold) %>%
            summarize(start = min(start), end = max(end), .groups = 'drop'),
            aes(xmin = start, xmax = end, ymin = as.numeric(Scaffold) - 0.1, ymax = as.numeric(Scaffold) + 0.1),
            fill = "gray", inherit.aes = FALSE) +
  # Add the TRUE stretches
  geom_rect(aes(fill = linked), data = rect_data %>% filter(linked == TRUE)) +
  scale_fill_manual(values = c("TRUE" = "green")) +
  facet_wrap(~ Scaffold, scales = "fixed", ncol = 1) +
  geom_vline(data = snps_157c, aes(xintercept = Position, color = SNP), linetype = "solid", color = "red") +
  geom_text(data = snps_157c, aes(x = Position, y = as.numeric(Scaffold) + 0.2, label = SNP), inherit.aes = FALSE, angle = 90, vjust = 0, size = 2, check_overlap = TRUE) +
  theme_minimal() +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.spacing = unit(0.1, "lines"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 12, face = "bold", hjust = 0.5, margin = margin(t = 1, b = 5))
  ) +
  scale_x_continuous(labels = label_mb, breaks = scales::pretty_breaks(n = 6)) +
  labs(x = "Position", title = "Continuous Stretches of Linked SNPs by Scaffold - Cluster 6 AUT", fill = "Linked") +
  guides(fill = "none")

# Save Venn diagram to PDF
output_path <- here("output", "ldna", "figures", "cluster_6_aut_scaffolds.pdf")
ggsave(output_path, height = 5, width = 8, units = "in")
```



To make sure the windows appear even if they are small
```{r plot-cluster-6-scaffolds-min-width, fig.height=5, fig.width=8}
# Set a minimum width for the windows
min_width <- 1e4  # for example, 1 million base pairs

# Adjust the rect_data calculation
rect_data <- cluster_6_all %>%
  arrange(Scaffold, Position) %>%
  mutate(change = linked != lag(linked, default = first(linked))) %>%
  group_by(Scaffold) %>%
  mutate(group_id = cumsum(change)) %>%
  group_by(Scaffold, group_id, linked) %>%
  summarize(start = min(Position), end = max(Position), .groups = 'drop') %>%
  ungroup() %>%
  mutate(width = end - start,
         adjusted_end = ifelse(width < min_width, start + min_width, end))

# Plotting using geom_rect with adjusted ends
ggplot(rect_data, aes(xmin = start, xmax = adjusted_end, ymin = as.numeric(Scaffold) - 0.1, ymax = as.numeric(Scaffold) + 0.1)) +
  geom_rect(data = rect_data %>% group_by(Scaffold) %>%
            summarize(start = min(start), end = max(adjusted_end), .groups = 'drop'),
            aes(xmin = start, xmax = end, ymin = as.numeric(Scaffold) - 0.1, ymax = as.numeric(Scaffold) + 0.1),
            fill = "gray", inherit.aes = FALSE) +
  geom_rect(aes(fill = linked), data = rect_data %>% filter(linked == TRUE)) +
  scale_fill_manual(values = c("TRUE" = "green")) +
  facet_wrap(~ Scaffold, scales = "fixed", ncol = 1) +
  theme_minimal() +
  geom_vline(data = snps_157c, aes(xintercept = Position, color = SNP), linetype = "solid", color = "red") +
  geom_text_repel(data = snps_157c, aes(x = Position, y = as.numeric(Scaffold) + 0.2, label = SNP),
                  inherit.aes = FALSE, angle = 45, size = 2,
                  nudge_y = 0.1,   # Adjust this value as needed
                  direction = "y") +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(), panel.spacing = unit(0.1, "lines"),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.text = element_text(size = 12, face = "bold", hjust = 0.5)) +
  scale_x_continuous(labels = label_mb, breaks = scales::pretty_breaks(n = 6)) +
  labs(x = "Position", title = "Continuous Stretches of Linked SNPs by Scaffold - Cluster 6 AUT", fill = "Linked") +
  guides(fill = "none")

# Save Venn diagram to PDF
output_path <- here("output", "ldna", "figures", "cluster_6_aut_scaffolds_with_small_windows.pdf")
ggsave(output_path, height = 5, width = 8, units = "in")
```

Calculate size
```{r calculate-stretch-sizes}
# Calculate the length of each stretch
rect_data <- rect_data %>%
  mutate(length = end - start + 1)

# Calculate average sizes for TRUE and FALSE stretches
average_sizes <- rect_data %>%
  group_by(linked) %>%
  summarize(average_size = mean(length))

# Display the average sizes
print(average_sizes)
```



We can also check what genes are in these scaffolds.

Import the data
```{r import-snps-genes-chr}
snps_genes_chr <- readRDS(here("output", "ldna", "snps_genes_chr.rds"))
```

The expression data
```{r import-gene-expression}
gene_expression <- read_delim(here("data", "files","MANvsAUTO_sig_mRNAs.csv"), delim = ",", col_names = TRUE, show_col_types = FALSE) |>
  dplyr::select(
    gene,log2FoldChange
  ) |>
  dplyr::rename(
    Gene_ID = gene
  )
head(gene_expression)
```

For example, what genes are in the cluster 14, so we can check what genes are in these scaffolds or more specifically what SNPs are in the genes

First, what genes are in these scaffolds?

```{r get-cluster-14-genes}
cluster_14_genes <- snps_genes_chr[snps_genes_chr$Scaffold %in% cluster_14_all$Scaffold, ]

# How many genes in the scaffolds
length(unique(cluster_14_genes$Gene_ID))
```


Now lets check what genes have the linked SNPs

```{r get-cluster-14-linked-genes}
cluster_14_snps2 <- snps_genes_chr[snps_genes_chr$SNP %in% aut_ch2$`3139_0.78`, ]

# How many genes with SNPs from the ld block
length(unique(cluster_14_snps2$Gene_ID))
```

Save as Excel file
```{r save-cluster-14-snps-excel}
# Save the data frame to an Excel file
write_xlsx(cluster_14_snps2, here("output", "ldna", "cluster_14_snps2.xlsx"))
```



We can count how many linked SNPs per gene
```{r count-snps-per-gene}
# Count the number of SNPs for each Gene_ID
snp_count_per_gene <- cluster_14_snps2 %>%
  group_by(Gene_ID) %>%
  summarise(SNP_count = n())

# View the first few rows of the result
head(snp_count_per_gene)
```

Save as Excel file
```{r save-snp-count-per-gene}
# Save the data frame to an Excel file
write_xlsx(snp_count_per_gene, here("output", "ldna", "snp_count_per_gene.xlsx"))
```


Now we can check how many DE genes are among the 109 genes that have SNPs on the cluster 14
```{r check-de-genes-in-cluster-14}
cluster_14_snps3 <- cluster_14_snps2[cluster_14_snps2$Gene_ID %in% gene_expression$Gene_ID, ]

# How many genes in the scaffolds
length(unique(cluster_14_snps3$Gene_ID))
unique(cluster_14_snps3$Gene_ID)
```

Get the 17 SNPs position
```{r get-17-snps-positions}
# Get the 17 SNPs
genes_17_snps <- snps_genes_chr |>
  dplyr::filter(SNP %in% snps_157b$SNP)

genes_17_snps <- snps_genes_chr[snps_genes_chr$SNP %in% snps_157b$SNP, ]

head(genes_17_snps)
```

Save as Excel file
```{r save-17-snps-excel}
# Save the data frame to an Excel file
write_xlsx(genes_17_snps, here("output", "ldna", "cluster_14_17_snps.xlsx"))
```


Get the SNPs for which we do not have genes
```{r filter-snps-without-genes}
genes_17_snps2 <- snps_157b |>
  dplyr::filter(!(SNP %in% genes_17_snps$SNP))
head(genes_17_snps2)
```

Save it (17 genes)
```{r save-snps-157b}
write.table(
  snps_157b,
  file      = here(
    "output", "ldna", "snps_157b.txt"
  ),
  sep       = "\t",
  row.names = FALSE,
  col.names = FALSE,
  quote     = FALSE
)
```
