---
title: "Autogenous - LDna data preparation"
author: "Luciano V Cosme"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: breezedark
    css:
      - "styles.css"
    toc: yes
    toc_float: no
    toc_depth: 5
editor_options:
  markdown:
    wrap: 120
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval                        = TRUE,
  echo                        = TRUE,
  cache                       = FALSE,
  class.output                = "bg-success"
)
knitr::opts_knit$set(
  root.dir = here::here()
)
```

> **Population label key:** The code labels used in the analysis below correspond to the following manuscript labels: AUT = AUTO, MAN = NON-AUTO, NEW = NON-AUTO-FIELD.

```{r libraries, message=FALSE, warning=FALSE, results='hide'}
# Load tidyverse FIRST to avoid rlang version conflict
library(tidyverse)
library(here)
library(scales)
library(ggrepel)

# Then add LDna v.2.15 library path and load
.libPaths(c("/workspace/lib/ldna_v215", .libPaths()))
library(LDna)
library(reshape2)
library(writexl)
```


## 2. Data Tidying

We will have to estimate LD with Plink and prepare the data to use it with LDna. The first step is to decided with SNP set we will use. For example, we did LD pruning for some analysis, so we should not use those files to start the linkage network analysis. We also have to decided about the minor allele frequency threshold we will use.

We can remove the individuals that failed some of our tests. One mosquito, AUT 399, failed the heterozygosity test. None failed the relatedness test, and 867 SNPs failed the HWE test.
We can start with file4 from the quality control. We can remove the mosquito and the SNPs that did not pass the HWE test.

Create file to remove mosquito

```{bash}
echo "AUT 399" > output/quality_control/remove_aut_399.txt
```

Remove mosquito and SNPs
```{bash}
plink2 \
--allow-extra-chr \
--bfile output/quality_control/file4 \
--remove output/quality_control/remove_aut_399.txt \
--extract output/quality_control/passed_hwe.txt \
--make-bed \
--out output/ldna/files/file1 \
--silent;

grep "samples\|variants" output/ldna/files/file1.log
```

The next step is to create the chromosomal scale for this data set. Check the bim file now to see that it is using the scaffold scale

```{bash}
head output/ldna/files/file1.bim
```

### 2.1 Create chromosomal scale

Import the .bim file with the SNPs to create a new chromosomal scale.

```{r import-bim-file1}
# Import the function
source(
  here(
    "notebooks", "helpers", "import_bim.R")
)

# Import the data
snps <- import_bim(here("output", "ldna", "files", "file1.bim"))

# Check it
head(snps)
```


Separate the tibbles into each chromosome.

```{r separate_data_into_chroms}
#   ____________________________________________________________________________
#   separate the SNP data per chromosome                                    ####
# chr1
chr1_snps <-
  snps |>
  filter(
    str_detect(
      Scaffold, "^1."
    )
  ) |> # here we get only Scaffold rows starting with 1
  as_tibble() # save as tibble
#
# chr2
chr2_snps <-
  snps |>
  filter(
    str_detect(
      Scaffold, "^2."
    )
  ) |>
  as_tibble()
#
# chr3
chr3_snps <-
  snps |>
  filter(
    str_detect(
      Scaffold, "^3."
    )
  ) |>
  as_tibble()
```

Import the file with sizes of each scaffold.

```{r import-scaffold-sizes}
#   ____________________________________________________________________________
#   import the file with the scaffold sizes                                 ####
sizes <-
  read_delim(
    here(
      "data", "genome", "scaffold_sizes.txt"
    ),
    col_names      = FALSE,
    show_col_types = FALSE,
    col_types      = "cd"
  )
#
# set column names
colnames(
  sizes
) <- c(
  "Scaffold", "Size"
)
#   ____________________________________________________________________________
#   create new column with the chromosome number                            ####
sizes <-
  sizes |>
  mutate(
    Chromosome = case_when( # we use mutate to create a new column called Chromosome
      startsWith(
        Scaffold, "1"
      ) ~ "1", # use startsWith to get Scaffold rows starting with 1 and output 1 on Chromosome column
      startsWith(
        Scaffold, "2"
      ) ~ "2",
      startsWith(
        Scaffold, "3"
      ) ~ "3"
    )
  ) |>
  arrange(
    Scaffold
  )                   # to sort the order of the scaffolds, fixing the problem we have with scaffold 1.86
# check it
head(sizes)
```

Create new scale. Get the scaffolds for each chromosome.

```{r get_list_scaffolds_per-chr}
#   ____________________________________________________________________________
#   separate the scaffold sizes tibble per chromosome                       ####
# chr1
chr1_scaffolds <-
  sizes |>
  filter(
    str_detect(
      Scaffold, "^1" # we use library stringr to get scaffolds starting with 1 (chromosome 1)
    )
  ) |>
  as_tibble()
#
# chr2
chr2_scaffolds <-
  sizes |>
  filter(
    str_detect(
      Scaffold, "^2" # we use library stringr to get scaffolds starting with 2 (chromosome 2)
    )
  ) |>
  as_tibble()
#
# # chr3
chr3_scaffolds <-
  sizes |>
  filter(
    str_detect(
      Scaffold, "^3" # we use library stringr to get scaffolds starting with 3 (chromosome 3)
    )
  ) |>
  as_tibble()
```

Create a scale for each chromosome.

```{r create_chr_scale}
#   ____________________________________________________________________________
#   create a new scale for each chromosome                                  ####
# chr1
chr1_scaffolds$overall_size_before_bp <-
  0                                                                        # we create a new column with zeros
for (i in 2:nrow(
  chr1_scaffolds
)
) {                                                                        # loop to start on second line
  chr1_scaffolds$overall_size_before_bp[i] <-                              # set position on the scale
    chr1_scaffolds$overall_size_before_bp[i - 1] + chr1_scaffolds$Size[i - # add the scaffold size and the location to get position on new scale
      1]
}
#
# chr2
chr2_scaffolds$overall_size_before_bp <- 0
for (i in 2:nrow(
  chr2_scaffolds
)
) {
  chr2_scaffolds$overall_size_before_bp[i] <-
    chr2_scaffolds$overall_size_before_bp[i - 1] + chr2_scaffolds$Size[i -
      1]
}
#
# chr3
chr3_scaffolds$overall_size_before_bp <- 0
for (i in 2:nrow(
  chr3_scaffolds
)
) {
  chr3_scaffolds$overall_size_before_bp[i] <-
    chr3_scaffolds$overall_size_before_bp[i - 1] + chr3_scaffolds$Size[i -
      1]
}
```

Merge the data frames scaffolds and SNPs.

```{r merge_snps_scales}
#   ____________________________________________________________________________
#   merge the data sets using the tidyverse function left_join              ####
# chr1
chr1_scale <-
  chr1_snps |>          # create data frame for each chromosome, get chr1_snps
  left_join(            # use lef_join function to merge it with chr1_scaffolds
    chr1_scaffolds,
    by = "Scaffold"
  ) |>                  # set column to use for merging (Scaffold in this case)
  na.omit() |>          # remove NAs, we don't have SNPs in every scaffold
  mutate(
    midPos_fullseq = as.numeric(
      Position
    ) +                 # make new columns numeric
      as.numeric(
        overall_size_before_bp
      )
  )
#
# chr2
chr2_scale <-
  chr2_snps |>
  left_join(
    chr2_scaffolds,
    by = "Scaffold"
  ) |>
  na.omit() |>
  mutate(
    midPos_fullseq = as.numeric(
      Position
    ) +
      as.numeric(
        overall_size_before_bp
      )
  )
#
# chr3
chr3_scale <-
  chr3_snps |>
  left_join(
    chr3_scaffolds,
    by = "Scaffold"
  ) |>
  na.omit() |>
  mutate(
    midPos_fullseq = as.numeric(
      Position
    ) +
      as.numeric(
        overall_size_before_bp
      )
  )
```

Merge all chromosome scales.

```{r bind_chr_scales}
#   ____________________________________________________________________________
#   merge the data sets, and select only the columns we need                ####
chroms <- rbind(
  chr1_scale, chr2_scale, chr3_scale
) |>
  dplyr::select(
    Chromosome, SNP, Cm, midPos_fullseq, Allele1, Allele2
  )
# check it
head(chroms)
```

Save the new .bim file

```{r save-new-bim-file}
#   ____________________________________________________________________________
#   save the new bim file with a new name, I added "B"                      ####
write.table(
  chroms,
  file      = here(
    "output", "ldna", "files", "file1B.bim"
  ),
  sep       = "\t",
  row.names = FALSE,
  col.names = FALSE,
  quote     = FALSE
)
```

Rename the .bim files
```{bash}
# change the name of the first .bim file, for example, append _backup.bim, and then replace the original file
mv output/ldna/files/file1.bim output/ldna/files/file1_backup.bim;
# than change the new bim we create to the original name (do it only once, otherwise it will mess up)
mv output/ldna/files/file1B.bim output/ldna/files/file1.bim
```

Create a new bed file with Plink2 to see if it works. For example, to see if the variants are in the
right order. Plink2 will give us a warning.

```{bash}
plink2 \
--bfile output/ldna/files/file1 \
--make-bed \
--out output/ldna/test01;
# then we remove the files
rm output/ldna/test01.*
```
No warnings from Plink2. Now, we can go ahead with our analysis.


Clean env and memory
```{r cleanup-memory-before-subset}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```

### 2.2 Subset by family

Estimate frequency
```{bash}
plink \
--keep-allele-order \
--bfile output/ldna/files/file1 \
--make-bed \
--freqx \
--out output/ldna/files/frq \
--silent;

grep "people\|variants" output/ldna/files/frq.log
```

Now we can use bash to check if there SNPs with low heterozygosity
```{bash}
# count SNPs with het < 0.5 (we have 60 mosquitoes)
cat output/ldna/files/frq.frqx | awk '{ if ($6 <= 30) print }' | wc -l
```

How many SNPs
```{r count-low-het-snps}
110353 - 104625
```
We can remove this SNPs since a limiting factor for the linkage network analysis is memory

Create a file with the SNPs
```{bash}
# get list of snps
cat output/ldna/files/frq.frqx | awk '{ if ($6 <= 30) print }' | awk '{print $2}' > output/ldna/files/snps_het.txt
```

We can start our analysis with MAF of 5%. If we have memory issues we can use 10% next. We will also set the genotyping missingness to zero.
```{bash}
# Filter MAF 5% and do not allow missing genotypes
plink \
--keep-allele-order \
--bfile output/ldna/files/file1 \
--out output/ldna/files/file2 \
--maf 0.05 \
--geno 0 \
--make-bed \
--extract output/ldna/files/snps_het.txt \
--silent;

grep "people\|variants" output/ldna/files/file2.log
```

Check the fam file families
```{bash}
# get the list of families
cat output/ldna/files/file2.fam | awk '!seen[$1]++' | awk '{print $1}'
```

We can check how many mosquitoes per population as well
```{bash get_pops_n, cache=TRUE}
awk '{print $1}' output/ldna/files/file2.fam | sort | uniq -c | awk '{print $2, $1}'
```

I did the analysis using the 3 populations, however the results of MAN were odd because we have only 10 mosquitoes. Therefore, I decided not to do linkage network analysis with MAN. I finished the analysis to see the odd pattern, but I came back and changed the code from this part foward to include only AUT and NEW for the linkage network analysis. I will still create files for each family, however the I will use only the shared SNPs between NEW and AUT. We will not consider MAN. I left it in the code, but we will not use it for comparisons.

Create a file for each of them
```{bash}
# make a text file with the name of each population
for pop in $(cat output/ldna/files/file2.fam | awk '!seen[$1]++' | awk '{print $1}');
do
  echo $pop > output/ldna/files/$pop\.txt
done
```

Now use Plink to create a file for each family. Now we set a MAF threshold of 5% within each family
```{bash}
# use plink to create a plink file for each population
for pop in $(cat output/ldna/files/file2.fam | awk '!seen[$1]++' | awk '{print $1}');
do
  plink --keep-allele-order --allow-no-sex --bfile output/ldna/files/file2 --make-bed --keep-fam output/ldna/files/$pop\.txt --out output/ldna/files/$pop --geno 0 --maf 0.05 --silent
done
```

Now we can check the number of SNPs and samples in each file

AUT
```{bash}
grep "people\|variants" output/ldna/files/AUT.log
```

MAN
```{bash}
grep "people\|variants" output/ldna/files/MAN.log
```

NEW
```{bash}
grep "people\|variants" output/ldna/files/NEW.log
```


We can see that the number of SNPs kept is not the same. We need to get the shared SNPs. We can import the bim files and use R to get the intersect

```{r import-pop-bim-files}
# Import the function
source(
  here(
    "notebooks", "helpers", "import_bim.R")
)

# Import the data
AUT <- import_bim(here("output", "ldna", "files", "AUT.bim"))
MAN <- import_bim(here("output", "ldna", "files", "MAN.bim")) # not using MAN because it has only 10 mosquitoes
NEW <- import_bim(here("output", "ldna", "files", "NEW.bim"))
```

Now get the shared SNPs

```{r find-common-snps}
# Identify common SNPs
# common_snps <- Reduce(intersect, list(AUT$SNP, MAN$SNP, NEW$SNP))
common_snps <- Reduce(intersect, list(AUT$SNP, NEW$SNP))

# Create a data frame with the common SNPs
common_snps_df <- data.frame(SNP = common_snps)

# Count them
length(common_snps)

# Write the data frame to a file
write.table(common_snps_df, here("output", "ldna", "files","common_snps.txt"), quote = FALSE, row.names = FALSE, col.names = FALSE)
```

We have 32,732 SNPs. It seems we remove a lot but still too much for LDna in a laptop. We can use the cluster but first I will try using a laptop.

Now we have to repeat the previous step when we create a file for each family but now using only the SNPs that have 5% MAF in each family

```{bash}
# use plink to create a plink file for each population
for pop in $(cat output/ldna/files/file2.fam | awk '!seen[$1]++' | awk '{print $1}');
do
  plink --keep-allele-order --allow-no-sex --bfile output/ldna/files/file2 --make-bed --keep-fam output/ldna/files/$pop\.txt --out output/ldna/files/$pop --extract output/ldna/files/common_snps.txt --silent
done
```


Now we can check the number of SNPs and they all should be the same

AUT
```{bash}
grep "people\|variants" output/ldna/files/AUT.log
```

MAN
```{bash}
grep "people\|variants" output/ldna/files/MAN.log
```

NEW
```{bash}
grep "people\|variants" output/ldna/files/NEW.log
```

Perfect. Now we can split the data by chromosome for each family.

### 2.3 Subset by chromosome within each family

We can create a new directory
```{bash}
# Create a directory
mkdir -p output/ldna/pop;

# We can copy the family files there
cp output/ldna/files/AUT* output/ldna/pop;
cp output/ldna/files/MAN* output/ldna/pop;
cp output/ldna/files/NEW* output/ldna/pop;

# remove the files we dont need
rm output/ldna/pop/*.txt;
rm output/ldna/pop/*.nosex
```

We can now get a list of SNPs for each chromosome. All the populations have the same SNPs, so we can use any of the files.

```{bash}
cat output/ldna/pop/AUT.bim | awk '$1 == 1' | awk '{print $2}' > output/ldna/pop/chr1_snps.txt;
cat output/ldna/pop/AUT.bim | awk '$1 == 2' | awk '{print $2}' > output/ldna/pop/chr2_snps.txt;
cat output/ldna/pop/AUT.bim | awk '$1 == 3' | awk '{print $2}' > output/ldna/pop/chr3_snps.txt
```

### 2.4 Estimate LD with Plink for each chromosome

```{bash}
# chr1
for pop in $(ls -1 output/ldna/pop/*.bim | sed 's/output\/ldna\/pop\///' | sed 's/\.[^.]*$//');
do
  plink --allow-no-sex --keep-allele-order --bfile output/ldna/pop/$pop --out output/ldna/pop/$pop\.chr1 --r2 triangle gz --extract output/ldna/pop/chr1_snps.txt --silent
done;

# chr2
for pop in $(ls -1 output/ldna/pop/*.bim | sed 's/output\/ldna\/pop\///' | sed 's/\.[^.]*$//');
do
  plink --allow-no-sex --keep-allele-order --bfile output/ldna/pop/$pop --out output/ldna/pop/$pop\.chr2 --r2 triangle gz --extract output/ldna/pop/chr2_snps.txt --silent
done;

# chr3
for pop in $(ls -1 output/ldna/pop/*.bim | sed 's/output\/ldna\/pop\///' | sed 's/\.[^.]*$//');
do
  plink --allow-no-sex --keep-allele-order --bfile output/ldna/pop/$pop --out output/ldna/pop/$pop\.chr3 --r2 triangle gz --extract output/ldna/pop/chr3_snps.txt --silent
done
```

Create directory for each chromosome and move the files
```{bash}
# make dirs and remove unnecessary files
rm output/ldna/pop/*.nosex;
mkdir -p output/ldna/pop/chr1; mv output/ldna/pop/*chr1.* output/ldna/pop/chr1/;
mkdir -p output/ldna/pop/chr2; mv output/ldna/pop/*chr2.* output/ldna/pop/chr2/;
mkdir -p output/ldna/pop/chr3; mv output/ldna/pop/*chr3.* output/ldna/pop/chr3/;

# copy files
cp output/ldna/pop/chr1*.txt output/ldna/pop/chr1/;
cp output/ldna/pop/chr2*.txt output/ldna/pop/chr2/;
cp output/ldna/pop/chr3*.txt output/ldna/pop/chr3/
```



### 2.5 Prepare LD matrices

Prepare files for LDna
```{bash}
# chr1
cat output/ldna/pop/chr1_snps.txt | awk '{print $1}' > output/ldna/pop/chr1/snps1.txt;
echo "" > output/ldna/pop/chr1/snps2.txt;
cat output/ldna/pop/chr1/snps2.txt output/ldna/pop/chr1/snps1.txt | gzip -9 > output/ldna/pop/chr1/snps3.txt.gz;
cat output/ldna/pop/chr1/snps1.txt | tr '\n' ' ' |  awk -v OFS='\t' '{$1=$1}1' | gzip -9 > output/ldna/pop/chr1/header.txt.gz;

# chr2
cat output/ldna/pop/chr2_snps.txt | awk '{print $1}' > output/ldna/pop/chr2/snps1.txt;
echo "" > output/ldna/pop/chr2/snps2.txt;
cat output/ldna/pop/chr2/snps2.txt output/ldna/pop/chr2/snps1.txt | gzip -9 > output/ldna/pop/chr2/snps3.txt.gz;
cat output/ldna/pop/chr2/snps1.txt | tr '\n' ' ' |  awk -v OFS='\t' '{$1=$1}1' | gzip -9 > output/ldna/pop/chr2/header.txt.gz;

# chr3
cat output/ldna/pop/chr3_snps.txt | awk '{print $1}' > output/ldna/pop/chr3/snps1.txt;
echo "" > output/ldna/pop/chr3/snps2.txt;
cat output/ldna/pop/chr3/snps2.txt output/ldna/pop/chr3/snps1.txt | gzip -9 > output/ldna/pop/chr3/snps3.txt.gz;
cat output/ldna/pop/chr3/snps1.txt | tr '\n' ' ' |  awk -v OFS='\t' '{$1=$1}1' | gzip -9 > output/ldna/pop/chr3/header.txt.gz;
```

Add header to the matrices
```{bash}
# chr1
for pop in $(ls -1 output/ldna/pop/chr1/*.ld.gz | sed 's/output\/ldna\/pop\/chr1\///' | sed 's/\.[^.]*$//');
do
  cat output/ldna/pop/chr1/header.txt.gz output/ldna/pop/chr1/$pop\.gz > output/ldna/pop/chr1/$pop\.txt.gz
done;

# chr2
for pop in $(ls -1 output/ldna/pop/chr2/*.ld.gz | sed 's/output\/ldna\/pop\/chr2\///' | sed 's/\.[^.]*$//');
do
  cat output/ldna/pop/chr2/header.txt.gz output/ldna/pop/chr2/$pop\.gz > output/ldna/pop/chr2/$pop\.txt.gz
done;

# chr3
for pop in $(ls -1 output/ldna/pop/chr3/*.ld.gz | sed 's/output\/ldna\/pop\/chr3\///' | sed 's/\.[^.]*$//');
do
  cat output/ldna/pop/chr3/header.txt.gz output/ldna/pop/chr3/$pop\.gz > output/ldna/pop/chr3/$pop\.txt.gz
done
```


We can rename the files
```{bash}
# rename the files (remove the ld)
for f in output/ldna/pop/chr1/*ld.txt.gz; do [ -f "$f" ] && mv "$f" "${f/ld.txt.gz/txt.gz}"; done
for f in output/ldna/pop/chr2/*ld.txt.gz; do [ -f "$f" ] && mv "$f" "${f/ld.txt.gz/txt.gz}"; done
for f in output/ldna/pop/chr3/*ld.txt.gz; do [ -f "$f" ] && mv "$f" "${f/ld.txt.gz/txt.gz}"; done
```

Now we can add the row names (this takes time and the output files are near 1Gb)
```{bash}
# chr1
for pop in $(ls -1 output/ldna/pop/chr1/*.chr1.txt.gz | sed 's/output\/ldna\/pop\/chr1\///' | sed 's/\.chr1\.txt\.gz$//');
do
  paste <(gzip -d < output/ldna/pop/chr1/snps3.txt.gz) <(gzip -d < output/ldna/pop/chr1/$pop.chr1.txt.gz) > output/ldna/pop/chr1/${pop}.chr1.txt
done

# chr2
for pop in $(ls -1 output/ldna/pop/chr2/*.chr2.txt.gz | sed 's/output\/ldna\/pop\/chr2\///' | sed 's/\.chr2\.txt\.gz$//');
do
  paste <(gzip -d < output/ldna/pop/chr2/snps3.txt.gz) <(gzip -d < output/ldna/pop/chr2/$pop.chr2.txt.gz) > output/ldna/pop/chr2/${pop}.chr2.txt
done

# chr3
for pop in $(ls -1 output/ldna/pop/chr3/*.chr3.txt.gz | sed 's/output\/ldna\/pop\/chr3\///' | sed 's/\.chr3\.txt\.gz$//');
do
  paste <(gzip -d < output/ldna/pop/chr3/snps3.txt.gz) <(gzip -d < output/ldna/pop/chr3/$pop.chr3.txt.gz) > output/ldna/pop/chr3/${pop}.chr3.txt
done
```

I deleted the MAN files for chromosome 2 and 3 since we will not estimate LD for them. I kept chromosome 1 to show how small sample size bias the results.


