---
title: "Autogenous - LDna chromosome 3"
author: "Luciano V Cosme"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: breezedark
    css:
      - "styles.css"
    toc: yes
    toc_float: no
    toc_depth: 5
editor_options:
  markdown:
    wrap: 120
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval                        = TRUE,
  echo                        = TRUE,
  cache                       = FALSE,
  class.output                = "bg-success"
)
knitr::opts_knit$set(
  root.dir = here::here()
)
```

> **Population label key:** The code labels used in the analysis below correspond to the following manuscript labels: AUT = AUTO, MAN = NON-AUTO, NEW = NON-AUTO-FIELD.

```{r libraries, message=FALSE, warning=FALSE, results='hide'}
# Load tidyverse FIRST to avoid rlang version conflict
library(tidyverse)
library(here)
library(scales)
library(ggrepel)

# Then add LDna v.2.15 library path and load
.libPaths(c("/workspace/lib/ldna_v215", .libPaths()))
library(LDna)
library(reshape2)
library(writexl)
```


## 5. LDna chromosome 3

Because MAN has only 10 mosquitoes we will not estimate LD for it

### 5.1 NEW (chromosome 3)
Clean env and memory
```{r cleanup-memory-new}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```

Import the data (make sure you have enough memory)
```{r load-new-ld-matrix, eval=TRUE}
ld1 <-
  read.delim(
    here("output", "ldna", "pop", "chr3", "NEW.chr3.txt"),
    sep = "\t",
    header = T,
    row.names = 1,
    stringsAsFactors = F,
    check.names = FALSE
  )
```

Check the LD matrix we built using bash
```{r check-new-ld-matrix, eval=TRUE}
head(ld1)
```

The row and column names are correct. Now we can convert the data frame to matrix

```{r convert-new-to-matrix, eval=TRUE}
# Create the matrix for ldna
ld2<- as.matrix(sapply(ld1, as.numeric))

# Get names
names<- rownames(ld1)
#names<-gsub("-", ".", names)

# Add names
row.names(ld2)<-names

# Clean non-numeric values
ld2[is.nan(ld2)] <- NA
ld2[is.infinite(ld2)] <- NA

# Remove diagonal values = 1
diag(ld2)=NA

# Make sure it is only the lower triangle matrix
ld2[!lower.tri(ld2)] <- NA
```

Create a LDna object
```{r run-new-ldna-raw, eval=TRUE}
# create the LDna object (change the number of cores as needed)
diag(ld2) <- NA
ld2[!lower.tri(ld2)] <- NA
ldna <- LDnaRaw(ld2, mc.cores=2, method = "single")
```

Save ldna
```{r save-new-ldna, eval=TRUE}
saveRDS(ldna, file = here("output", "ldna", "pop", "chr3", "NEW.rds"))
saveRDS(ld2, file = here("output", "ldna", "pop", "chr3", "NEW_ld2.rds"))
```

```{r load-new-ldna-results}
ldna <- readRDS(here("output", "ldna", "pop", "chr3", "NEW.rds"))
ld2 <- readRDS(here("output", "ldna", "pop", "chr3", "NEW_ld2.rds"))
```


Check the object
```{r check-new-ldna-str}
str(ldna)
```

Extract clusters using min.edges=100
```{r extract-new-branches}
pdf(file = here("output", "ldna", "pop", "chr3", "NEW_extractBranches.pdf"), width = 20, height = 12)
op <- par(mfcol=c(1,4))

clusters <- extractBranches(ldna, min.edges=100, plot.tree=TRUE, cores=1)
summary <- summaryLDna(ldna, clusters, ld2)
write.table(summary, file = here("output", "ldna", "pop", "chr3", "summary_NEW.txt"), row.names=FALSE, sep="\t", quote=FALSE)

par(op)
while (!is.null(dev.list())) dev.off()
```

Get cluster information
```{r get-new-cluster-info}
# to get the list of names of the clusters
# Find the length of the longest vector
max_length <- max(sapply(clusters, length))

# Function to pad vectors with NAs to make them the same length
pad_vector_to_max_length <- function(vec, max_length) {
  c(vec, rep(NA, max_length - length(vec)))
}

# Apply the padding function to each vector in the list and combine into a data frame
do1 <- as.data.frame(do.call(cbind, lapply(clusters, pad_vector_to_max_length, max_length)))

# to get list of snps for each cluster
do <- do.call(cbind, unname(lapply(clusters, `length<-`, max(lengths(clusters)))))
colnames(do) <- colnames(do1)
```

Save it
```{r save-new-clusters-df}
max_length <- max(sapply(clusters, length))
# Initialize an empty list
cluster_df_list <- list()

# Loop through each cluster and pad with NAs
for (cluster_name in names(clusters)) {
  cluster_length <- length(clusters[[cluster_name]])
  padded_cluster <- c(clusters[[cluster_name]], rep(NA, max_length - cluster_length))
  cluster_df_list[[cluster_name]] <- padded_cluster
}

# Combine the lists into a data frame
df <- data.frame(cluster_df_list)
# Remove 'X' from column names
names(df) <- make.names(names(df), unique = FALSE)
names(df) <- sub("X", "", names(df), fixed = TRUE)

# Save it
saveRDS(df, file = here("output", "ldna", "pop", "chr3", "NEW_clusters_snps.rds"))
head(df)
```
We can save it in a different format as well
```{r save-new-clusters-long-format}
# Assuming your data frame is named df
long_df <- gather(df, Cluster, SNP, everything())

# View the first few rows of the transformed data
head(long_df)

# Save it
saveRDS(long_df, file = here("output", "ldna", "pop", "chr3", "NEW_clusters_snps1.rds"))
```


Make long format
```{r melt-new-clusters}
new_chr3 <- melt(do, na.rm = T, value.name = "value")
head(new_chr3)
```


Update names
```{r update-new-cluster-colnames}
colnames(new_chr3)<- c("v1", "cluster", "SNP")
head(new_chr3)
```



Import the bim file to get the SNP positions
```{r import-new-bim-positions}
# Import the function
source(
  here(
    "notebooks", "helpers", "import_bim.R")
)

# Import the data
new_snps_chr <- import_bim(here("output", "ldna", "pop", "NEW.bim")) |>
  dplyr::select(
    SNP, Scaffold, Position
  ) |>
  dplyr::rename(
    Chromosome = Scaffold
  )

# Check it
head(new_snps_chr)
```

Merge the cluster data and the SNP data
```{r merge-new-clusters-snps}
# merge dataframes
merged<- merge(new_chr3, new_snps_chr, by="SNP", all.x=TRUE)
merged<-na.omit(merged)
head(merged)
```

Select the columns we need
```{r select-new-cluster-columns}
# subset
merged <- subset(merged, select = c(cluster, SNP, Position))
head(merged)
```

Sort by distance
```{r sort-new-by-position}
merged<- merged[order(merged$Position),,drop=FALSE]
head(merged)
```


```{r subset-new-sushi}
# for parsing later
sushi1<- subset(merged, select = c(cluster, Position))
head(sushi1)
```

Save the data
```{r save-new-cluster-data, warning=FALSE}
# Create directory
new_directory_path <- here("output", "ldna", "pop", "chr3", "clusters")
dir.create(new_directory_path, recursive = TRUE)

# save file
write.table(merged, file = here("output", "ldna", "pop", "chr3", "SNPs_clusters_NEW_chr3.txt"), row.names = F, sep = "\t", quote = F)
write.table(sushi1, file = here("output", "ldna", "pop", "chr3", "clusters", "sushi_NEW_chr3.txt"), row.names = F, sep = "\t", quote = F)
```

Parse the file1 to get the start and end of the clusters as well as their size. Remember, recombination make it a mosaic, so the clusters are "mixed".
```{bash}
awk '
function print_row() {
    if ( feature != "" )
       print feature, start, end, (end - start)
}

BEGIN {
    FS=OFS="\t";
    print "Cluster", "Start", "End", "Size";
}
NR == 1 { next } # Skip the first line
$1 != feature {
    print_row();
    feature = $1;
    start = $2;
    end = $2;
    next;
}
{
    end = $2;
}
END {
    print_row();
}
' output/ldna/pop/chr3/clusters/sushi_NEW_chr3.txt > output/ldna/pop/chr3/clusters/sushi2_NEW_chr3.txt;
head output/ldna/pop/chr3/clusters/sushi2_NEW_chr3.txt
```

```{r count-new-segments}
# Define the path
input_path <- here("output", "ldna", "pop", "chr3", "clusters", "sushi2_NEW_chr3.txt")

# Read the data into R
clusters_data <- read_table(input_path, col_names = TRUE, col_types = NULL)

# Arrange by position
clusters_data <- clusters_data |>
  arrange(Start)

# Count segments for each cluster - if you want only the counts
segment_counts <- clusters_data %>%
  group_by(Cluster) %>%
  summarize(nSegments = n(), .groups = 'drop')


# View the result
head(segment_counts)
```


Plot it
```{r plot-new-clusters-basic, fig.height=5, fig.width=8}
# Define the path
input_path <-
  here("output",
       "ldna",
       "pop",
       "chr3",
       "clusters",
       "sushi2_NEW_chr3.txt")

# Read the data into R
clusters_data <- read.table(input_path, header = TRUE, sep = "\t") |>
  arrange(Start)

# Filter out rows with Size equal to 0
clusters_data <- subset(clusters_data, Size > 100000) # show only bigger than 100kb


# Calculate the maximum size for scaling
max_size <- max(clusters_data$Size)

# Create Start_Mb and End_Mb within clusters_data
clusters_data$Start_Mb <- clusters_data$Start / 1e6
clusters_data$End_Mb <- clusters_data$End / 1e6

# Create the plot
ggplot(clusters_data, aes(xmin = Start_Mb, xmax = End_Mb, ymin = 0, ymax = Size / 1e6)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = label_number(unit = "M"), breaks = pretty_breaks(n = 10)) +
  scale_y_continuous(labels = label_number(unit = "M"), breaks = pretty_breaks(n = 10)) +
  labs(x = "Chromosome 3 (Mb)", y = "Cluster Size (Mb)", fill = "Cluster ID") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    panel.grid.minor.x = element_blank(),
    legend.position = "top",
    legend.box = "horizontal"
  ) +
  guides(fill = guide_legend(nrow = 3, title.position = "top", title.hjust = 0.5))


# Use ggsave to save the plot as a PDF
ggsave(
  filename = here("output", "ldna", "pop", "chr3", "clusters", "NEW_chr3.pdf"),
  device = "pdf",
  width = 8,
  height = 5,
  units = "in"
)
```


We can annotate the number of segments in the legend.

```{r plot-new-clusters-annotated, fig.height=5, fig.width=8}
# Merge SNP counts into clusters_data while specifying suffixes
annotated_data <- merge(clusters_data, segment_counts, by = "Cluster")

# Create a new column with Cluster names and SNP counts
annotated_data$Cluster_with_SNPs <- with(annotated_data, paste(Cluster, " (", nSegments, ")", sep=""))

# Now the plot uses the new Cluster_with_SNPs for the fill legend
ggplot(annotated_data, aes(xmin = Start_Mb, xmax = End_Mb, ymin = 0, ymax = Size / 1e6)) +
  geom_rect(aes(fill = as.factor(Cluster_with_SNPs)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = scales::label_number(unit = "M"), breaks = scales::pretty_breaks(n = 10)) +
  scale_y_continuous(labels = scales::label_number(unit = "M"), breaks = scales::pretty_breaks(n = 10)) +
  labs(x = "Chromosome 3 (Mb)", y = "Cluster Size (Mb)", fill = "Cluster ID") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    panel.grid.minor.x = element_blank(),
    legend.position = "top",
    legend.box = "horizontal"
  ) +
  guides(fill = guide_legend(nrow = 3, title.position = "top", title.hjust = 0.5))

# Use ggsave to save the plot as a PDF
ggsave(
  filename = here("output", "ldna", "pop", "chr3", "clusters", "NEW_chr3b.pdf"),
  device = "pdf",
  width = 8,
  height = 5,
  units = "in"
)
```



We can save the data to plot all chromosomes togeter using a facet plot
```{r check-new-annotated-data}
head(annotated_data)
```

Save plotting data, we can arrange the column order first
```{r save-new-plot-data}
plot_data <- annotated_data |>
  dplyr::select(
    Cluster, nSegments, Start, End, Size, Start_Mb, End_Mb
  )

# Create a new column with Chromosome
plot_data <- data.frame(Chromosome = rep(1, nrow(plot_data)), plot_data)

# Split the 'Cluster' column at "_", and create the 'r2' column with the second part
plot_data$r2 <- sapply(strsplit(as.character(plot_data$Cluster), "_"), `[`, 2)

# Move the 'r2' column to be the second column, right after 'Chromosome'
plot_data <- plot_data[, c(1:2, ncol(plot_data), 3:(ncol(plot_data)-1))]


# Arrange
plot_data <- plot_data |>
  dplyr::arrange(Start)

# Check it
head(plot_data)

# Save it
saveRDS(plot_data, file = here("output", "ldna", "pop", "chr3", "NEW_plot.rds"))
```

We can also import the summary file to create a object with SNP counts per segment
```{r save-new-cluster-seg-snp}
# Define the path
input_path <-
  here("output",
       "ldna",
       "pop",
       "chr3",
       "summary_NEW.txt")

# Read the data into R
clusters_snps <- read.table(input_path, header = TRUE, sep = "\t")

# We subset and rename the column 1
clusters_snps <- clusters_snps |>
  dplyr::select(
    Name, nLoci
  ) |>
  dplyr::rename(
    Cluster = Name,
    nSNPs = nLoci
  )

# Merge
cluster_seg_snp <- inner_join(segment_counts, clusters_snps, by = "Cluster")
cluster_seg_snp <- inner_join(clusters_data, cluster_seg_snp, by = "Cluster")

# Split the 'Cluster' column at "_", and create the 'r2' column with the second part
cluster_seg_snp$r2 <- sapply(strsplit(as.character(cluster_seg_snp$Cluster), "_"), `[`, 2)

# Add chromosome number and population name
cluster_seg_snp <- data.frame(Chromosome = rep(3, nrow(cluster_seg_snp)), cluster_seg_snp)
cluster_seg_snp <- data.frame(Population = rep("NEW", nrow(cluster_seg_snp)), cluster_seg_snp)


# Reorder the columns
cluster_seg_snp <- cluster_seg_snp |>
  dplyr::select(
    Chromosome, Cluster, r2, Start, End, nSegments, nSNPs
  ) |>
  dplyr::arrange(
    Chromosome, Start
  )

head(cluster_seg_snp)


# Save the data for plotting later
saveRDS(cluster_seg_snp, file = here("output", "ldna", "pop", "chr3", "NEW_plot2.rds"))
```


### 5.2 AUT (chromosome 3)
Clean env and memory
```{r cleanup-memory-aut}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```

Import the data
```{r load-aut-ld-matrix, eval=TRUE}
ld1 <-
  read.delim(
    here("output", "ldna", "pop", "chr3", "AUT.chr3.txt"),
    sep = "\t",
    header = T,
    row.names = 1,
    stringsAsFactors = F,
    check.names = FALSE
  )
```

Check the LD matrix we built using bash
```{r check-aut-ld-matrix, eval=TRUE}
head(ld1)
```

The row and column names are correct. Now we can convert the data frame to matrix

```{r convert-aut-to-matrix, eval=TRUE}
# Create the matrix for ldna
ld2<- as.matrix(sapply(ld1, as.numeric))

# Get names
names<- rownames(ld1)
#names<-gsub("-", ".", names)

# Add names
row.names(ld2)<-names

# Clean non-numeric values
ld2[is.nan(ld2)] <- NA
ld2[is.infinite(ld2)] <- NA

# Remove diagonal values = 1
diag(ld2)=NA

# Make sure it is only the lower triangle matrix
ld2[!lower.tri(ld2)] <- NA
```

Create a LDna object
```{r run-aut-ldna-raw, eval=TRUE}
# create the LDna object (reduce cores to limit memory during full render)
diag(ld2) <- NA
ld2[!lower.tri(ld2)] <- NA
gc()
ldna <- LDnaRaw(ld2, mc.cores=2, method = "single")
```

Check the object
```{r check-aut-ldna-str, eval=TRUE}
str(ldna)
```

Save it
```{r save-aut-ldna, eval=TRUE}
saveRDS(ldna, file = here("output", "ldna", "pop", "chr3", "AUT.rds"))
saveRDS(ld2, file = here("output", "ldna", "pop", "chr3", "AUT_ld2.rds"))
```

Load the data
```{r load-aut-ldna-results}
ldna <- readRDS(here("output", "ldna", "pop", "chr3", "AUT.rds"))
ld2 <- readRDS(here("output", "ldna", "pop", "chr3", "AUT_ld2.rds"))
```

Extract clusters using min.edges=100
```{r extract-aut-branches}
pdf(file = here("output", "ldna", "pop", "chr3", "AUT_extractBranches.pdf"), width = 20, height = 12)
op <- par(mfcol=c(1,4))

clusters <- extractBranches(ldna, min.edges=100, plot.tree=TRUE, cores=1)
summary <- summaryLDna(ldna, clusters, ld2)
write.table(summary, file = here("output", "ldna", "pop", "chr3", "summary_AUT.txt"), row.names=FALSE, sep="\t", quote=FALSE)

par(op)
while (!is.null(dev.list())) dev.off()
```

Get cluster information
```{r get-aut-cluster-info}
# to get the list of names of the clusters
# Find the length of the longest vector
max_length <- max(sapply(clusters, length))

# Function to pad vectors with NAs to make them the same length
pad_vector_to_max_length <- function(vec, max_length) {
  c(vec, rep(NA, max_length - length(vec)))
}

# Apply the padding function to each vector in the list and combine into a data frame
do1 <- as.data.frame(do.call(cbind, lapply(clusters, pad_vector_to_max_length, max_length)))

# to get list of snps for each cluster
do <- do.call(cbind, unname(lapply(clusters, `length<-`, max(lengths(clusters)))))
colnames(do) <- colnames(do1)
```

Save it
```{r save-aut-clusters-df}
max_length <- max(sapply(clusters, length))
# Initialize an empty list
cluster_df_list <- list()

# Loop through each cluster and pad with NAs
for (cluster_name in names(clusters)) {
  cluster_length <- length(clusters[[cluster_name]])
  padded_cluster <- c(clusters[[cluster_name]], rep(NA, max_length - cluster_length))
  cluster_df_list[[cluster_name]] <- padded_cluster
}

# Combine the lists into a data frame
df <- data.frame(cluster_df_list)
# Remove 'X' from column names
names(df) <- make.names(names(df), unique = FALSE)
names(df) <- sub("X", "", names(df), fixed = TRUE)

# Save it
saveRDS(df, file = here("output", "ldna", "pop", "chr3", "AUT_clusters_snps.rds"))
head(df)
```
We can save it in a different format as well
```{r save-aut-clusters-long-format}
# Assuming your data frame is named df
long_df <- gather(df, Cluster, SNP, everything())

# View the first few rows of the transformed data
head(long_df)

# Save it
saveRDS(long_df, file = here("output", "ldna", "pop", "chr3", "AUT_clusters_snps1.rds"))
```

Make long format
```{r melt-aut-clusters}
aut_chr2 <- melt(do, na.rm = T, value.name = "value")
head(aut_chr2)
```


Update names
```{r update-aut-cluster-colnames}
colnames(aut_chr2)<- c("v1", "cluster", "SNP")
head(aut_chr2)
```


Import the bim file to get the SNP positions
```{r import-aut-bim-positions}
# Import the function
source(
  here(
    "notebooks", "helpers", "import_bim.R")
)

# Import the data
aut_snps_chr <- import_bim(here("output", "ldna", "pop", "AUT.bim")) |>
  dplyr::select(
    SNP, Scaffold, Position
  ) |>
  dplyr::rename(
    Chromosome = Scaffold
  )

# Check it
head(aut_snps_chr)
```

Merge the cluster data and the SNP data
```{r merge-aut-clusters-snps}
# merge dataframes
merged<- merge(aut_chr2, aut_snps_chr, by="SNP", all.x=TRUE)
merged<-na.omit(merged)
head(merged)
```

Select the columns we need
```{r select-aut-cluster-columns}
# subset
merged <- subset(merged, select = c(cluster, SNP, Position))
head(merged)
```

Sort by distance
```{r sort-aut-by-position}
merged<- merged[order(merged$Position),,drop=FALSE]
head(merged)
```


```{r subset-aut-sushi}
# for parsing later
sushi1<- subset(merged, select = c(cluster, Position))
head(sushi1)
```

Save the data
```{r save-aut-cluster-data, warning=FALSE}
# Create directory
new_directory_path <- here("output", "ldna", "pop", "chr3", "clusters")
dir.create(new_directory_path, recursive = TRUE)

# save file
write.table(merged, file = here("output", "ldna", "pop", "chr3", "SNPs_clusters_AUT_chr3.txt"), row.names = F, sep = "\t", quote = F)
write.table(sushi1, file = here("output", "ldna", "pop", "chr3", "clusters", "sushi_AUT_chr3.txt"), row.names = F, sep = "\t", quote = F)
```

Parse the file1 to get the start and end of the clusters as well as their size. Remember, recombination make it a mosaic, so the clusters are "mixed".
```{bash}
awk '
function print_row() {
    if ( feature != "" )
       print feature, start, end, (end - start)
}

BEGIN {
    FS=OFS="\t";
    print "Cluster", "Start", "End", "Size";
}
NR == 1 { next } # Skip the first line
$1 != feature {
    print_row();
    feature = $1;
    start = $2;
    end = $2;
    next;
}
{
    end = $2;
}
END {
    print_row();
}
' output/ldna/pop/chr3/clusters/sushi_AUT_chr3.txt > output/ldna/pop/chr3/clusters/sushi2_AUT_chr3.txt;
head output/ldna/pop/chr3/clusters/sushi2_AUT_chr3.txt
```

Get SNP count
```{r count-aut-segments}
# Define the path
input_path <- here("output", "ldna", "pop", "chr3", "clusters", "sushi2_AUT_chr3.txt")

# Read the data into R
clusters_data <- read_table(input_path, col_names = TRUE, col_types = NULL)

# Arrange by position
clusters_data <- clusters_data |>
  arrange(Start)

# Count segments for each cluster - if you want only the counts
segment_counts <- clusters_data %>%
  group_by(Cluster) %>%
  summarize(nSegments = n(), .groups = 'drop')

# View the result
head(segment_counts)
```

Plot it
```{r plot-aut-clusters-basic, fig.height=5, fig.width=8}
# Define the path
input_path <-
  here("output",
       "ldna",
       "pop",
       "chr3",
       "clusters",
       "sushi2_AUT_chr3.txt")

# Read the data into R
clusters_data <- read.table(input_path, header = TRUE, sep = "\t") |>
  arrange(Start)

# Arrange by position
clusters_data <- clusters_data |>
  arrange(Start)

# Filter out rows with Size equal to 0
clusters_data <- subset(clusters_data, Size > 100000) # show only bigger than 100kb


# Calculate the maximum size for scaling
max_size <- max(clusters_data$Size)

# Create Start_Mb and End_Mb within clusters_data
clusters_data$Start_Mb <- clusters_data$Start / 1e6
clusters_data$End_Mb <- clusters_data$End / 1e6

# Create the plot
ggplot(clusters_data, aes(xmin = Start_Mb, xmax = End_Mb, ymin = 0, ymax = Size / 1e6)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = label_number(unit = "M"), breaks = pretty_breaks(n = 10)) +
  scale_y_continuous(labels = label_number(unit = "M"), breaks = pretty_breaks(n = 10)) +
  labs(x = "Chromosome 3 (Mb)", y = "Cluster Size (Mb)", fill = "Cluster ID") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    panel.grid.minor.x = element_blank(),
    legend.position = "top",
    legend.box = "horizontal"
  ) +
  guides(fill = guide_legend(ncol = 7, title.position = "top", title.hjust = 0.5))


# Use ggsave to save the plot as a PDF
ggsave(
  filename = here("output", "ldna", "pop", "chr3", "clusters", "AUT_chr3.pdf"),
  device = "pdf",
  width = 8,
  height = 5,
  units = "in"
)
```


We can annotate the number of segments in the legend.

```{r plot-aut-clusters-annotated, fig.height=5, fig.width=10}
# Merge SNP counts into clusters_data while specifying suffixes
annotated_data <- merge(clusters_data, segment_counts, by = "Cluster")

# Create a new column with Cluster names and SNP counts
annotated_data$Cluster_with_SNPs <- with(annotated_data, paste(Cluster, " (", nSegments, ")", sep=""))

# Now the plot uses the new Cluster_with_SNPs for the fill legend
ggplot(annotated_data, aes(xmin = Start_Mb, xmax = End_Mb, ymin = 0, ymax = Size / 1e6)) +
  geom_rect(aes(fill = as.factor(Cluster_with_SNPs)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = scales::label_number(unit = "M"), breaks = scales::pretty_breaks(n = 10)) +
  scale_y_continuous(labels = scales::label_number(unit = "M"), breaks = scales::pretty_breaks(n = 10)) +
  labs(x = "Chromosome 3 (Mb)", y = "Cluster Size (Mb)", fill = "Cluster ID") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    panel.grid.minor.x = element_blank(),
    legend.position = "top",
    legend.box = "horizontal"
  ) +
  guides(fill = guide_legend(nrow = 10, title.position = "top", title.hjust = 0.5))

# Use ggsave to save the plot as a PDF
ggsave(
  filename = here("output", "ldna", "pop", "chr3", "clusters", "AUT_chr3b.pdf"),
  device = "pdf",
  width = 10,
  height = 5,
  units = "in"
)
```

Save plotting data, we can arrange the column order first
```{r save-aut-plot-data}
plot_data <- annotated_data |>
  dplyr::select(
    Cluster, nSegments, Start, End, Size, Start_Mb, End_Mb
  )

# Create a new column with Chromosome
plot_data <- data.frame(Chromosome = rep(1, nrow(plot_data)), plot_data)

# Split the 'Cluster' column at "_", and create the 'r2' column with the second part
plot_data$r2 <- sapply(strsplit(as.character(plot_data$Cluster), "_"), `[`, 2)

# Move the 'r2' column to be the second column, right after 'Chromosome'
plot_data <- plot_data[, c(1:2, ncol(plot_data), 3:(ncol(plot_data)-1))]


# Arrange
plot_data <- plot_data |>
  dplyr::arrange(Start)

# Check it
head(plot_data)

# Save it
saveRDS(plot_data, file = here("output", "ldna", "pop", "chr3", "AUT_plot.rds"))
```

We can also import the summary file to create a object with SNP counts per segment
```{r save-aut-cluster-seg-snp}
# Define the path
input_path <-
  here("output",
       "ldna",
       "pop",
       "chr3",
       "summary_AUT.txt")

# Read the data into R
clusters_snps <- read.table(input_path, header = TRUE, sep = "\t")

# We subset and rename the column 1
clusters_snps <- clusters_snps |>
  dplyr::select(
    Name, nLoci
  ) |>
  dplyr::rename(
    Cluster = Name,
    nSNPs = nLoci
  )

# Merge
cluster_seg_snp <- inner_join(segment_counts, clusters_snps, by = "Cluster")
cluster_seg_snp <- inner_join(clusters_data, cluster_seg_snp, by = "Cluster")

# Split the 'Cluster' column at "_", and create the 'r2' column with the second part
cluster_seg_snp$r2 <- sapply(strsplit(as.character(cluster_seg_snp$Cluster), "_"), `[`, 2)

# Add chromosome number and population name
cluster_seg_snp <- data.frame(Chromosome = rep(3, nrow(cluster_seg_snp)), cluster_seg_snp)
cluster_seg_snp <- data.frame(Population = rep("AUT", nrow(cluster_seg_snp)), cluster_seg_snp)


# Reorder the columns
cluster_seg_snp <- cluster_seg_snp |>
  dplyr::select(
    Chromosome, Cluster, r2, Start, End, nSegments, nSNPs
  ) |>
  dplyr::arrange(
    Chromosome, Start
  )

head(cluster_seg_snp)


# Save the data for plotting later
saveRDS(cluster_seg_snp, file = here("output", "ldna", "pop", "chr3", "AUT_plot2.rds"))
```


