---
title: "Autogenous - LDna combined analysis"
author: "Luciano V Cosme"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: breezedark
    css:
      - "styles.css"
    toc: yes
    toc_float: no
    toc_depth: 5
editor_options:
  markdown:
    wrap: 120
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval                        = TRUE,
  echo                        = TRUE,
  cache                       = FALSE,
  class.output                = "bg-success"
)
knitr::opts_knit$set(
  root.dir = here::here()
)
```

> **Population label key:** The code labels used in the analysis below correspond to the following manuscript labels: AUT = AUTO, MAN = NON-AUTO, NEW = NON-AUTO-FIELD.

```{r libraries, message=FALSE, warning=FALSE, results='hide'}
# Load tidyverse FIRST to avoid rlang version conflict
library(tidyverse)
library(here)
library(scales)
library(ggrepel)

# Then add LDna v.2.15 library path and load
.libPaths(c("/workspace/lib/ldna_v215", .libPaths()))
library(LDna)
library(reshape2)
library(writexl)
```


## 6. Combine the chromosomal data

Clean env and memory
```{r clean-env-memory}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```

Read the plotting data for NEW and AUT
```{r load-plot-data-new-aut}
# NEW
NEW1 <- readRDS(here("output", "ldna", "pop", "chr1", "NEW_plot2.rds")) |> mutate(Population = "NEW")
NEW2 <- readRDS(here("output", "ldna", "pop", "chr2", "NEW_plot2.rds")) |> mutate(Population = "NEW")
NEW3 <- readRDS(here("output", "ldna", "pop", "chr3", "NEW_plot2.rds")) |> mutate(Population = "NEW")

# AUT
AUT1 <- readRDS(here("output", "ldna", "pop", "chr1", "AUT_plot2.rds")) |> mutate(Population = "AUT")
AUT2 <- readRDS(here("output", "ldna", "pop", "chr2", "AUT_plot2.rds")) |> mutate(Population = "AUT")
AUT3 <- readRDS(here("output", "ldna", "pop", "chr3", "AUT_plot2.rds")) |> mutate(Population = "AUT")
```

Bind the objects

```{r bind-all-populations}
albo <- rbind(NEW1, NEW2, NEW3, AUT1, AUT2, AUT3)

head(albo)
```

We can create a table
```{r create-cluster-table}
table1 <- albo |>
  mutate(Size = End - Start) |>
  dplyr::select(
    Population, Chromosome, Cluster, r2, nSegments, nSNPs, Start, End, Size
  ) |>
  arrange(Population, Chromosome, Start)
head(table1)
```

We can focus on LD blocks equal or bigger than 1Mb
```{r filter-clusters-1mb}
table1 <- table1 %>%
  dplyr::filter(Size >= 1000000) %>%
  dplyr::mutate(`Size (Mb)` = round(Size / 1000000, 2)) # %>%
  # dplyr::select(Size)
head(table1)
```

Create table
```{r kable-cluster-table}
knitr::kable(table1)
```

We can create a facet plot now for all chromosomes

```{r plot-clusters-fixed-scale, fig.height=5, fig.width=10}
# To plot only clusters equal or bigger than 1Mb
albo2 <- table1 |> dplyr::filter(Size >= 1000000)

# Function to format numbers as Mb
label_mb <- function(x) {
  sprintf("%.0fMb", x / 1e6)
}

# Plot it
ggplot(albo2, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    strip.background = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position = "none",
    panel.spacing.x = unit(1, "lines")  # Adjust the unit and number to increase space as needed
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed") +
  guides(fill = "none")

# Use ggsave to save the plot as a PDF
# ggsave(
#   filename = here("output", "ldna", "AUT_NEW_fixed.pdf"),
#   device = "pdf",
#   width = 10,
#   height = 5,
#   units = "in"
# )
```

Make each plot with different scales
```{r plot-clusters-free-scale, fig.height=5, fig.width=8}
# Plot it
ggplot(albo2, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    panel.grid.minor.x = element_blank(),
    # strip.background = element_rect(fill = "#e8e8e8", colour = NA),
    legend.position = "none",
    panel.spacing.x = unit(1, "lines")
  ) +
  facet_wrap(~ Population + Chromosome, scales = "free_y", ncol = 3) +  # Free y scale, and each chromosome gets its own row
  guides(fill = "none")

# Use ggsave to save the plot as a PDF
# ggsave(
#   filename = here("output", "ldna", "AUT_NEW_free.pdf"),
#   device = "pdf",
#   width = 8,
#   height = 5,
#   units = "in"
# )
```

We can also remove the large clusters and plot with fixed scale to look at the cluster between 1 and 10Mb
```{r plot-clusters-1-to-10mb, fig.height=5, fig.width=8}
# To plot only clusters equal or bigger than 1Mb but less that 10Mb
albo3 <- albo2 |> dplyr::filter(Size >= 1000000, Size <= 10000000)


# Plot it
ggplot(albo3, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    strip.background = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position = "none",
    panel.spacing.x = unit(1, "lines")  # Adjust the unit and number to increase space as needed
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed") +
  guides(fill = "none")

# Use ggsave to save the plot as a PDF
# ggsave(
#   filename = here("output", "ldna", "AUT_NEW_fixed_1_to_10Mb.pdf"),
#   device = "pdf",
#   width = 8,
#   height = 5,
#   units = "in"
# )
```

It seems that the linkage patters are quite different from AUT to the wild population. I am not sure why we are seeing clusters at such long distance the wild population for chromosome 2 and 3

We could do the analysis by sex, but we only have 13 females and 15 males in AUT. Perhaps we could use the same number of samples for each. However, the sample size will be too small still.

## 7. Plot SNPs on DE genes with LD blocks
```{r load-snps-scan-de}
snps_scan <-
  read.table(
    # here("output", "pcadapt", "man_aut_common_SNPs_pcadapt_outflank.txt"),
    here("output", "snpeff", "SNPs_79_DE.txt"),
    stringsAsFactors = FALSE
  ) |>
  dplyr::rename(
    SNP = V1
  )

head(snps_scan)
```

Import bim file
```{r import-bim-ldna-file1}
# Import the function
source(
  here(
    "notebooks", "helpers", "import_bim.R")
)

# Import the data
snps <- import_bim(here("output", "ldna", "files", "file1.bim"))

# Check it
head(snps)
```

Merge the objects
```{r merge-scan-snps-bim}
merged_snps <- inner_join(snps_scan, snps, by = "SNP") |>
  dplyr::rename(
    Chromosome = Scaffold
  )
head(merged_snps)
```

We can add them as lines in the plot
```{r plot-snps-on-ld-blocks, fig.height=5, fig.width=8}
# First, create a new data frame with 1Mb windows and count SNPs in each window
merged_snps$Window <- floor(merged_snps$Position / 20e6) * 20e6 # 20mb
snp_counts <- merged_snps %>%
  group_by(Chromosome, Window) %>%
  summarize(SNP_count = n(), .groups = 'drop')

# Now, add this SNP information to the plot
snp_plot <- ggplot() +
  geom_rect(data = albo2, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size, fill = as.factor(Cluster)),
            color = "black", linewidth = 0.2) +
  geom_vline(data = snp_counts, aes(xintercept = Window + 0.5e6), color = "pink", linetype = "dotted", linewidth = 0.5) +
  geom_text(data = snp_counts, aes(x = Window + 0.5e6, y = 8e6, label = SNP_count), vjust = -0.5, color = "blue", size = 2) +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position (Mb)", y = "Cluster Size (Mb)") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotdash"),
    strip.background = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position = "none",
    panel.spacing.x = unit(1, "lines")
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed")

# Display the plot
print(snp_plot)

#
# # Use ggsave to save the plot as a PDF
# ggsave(
#   filename = here("output", "ldna", "linkage_snps_genes_de.pdf"),
#   device = "pdf",
#   width = 8,
#   height = 5,
#   units = "in"
# )
```

Find the non-overlapping clusters
```{r find-non-overlapping-clusters-v1}
# Split the data into two subsets based on Population
albo_NEW <- subset(albo2, Population == "NEW")
albo_AUT <- subset(albo2, Population == "AUT")

# Function to find non-overlapping clusters
find_non_overlapping_clusters <- function(new_clusters, aut_clusters) {
  non_overlapping <- vector("list", length(new_clusters$Chromosome))
  names(non_overlapping) <- unique(new_clusters$Chromosome)

  for (chr in unique(new_clusters$Chromosome)) {
    new_chrom_clusters <- new_clusters[new_clusters$Chromosome == chr, ]
    aut_chrom_clusters <- aut_clusters[aut_clusters$Chromosome == chr, ]

    non_overlap <- vector("list", nrow(new_chrom_clusters))

    for (i in 1:nrow(new_chrom_clusters)) {
      overlaps <- any(aut_chrom_clusters$Start <= new_chrom_clusters$End[i] & aut_chrom_clusters$End >= new_chrom_clusters$Start[i])
      if (!overlaps) {
        non_overlap[[i]] <- new_chrom_clusters[i, ]
      }
    }

    non_overlapping[[chr]] <- do.call(rbind, non_overlap)
  }

  do.call(rbind, non_overlapping)
}

# Find non-overlapping clusters
non_overlapping_clusters <- find_non_overlapping_clusters(albo_NEW, albo_AUT)

# View the non-overlapping clusters
non_overlapping_clusters
```
Sanity check: Second way to get it
```{r find-non-overlapping-clusters-v2}
# Function to find non-overlapping segments
find_non_overlapping_segments <- function(new_clusters, aut_clusters) {
  non_overlapping <- list()

  for (chr in unique(new_clusters$Chromosome)) {
    new_chrom_segments <- new_clusters[new_clusters$Chromosome == chr, ]
    aut_chrom_segments <- aut_clusters[aut_clusters$Chromosome == chr, ]

    non_overlap <- list()

    for (i in 1:nrow(new_chrom_segments)) {
      overlaps <- FALSE
      for (j in 1:nrow(aut_chrom_segments)) {
        if (new_chrom_segments$Start[i] <= aut_chrom_segments$End[j] && new_chrom_segments$End[i] >= aut_chrom_segments$Start[j]) {
          overlaps <- TRUE
          break
        }
      }

      if (!overlaps) {
        non_overlap <- c(non_overlap, list(new_chrom_segments[i, ]))
      }
    }

    non_overlapping[[chr]] <- do.call(rbind, non_overlap)
  }

  do.call(rbind, non_overlapping)
}

# Find non-overlapping segments
non_overlapping_segments <- find_non_overlapping_segments(albo_NEW, albo_AUT)

# View the non-overlapping segments
non_overlapping_segments
```
```{r find-unique-clusters-per-population}
# Function to find unique non-overlapping clusters for each population
find_unique_clusters <- function(population1, population2) {
  unique_clusters <- list()

  for (chr in unique(population1$Chromosome)) {
    pop1_chrom_clusters <- population1[population1$Chromosome == chr, ]
    pop2_chrom_clusters <- population2[population2$Chromosome == chr, ]

    unique_clusters_chr <- list()

    for (i in 1:nrow(pop1_chrom_clusters)) {
      overlaps <- any(
        pop2_chrom_clusters$Start <= pop1_chrom_clusters$End[i] &
        pop2_chrom_clusters$End >= pop1_chrom_clusters$Start[i]
      )

      if (!overlaps) {
        unique_clusters_chr <- c(unique_clusters_chr, list(pop1_chrom_clusters[i, ]))
      }
    }

    if (length(unique_clusters_chr) > 0) {
      unique_clusters[[chr]] <- do.call(rbind, unique_clusters_chr)
    }
  }

  do.call(rbind, unique_clusters)
}

# Find unique non-overlapping clusters for NEW and AUT populations
unique_clusters_NEW <- find_unique_clusters(albo2[albo2$Population == "NEW",], albo2[albo2$Population == "AUT",])
unique_clusters_AUT <- find_unique_clusters(albo2[albo2$Population == "AUT",], albo2[albo2$Population == "NEW",])

# View the unique non-overlapping clusters for each population
unique_clusters_NEW
unique_clusters_AUT
```


Plot it
```{r plot-unique-clusters-combined, fig.height=5, fig.width=10}
# Combine the two data frames
combined_clusters <- rbind(unique_clusters_NEW, unique_clusters_AUT)


# Function to format numbers as Mb
label_mb <- function(x) {
  sprintf("%.0fMb", x / 1e6)
}

# Plot it
ggplot(combined_clusters, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    strip.background = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position = "none",
    panel.spacing.x = unit(1, "lines")  # Adjust the unit and number to increase space as needed
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed") +
  guides(fill = "none")

# Use ggsave to save the plot as a PDF
# ggsave(
#   filename = here("output", "ldna", "AUT_NEW_fixed_non_overlapping.pdf"),
#   device = "pdf",
#   width = 10,
#   height = 5,
#   units = "in"
# )
```

```{r load-snps-on-de-genes}
snps_DEgenes<- read_delim(here("output", "snpeff", "SNPs_on_DE_genes.txt"), delim = "\t", col_names = TRUE, show_col_types = FALSE)
head(snps_DEgenes)
```

Check if any SNP fall within the clusters

```{r intersect-de-snps-clusters}
# Initialize an empty data frame to store the results
results <- data.frame(Population = character(), Chromosome = integer(),
                      Cluster = character(), r2 = numeric(),
                      nSegments = integer(), nSNPs = integer(),
                      Start = integer(), End = integer(), Size = integer(),
                      SizeMb = numeric(), SNP = character(), Position_chr = integer(),
                      stringsAsFactors = FALSE)

for (i in 1:nrow(snps_DEgenes)) {
    for (j in 1:nrow(combined_clusters)) {
        if (snps_DEgenes$Chromosome[i] == combined_clusters$Chromosome[j] &&
            snps_DEgenes$Position_chr[i] >= combined_clusters$Start[j] &&
            snps_DEgenes$Position_chr[i] <= combined_clusters$End[j]) {

            # Create a new row as a data frame with the same column names
            new_row <- data.frame(Population = combined_clusters$Population[j],
                                  Chromosome = combined_clusters$Chromosome[j],
                                  Cluster = combined_clusters$Cluster[j],
                                  r2 = combined_clusters$r2[j],
                                  nSegments = combined_clusters$nSegments[j],
                                  nSNPs = combined_clusters$nSNPs[j],
                                  Start = combined_clusters$Start[j],
                                  End = combined_clusters$End[j],
                                  Size = combined_clusters$Size[j],
                                  SizeMb = combined_clusters$Size[j],
                                  SNP = snps_DEgenes$SNP[i],
                                  Position_chr = snps_DEgenes$Position_chr[i],
                                  stringsAsFactors = FALSE)

            # Append the new row to the results data frame
            results <- rbind(results, new_row)
        }
    }
}


head(results)
```

We can count how many SNPs we have per cluster per population
```{r count-snps-per-cluster}
snp_counts <- results %>%
  select(Population, Chromosome, Cluster, nSNPs) %>%
  distinct() %>%
  rename(SNP_Count = nSNPs)


head(snp_counts)
```

Plot the clusters for which we have SNPs on the DE genes
```{r plot-clusters-with-de-snps, fig.height=5, fig.width=8}
# Merge SNP counts into results
results_with_counts <- merge(results, snp_counts, by = c("Population", "Chromosome", "Cluster"))

ggplot(results_with_counts, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  geom_text(aes(x = (Start + End) / 2, y = Size + 0.5, label = SNP_Count), size = 3, vjust = -0.2) + # Annotate SNP counts on top
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    strip.background = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position = "none",
    panel.spacing.x = unit(1, "lines")
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed") +
  guides(fill = "none")


```
```{r assign-cluster-integer-ids}
# 1. Generate integer IDs for each cluster in plotting order
cluster_ids <- data.frame(
  Cluster     = sort(unique(results$Cluster)),
  Cluster_ID  = seq_along(unique(results$Cluster))
)

# 2. Merge IDs into your results
results_with_ids <- merge(results, cluster_ids, by = "Cluster")

# 3. Plot, labeling by your original 1,2,3,… IDs
ggplot(results_with_ids, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = factor(Cluster_ID)), color = "black", linewidth = 0.2) +
  geom_text(aes(x = (Start + End) / 2,
                y = Size + 0.5,
                label = Cluster_ID),
            size = 3, vjust = -0.2) +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y        = element_text(margin = margin(t = 0, r = 5)),
    axis.ticks.y       = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    strip.background   = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position    = "none",
    panel.spacing.x    = unit(1, "lines")
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed") +
  guides(fill = "none")

```

The table below maps the integer cluster IDs used in the plots to the `extractBranches` cluster names, the chromosome, the population, and the number of SNPs. Clusters referenced in the manuscript as **cluster 14** and **cluster 6** are highlighted.

```{r build-cluster-id-mapping-table}
# Cluster ID mapping table
cluster_ids <- data.frame(
  Cluster     = sort(unique(results_with_counts$Cluster)),
  Cluster_ID  = seq_along(unique(results_with_counts$Cluster))
)

# Merge with population/chromosome/SNP count info
cluster_map <- merge(cluster_ids,
  results_with_counts %>%
    select(Cluster, Population, Chromosome, SNP_Count) %>%
    distinct(),
  by = "Cluster")

# Add manuscript name column
cluster_map$Manuscript_Name <- ifelse(cluster_map$Cluster == "3215_0.78", "Cluster 14",
                              ifelse(cluster_map$Cluster == "1971_0.64", "Cluster 6", ""))

knitr::kable(cluster_map %>% arrange(Cluster_ID),
  caption = "Cluster ID mapping: plot labels to extractBranches names. Cluster 14 (chr2 AUT, 3215_0.78) and Cluster 6 (chr1 AUT, 1971_0.64) are the two major LD clusters discussed in the manuscript.")
```

```{r plot-clusters-with-id-labels, fig.height=5, fig.width=8}
# 1. Merge integer IDs into your SNP‐counted data
results_with_counts <- merge(results_with_counts, cluster_ids, by = "Cluster")

# 2. Build the fill label
results_with_counts$Cluster_SNP <- with(
  results_with_counts,
  paste0(Cluster, " (", SNP_Count, " SNPs)")
)

# 3. Plot with fill by SNP count label and text by cluster ID
ggplot(results_with_counts, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = Cluster_SNP), color = "black", linewidth = 0.2) +
  geom_text(
    aes(x = (Start + End) / 2, y = Size + 0.5, label = Cluster_ID),
    size = 3, vjust = -0.2
  ) +
  scale_fill_discrete(name = "Cluster and SNP Count") +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y        = element_text(margin = margin(t = 0, r = 5)),
    axis.ticks.y       = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    strip.background   = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position    = "right",
    panel.spacing.x    = unit(1, "lines")
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed")
```



Save
```{r save-results-with-counts, eval=TRUE}
saveRDS(results_with_counts, file = here("output", "ldna", "results_with_counts.rds"))
```

Read the data
```{r reload-results-with-counts}
results_with_counts <- readRDS(file = here("output", "ldna", "results_with_counts.rds"))
```


```{r plot-clusters-sorted-legend, fig.height=5, fig.width=12}
# Rebuild the fill label: add ID before Cluster and drop “ SNPs”
results_with_counts$Cluster_SNP <- with(
  results_with_counts,
  paste0(Cluster_ID, ": ", Cluster, " (", SNP_Count, ")")
)

# 1. Reorder Cluster_SNP factor by Cluster_ID so the legend follows numeric order
results_with_counts$Cluster_SNP <- factor(
  results_with_counts$Cluster_SNP,
  levels = results_with_counts %>%
    distinct(Cluster_ID, Cluster_SNP) %>%
    arrange(Cluster_ID) %>%
    pull(Cluster_SNP)
)

# 2. Plot with sorted legend
ggplot(results_with_counts, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = Cluster_SNP), color = "black", linewidth = 0.2) +
  geom_text(
    aes(x = (Start + End) / 2, y = Size + 0.5, label = Cluster_ID),
    size = 3, vjust = -0.2
  ) +
  scale_fill_discrete(name = "Cluster and SNP Count") +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y        = element_text(margin = margin(t = 0, r = 5)),
    axis.ticks.y       = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    strip.background   = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position    = "right",
    panel.spacing.x    = unit(1, "lines")
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed")



# Use ggsave to save the plot as a PDF
ggsave(
  filename = here("output", "ldna", "AUT_NEW_non_overlapping_clusters_with_SNPs_on_DE_genes.pdf"),
  device = "pdf",
  width = 10,
  height = 5,
  units = "in"
)
```


Save it
```{r save-results-xlsx}
library(writexl)

# 1. Drop Cluster_ID column(s)
cleaned_results <- results_with_counts %>%
  select(-starts_with("Cluster_ID")) %>%
  mutate(
    Start        = formatC(Start,        format = "f", digits = 0),
    End          = formatC(End,          format = "f", digits = 0),
    Position_chr = formatC(Position_chr, format = "f", digits = 0)
  )


# 2. Export cleaned results_with_counts to Excel
write_xlsx(
  list(Results = cleaned_results),
  here("output/ldna/results_with_counts.xlsx")
)

```

<!-- Development marker removed -->

Now we can get the SNPs on the DE genes
```{r load-snps-79-de-genes}
snps79_DE_genes<- read_delim(here("output", "snpeff","SNPs_79_DE.txt"), delim = "\t", col_names = FALSE, show_col_types = FALSE)
head(snps79_DE_genes)
```

Or the SNPs from the selection scan
```{r load-snps-157-selection}
snps_selection <- read_delim(here("output", "pcadapt","outlier_157_SNPs.txt"), delim = "\t", col_names = FALSE, show_col_types = FALSE)
head(snps_selection)
```



Filter
```{r filter-results-by-de-snps}
# Create a logical vector to filter rows based on SNP IDs
filter_vector <- results_with_counts$SNP %in% snps79_DE_genes$X1
# filter_vector <- results_with_counts$SNP %in% snps_selection$X1  # we do not have any

# Subset results_with_counts to keep only rows where SNP matches
filtered_results <- results_with_counts[filter_vector, ]

head(filtered_results)
# length(filtered_results)
```

We have 16 SNPs
```{r plot-filtered-results-16-snps, fig.height=5, fig.width=8}
# Assign unique numbers to each cluster and create labels
filtered_results <- filtered_results %>%
  mutate(Cluster_Number = as.numeric(as.factor(Cluster))) %>%
  arrange(Cluster_Number) %>%
  mutate(Cluster_Label = paste(Cluster_Number, "_", Cluster, " (", SNP_Count, ")", sep=""))

# Reorder the factor levels of Cluster_Label based on Cluster_Number
filtered_results$Cluster_Label <- factor(filtered_results$Cluster_Label,
                                            levels = unique(filtered_results$Cluster_Label[order(filtered_results$Cluster_Number)]))

# Plotting
ggplot(filtered_results, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = Cluster_Label), color = "black", linewidth = 0.2) +
  geom_text(aes(x = (Start + End) / 2, y = Size + 0.5, label = Cluster_Number), size = 3, vjust = -0.2) +
  scale_fill_discrete(name = "Cluster ID") +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    strip.background = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position = "right",
    panel.spacing.x = unit(1, "lines")
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed")


# Use ggsave to save the plot as a PDF
ggsave(
  filename = here("output", "ldna", "AUT_NEW_overlapping_clusters_with_SNPs_on_DE_genes_intersect.pdf"),
  device = "pdf",
  width = 10,
  height = 5,
  units = "in"
)
```
This are the clusters for which we found the SNP as outlier and the gene the SNP in on is DE



Get the overlapping clusters
```{r find-overlapping-clusters}
# Function to find overlapping clusters
find_overlapping_clusters <- function(population1, population2) {
  overlapping_clusters <- list()

  for (chr in unique(population1$Chromosome)) {
    pop1_chrom_clusters <- population1[population1$Chromosome == chr, ]
    pop2_chrom_clusters <- population2[population2$Chromosome == chr, ]

    overlapping_clusters_chr <- list()

    for (i in 1:nrow(pop1_chrom_clusters)) {
      for (j in 1:nrow(pop2_chrom_clusters)) {
        if (pop1_chrom_clusters$Start[i] <= pop2_chrom_clusters$End[j] &&
            pop1_chrom_clusters$End[i] >= pop2_chrom_clusters$Start[j]) {
          overlapping_clusters_chr <- c(overlapping_clusters_chr, list(pop1_chrom_clusters[i, ]))
          break
        }
      }
    }

    if (length(overlapping_clusters_chr) > 0) {
      overlapping_clusters[[chr]] <- do.call(rbind, overlapping_clusters_chr)
    }
  }

  do.call(rbind, overlapping_clusters)
}

# Subset data for NEW and AUT populations
albo_NEW <- subset(albo2, Population == "NEW")
albo_AUT <- subset(albo2, Population == "AUT")

# Find overlapping clusters between NEW and AUT
overlapping_clusters <- find_overlapping_clusters(albo_NEW, albo_AUT)

# View the overlapping clusters
overlapping_clusters
```

Plot it
```{r plot-overlapping-clusters, fig.height=5, fig.width=10}
# Function to format numbers as Mb
label_mb <- function(x) {
  sprintf("%.0fMb", x / 1e6)
}

# Plot it
ggplot(overlapping_clusters, aes(xmin = Start, xmax = End, ymin = 0, ymax = Size)) +
  geom_rect(aes(fill = as.factor(Cluster)), color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = label_mb, breaks = pretty_breaks(n = 3)) +
  scale_y_continuous(labels = label_mb, breaks = pretty_breaks(n = 5)) +
  labs(x = "Position", y = "Cluster Size") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(margin = margin(t = 0, r = 5, b = 0, l = 5)),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
    strip.background = element_rect(fill = "#e8e8e8", colour = NA),
    panel.grid.minor.x = element_blank(),
    legend.position = "none",
    panel.spacing.x = unit(1, "lines")  # Adjust the unit and number to increase space as needed
  ) +
  facet_grid(Population ~ Chromosome, scales = "fixed", space = "fixed") +
  guides(fill = "none")

# Use ggsave to save the plot as a PDF
ggsave(
  filename = here("output", "ldna", "AUT_NEW_fixed_overlapping.pdf"),
  device = "pdf",
  width = 10,
  height = 5,
  units = "in"
)
```

## 8. Mapping back SNP clusters to scaffolds

Clean env and memory
```{r clean-env-section8}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```

With chromosomal-scale coordinates established for the linkage network analysis, the next step is to map SNPs from clusters of interest back to scaffold coordinates, enabling cross-referencing with the gene annotation file to identify genes within each linkage group.

Cluster 14 is on chromosome 2 and only in the AUT line. The cluster ID from `extractBranches` (LDna v2.15) is `3215_0.78`. This corresponds to cluster `3139_0.78` from the original analysis with `extractClusters` (LDna v.63). The cluster name changed because v2.15 `LDnaRaw` produces a slightly different dendrogram structure, but the SNP content is the same — all 316 SNPs from the original cluster are contained within this cluster (plus 8 additional SNPs at the cluster boundary). The LD threshold (0.78) is identical.
```{r load-aut-chr2-clusters}
aut_ch2 <- readRDS(here("output", "ldna", "pop", "chr2", "AUT_clusters_snps.rds"))
str(aut_ch2$`3215_0.78`)
```

Now we can import the bim file with the scaffolds. We can use the backup file we created when we created the chromosomal scale.
```{bash}
head output/quality_control/file7_backup.bim
```

Import it
```{r import-bim-file7-backup}
# Import the function
source(
  here(
    "notebooks", "helpers", "import_bim.R")
)

# Import the data
snps <- import_bim(here("output", "quality_control", "file7_backup.bim"))

# Check it
head(snps)
```


Now we can filter the 566 SNPs that are in the cluster and see what scaffolds there are located

```{r extract-cluster-14-snps}
cluster_14 <- snps[snps$SNP %in% aut_ch2$`3215_0.78`, ]
head(cluster_14) # 316 SNPs

# Save it
saveRDS(
  cluster_14, here(
    "output", "ldna", "cluster_14.rds"
    )
  )
```

Create table
```{r kable-cluster-14}
knitr::kable(cluster_14)
```

Cluster 6 is on chromosome 1, AUT only. The cluster ID from `extractBranches` (LDna v2.15) is `1971_0.64`, corresponding to `1942_0.64` from the original `extractClusters` (LDna v.63) analysis. Same as cluster 14, all 88 original SNPs are contained within this cluster (plus 2 additional SNPs). The LD threshold (0.64) is identical.
```{r load-aut-chr1-clusters}
aut_ch1 <- readRDS(here("output", "ldna", "pop", "chr1", "AUT_clusters_snps.rds"))
str(aut_ch1$`1971_0.64`)
```

```{r extract-cluster-6-snps}
cluster_6 <- snps[snps$SNP %in% aut_ch1$`1971_0.64`, ]
head(cluster_6) # 316 SNPs

# Save it
saveRDS(
  cluster_6, here(
    "output", "ldna", "cluster_6.rds"
    )
  )
```


Create table
```{r kable-cluster-6}
knitr::kable(cluster_6)
```


```{r get-cluster-6-all-scaffold-snps}
cluster_6_all <- snps[snps$Scaffold %in% cluster_6$Scaffold, ]
head(cluster_6_all)
```


However we need the other SNPs to know if the blocks are connected or not. Now we know what scaffolds the SNPs are located and we can check it
```{r get-cluster-14-all-scaffold-snps}
cluster_14_all <- snps[snps$Scaffold %in% cluster_14$Scaffold, ]
head(cluster_14_all)
```

Now we can create a new column and tag the SNPs that are linked
```{r tag-linked-snps}
# Creating the new column 'linked'
cluster_14_all <- cluster_14_all %>%
  mutate(linked = SNP %in% aut_ch2$`3215_0.78`)
head(cluster_14_all)

# Creating the new column 'linked'
cluster_6_all <- cluster_6_all %>%
  mutate(linked = SNP %in% aut_ch1$`1971_0.64`)
head(cluster_6_all)
```

Lets check how many linked SNPs we have and how many aren't
```{r count-linked-snps-cluster-14}
count_linked <- cluster_14_all %>%
  group_by(linked) %>%
  summarise(count = n())
count_linked
```

```{r count-linked-snps-cluster-6}
count_linked <- cluster_6_all %>%
  group_by(linked) %>%
  summarise(count = n())
count_linked
```

Get the 157 outliers
```{r load-outlier-157-snps}
snps_157 <-
  read.table(
    here("output", "pcadapt", "outlier_157_SNPs.txt"),
    stringsAsFactors = FALSE
    )

# Get the 157 SNPs
snps_157b <- cluster_14_all |>
  filter(SNP %in% snps_157$V1)

head(snps_157b)
```


```{r filter-157-snps-cluster-6}
# Get the 157 SNPs
snps_157c <- cluster_6_all |>
  filter(SNP %in% snps_157$V1)

head(snps_157c)
```

Plot it
```{r plot-cluster-14-linked-scaffolds, fig.height=5, fig.width=8}
# Function to format numbers as Mb
label_mb <- function(x) {
  sprintf("%.0fMb", x / 1e6)
}


# Calculate the start and end positions for each stretch of TRUE or FALSE
rect_data <- cluster_14_all %>%
  arrange(Scaffold, Position) %>%
  mutate(change = linked != lag(linked, default = first(linked))) %>%
  group_by(Scaffold) %>%
  mutate(group_id = cumsum(change)) %>%
  group_by(Scaffold, group_id, linked) %>%
  summarize(start = min(Position), end = max(Position), .groups = 'drop') %>%
  ungroup()

# Plotting using geom_rect
ggplot(rect_data, aes(xmin = start, xmax = end, ymin = as.numeric(Scaffold) - 0.1, ymax = as.numeric(Scaffold) + 0.1)) +
  # Add a background layer for each scaffold with gray color
  geom_rect(data = rect_data %>% group_by(Scaffold) %>%
            summarize(start = min(start), end = max(end), .groups = 'drop'),
            aes(xmin = start, xmax = end, ymin = as.numeric(Scaffold) - 0.1, ymax = as.numeric(Scaffold) + 0.1),
            fill = "gray", inherit.aes = FALSE) +
  # Add the TRUE stretches
  geom_rect(aes(fill = linked), data = rect_data %>% filter(linked == TRUE)) +
  scale_fill_manual(values = c("TRUE" = "green")) +
  facet_wrap(~ Scaffold, scales = "fixed", ncol = 1) +
  geom_vline(data = snps_157b, aes(xintercept = Position, color = SNP), linetype = "solid", color = "red") +
  geom_text(data = snps_157b, aes(x = Position, y = as.numeric(Scaffold) + 0.2, label = SNP), inherit.aes = FALSE, angle = 90, vjust = 0, size = 2, check_overlap = TRUE) +
  theme_minimal() +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.spacing = unit(0.1, "lines"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 12, face = "bold", hjust = 0.5, margin = margin(t = 1, b = 5))
  ) +
  scale_x_continuous(labels = label_mb, breaks = scales::pretty_breaks(n = 6)) +
  labs(x = "Position", title = "Continuous Stretches of Linked SNPs by Scaffold - Cluster 14 AUT", fill = "Linked") +
  guides(fill = "none")

# Save Venn diagram to PDF
output_path <- here("output", "ldna", "figures", "cluster_14_aut_scaffolds.pdf")
ggsave(output_path, height = 5, width = 8, units = "in")
```



To make sure the windows appear even if they are small
```{r plot-cluster-14-min-window, fig.height=5, fig.width=8}
# Set a minimum width for the windows
min_width <- 1e4  # for example, 1 million base pairs

# Adjust the rect_data calculation
rect_data <- cluster_14_all %>%
  arrange(Scaffold, Position) %>%
  mutate(change = linked != lag(linked, default = first(linked))) %>%
  group_by(Scaffold) %>%
  mutate(group_id = cumsum(change)) %>%
  group_by(Scaffold, group_id, linked) %>%
  summarize(start = min(Position), end = max(Position), .groups = 'drop') %>%
  ungroup() %>%
  mutate(width = end - start,
         adjusted_end = ifelse(width < min_width, start + min_width, end))

# Plotting using geom_rect with adjusted ends
ggplot(rect_data, aes(xmin = start, xmax = adjusted_end, ymin = as.numeric(Scaffold) - 0.1, ymax = as.numeric(Scaffold) + 0.1)) +
  geom_rect(data = rect_data %>% group_by(Scaffold) %>%
            summarize(start = min(start), end = max(adjusted_end), .groups = 'drop'),
            aes(xmin = start, xmax = end, ymin = as.numeric(Scaffold) - 0.1, ymax = as.numeric(Scaffold) + 0.1),
            fill = "gray", inherit.aes = FALSE) +
  geom_rect(aes(fill = linked), data = rect_data %>% filter(linked == TRUE)) +
  scale_fill_manual(values = c("TRUE" = "green")) +
  facet_wrap(~ Scaffold, scales = "fixed", ncol = 1) +
  theme_minimal() +
  geom_vline(data = snps_157b, aes(xintercept = Position, color = SNP), linetype = "solid", color = "red") +
  geom_text_repel(data = snps_157b, aes(x = Position, y = as.numeric(Scaffold) + 0.2, label = SNP),
                  inherit.aes = FALSE, angle = 45, size = 2,
                  nudge_y = 0.1,   # Adjust this value as needed
                  direction = "y") +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(), panel.spacing = unit(0.1, "lines"),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.text = element_text(size = 12, face = "bold", hjust = 0.5)) +
  scale_x_continuous(labels = label_mb, breaks = scales::pretty_breaks(n = 6)) +
  labs(x = "Position", title = "Continuous Stretches of Linked SNPs by Scaffold - Cluster 14 AUT", fill = "Linked") +
  guides(fill = "none")

# Save Venn diagram to PDF
output_path <- here("output", "ldna", "figures", "cluster_14_aut_scaffolds_with_small_windows.pdf")
ggsave(output_path, height = 5, width = 8, units = "in")
```

Cluster 6
```{r plot-cluster-6-linked-scaffolds, fig.height=5, fig.width=8}
# Function to format numbers as Mb
label_mb <- function(x) {
  sprintf("%.0fMb", x / 1e6)
}


# Calculate the start and end positions for each stretch of TRUE or FALSE
rect_data <- cluster_6_all %>%
  arrange(Scaffold, Position) %>%
  mutate(change = linked != lag(linked, default = first(linked))) %>%
  group_by(Scaffold) %>%
  mutate(group_id = cumsum(change)) %>%
  group_by(Scaffold, group_id, linked) %>%
  summarize(start = min(Position), end = max(Position), .groups = 'drop') %>%
  ungroup()

# Plotting using geom_rect
ggplot(rect_data, aes(xmin = start, xmax = end, ymin = as.numeric(Scaffold) - 0.1, ymax = as.numeric(Scaffold) + 0.1)) +
  # Add a background layer for each scaffold with gray color
  geom_rect(data = rect_data %>% group_by(Scaffold) %>%
            summarize(start = min(start), end = max(end), .groups = 'drop'),
            aes(xmin = start, xmax = end, ymin = as.numeric(Scaffold) - 0.1, ymax = as.numeric(Scaffold) + 0.1),
            fill = "gray", inherit.aes = FALSE) +
  # Add the TRUE stretches
  geom_rect(aes(fill = linked), data = rect_data %>% filter(linked == TRUE)) +
  scale_fill_manual(values = c("TRUE" = "green")) +
  facet_wrap(~ Scaffold, scales = "fixed", ncol = 1) +
  geom_vline(data = snps_157c, aes(xintercept = Position, color = SNP), linetype = "solid", color = "red") +
  geom_text(data = snps_157c, aes(x = Position, y = as.numeric(Scaffold) + 0.2, label = SNP), inherit.aes = FALSE, angle = 90, vjust = 0, size = 2, check_overlap = TRUE) +
  theme_minimal() +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.spacing = unit(0.1, "lines"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 12, face = "bold", hjust = 0.5, margin = margin(t = 1, b = 5))
  ) +
  scale_x_continuous(labels = label_mb, breaks = scales::pretty_breaks(n = 6)) +
  labs(x = "Position", title = "Continuous Stretches of Linked SNPs by Scaffold - Cluster 6 AUT", fill = "Linked") +
  guides(fill = "none")

# Save Venn diagram to PDF
output_path <- here("output", "ldna", "figures", "cluster_6_aut_scaffolds.pdf")
ggsave(output_path, height = 5, width = 8, units = "in")
```



To make sure the windows appear even if they are small
```{r plot-cluster-6-min-window, fig.height=5, fig.width=8}
# Set a minimum width for the windows
min_width <- 1e4  # for example, 1 million base pairs

# Adjust the rect_data calculation
rect_data <- cluster_6_all %>%
  arrange(Scaffold, Position) %>%
  mutate(change = linked != lag(linked, default = first(linked))) %>%
  group_by(Scaffold) %>%
  mutate(group_id = cumsum(change)) %>%
  group_by(Scaffold, group_id, linked) %>%
  summarize(start = min(Position), end = max(Position), .groups = 'drop') %>%
  ungroup() %>%
  mutate(width = end - start,
         adjusted_end = ifelse(width < min_width, start + min_width, end))

# Plotting using geom_rect with adjusted ends
ggplot(rect_data, aes(xmin = start, xmax = adjusted_end, ymin = as.numeric(Scaffold) - 0.1, ymax = as.numeric(Scaffold) + 0.1)) +
  geom_rect(data = rect_data %>% group_by(Scaffold) %>%
            summarize(start = min(start), end = max(adjusted_end), .groups = 'drop'),
            aes(xmin = start, xmax = end, ymin = as.numeric(Scaffold) - 0.1, ymax = as.numeric(Scaffold) + 0.1),
            fill = "gray", inherit.aes = FALSE) +
  geom_rect(aes(fill = linked), data = rect_data %>% filter(linked == TRUE)) +
  scale_fill_manual(values = c("TRUE" = "green")) +
  facet_wrap(~ Scaffold, scales = "fixed", ncol = 1) +
  theme_minimal() +
  geom_vline(data = snps_157c, aes(xintercept = Position, color = SNP), linetype = "solid", color = "red") +
  geom_text_repel(data = snps_157c, aes(x = Position, y = as.numeric(Scaffold) + 0.2, label = SNP),
                  inherit.aes = FALSE, angle = 45, size = 2,
                  nudge_y = 0.1,   # Adjust this value as needed
                  direction = "y") +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(), panel.spacing = unit(0.1, "lines"),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.text = element_text(size = 12, face = "bold", hjust = 0.5)) +
  scale_x_continuous(labels = label_mb, breaks = scales::pretty_breaks(n = 6)) +
  labs(x = "Position", title = "Continuous Stretches of Linked SNPs by Scaffold - Cluster 6 AUT", fill = "Linked") +
  guides(fill = "none")

# Save Venn diagram to PDF
output_path <- here("output", "ldna", "figures", "cluster_6_aut_scaffolds_with_small_windows.pdf")
ggsave(output_path, height = 5, width = 8, units = "in")
```

Calculate size
```{r calculate-stretch-sizes}
# Calculate the length of each stretch
rect_data <- rect_data %>%
  mutate(length = end - start + 1)

# Calculate average sizes for TRUE and FALSE stretches
average_sizes <- rect_data %>%
  group_by(linked) %>%
  summarize(average_size = mean(length))

# Display the average sizes
print(average_sizes)
```



We can also check what genes are in these scaffolds.

Import the data
```{r load-snps-genes-chr}
snps_genes_chr <- readRDS(here("output", "ldna", "snps_genes_chr.rds"))
```

The expression data
```{r load-gene-expression-data}
gene_expression <- read_delim(here("data", "files","MANvsAUTO_sig_mRNAs.csv"), delim = ",", col_names = TRUE, show_col_types = FALSE) |>
  dplyr::select(
    gene,log2FoldChange
  ) |>
  dplyr::rename(
    Gene_ID = gene
  )
head(gene_expression)
```

For example, what genes are in the cluster 14, so we can check what genes are in these scaffolds or more specifically what SNPs are in the genes

First, what genes are in these scaffolds?

```{r get-genes-cluster-14-scaffolds}
cluster_14_genes <- snps_genes_chr[snps_genes_chr$Scaffold %in% cluster_14_all$Scaffold, ]

# How many genes in the scaffolds
length(unique(cluster_14_genes$Gene_ID))
```


Now lets check what genes have the linked SNPs

```{r get-genes-cluster-14-linked-snps}
cluster_14_snps2 <- snps_genes_chr[snps_genes_chr$SNP %in% aut_ch2$`3215_0.78`, ]

# How many genes with SNPs from the ld block
length(unique(cluster_14_snps2$Gene_ID))
```

Save as Excel file
```{r save-cluster-14-snps2-xlsx}
# Save the data frame to an Excel file
write_xlsx(cluster_14_snps2, here("output", "ldna", "cluster_14_snps2.xlsx"))
```



We can count how many linked SNPs per gene
```{r count-snps-per-gene-cluster-14}
# Count the number of SNPs for each Gene_ID
snp_count_per_gene <- cluster_14_snps2 %>%
  group_by(Gene_ID) %>%
  summarise(SNP_count = n())

# View the first few rows of the result
head(snp_count_per_gene)
```

Save as Excel file
```{r save-snp-count-per-gene-xlsx}
# Save the data frame to an Excel file
write_xlsx(snp_count_per_gene, here("output", "ldna", "snp_count_per_gene.xlsx"))
```


Now we can check how many DE genes are among the 109 genes that have SNPs on the cluster 14
```{r check-de-genes-in-cluster-14}
cluster_14_snps3 <- cluster_14_snps2[cluster_14_snps2$Gene_ID %in% gene_expression$Gene_ID, ]

# How many genes in the scaffolds
length(unique(cluster_14_snps3$Gene_ID))
unique(cluster_14_snps3$Gene_ID)
```

Get the 17 SNPs position
```{r get-genes-for-157b-snps}
# Get the 17 SNPs
genes_17_snps <- snps_genes_chr |>
  dplyr::filter(SNP %in% snps_157b$SNP)

genes_17_snps <- snps_genes_chr[snps_genes_chr$SNP %in% snps_157b$SNP, ]

head(genes_17_snps)
```

Save as Excel file
```{r save-cluster-14-17-snps-xlsx}
# Save the data frame to an Excel file
write_xlsx(genes_17_snps, here("output", "ldna", "cluster_14_17_snps.xlsx"))
```


Get the SNPs for which we do not have genes
```{r get-snps-without-gene-annotation}
genes_17_snps2 <- snps_157b |>
  dplyr::filter(!(SNP %in% genes_17_snps$SNP))
head(genes_17_snps2)
```

Save it (17 genes)
```{r save-snps-157b-txt}
write.table(
  snps_157b,
  file      = here(
    "output", "ldna", "snps_157b.txt"
  ),
  sep       = "\t",
  row.names = FALSE,
  col.names = FALSE,
  quote     = FALSE
)
```


## 8. Cluster gene intersection with differentially expressed genes

This section identifies genes within the two major LD clusters (cluster 14 on chromosome 2 and cluster 6 on chromosome 1) and checks whether any overlap with differentially expressed genes from the NON-AUTO vs AUTO comparison.

The gene coordinate data (`snps_genes_chr.rds`) and expression data (`snps_expression.rds`) were generated in File S5.

### 8.1 Cluster 14 genes

```{r load-snps-genes-chr-section8}
snps_genes_chr <-
  readRDS(
    file = here(
      "output", "ldna", "snps_genes_chr.rds"
      )
    )
head(snps_genes_chr)
```

```{r load-snps-expression}
snps_expression <-
  readRDS(
    here(
      "output", "ldna", "snps_expression.rds"
      )
    )
head(snps_expression)
```

```{r select-cluster-14-snp-ids}
cluster_14_SNPs <-
  cluster_14 |>
  select(
    SNP
  )

head(cluster_14_SNPs)
```


##### 8.1.1 Genes from coding regions

```{r join-cluster-14-snps-genes}
genes_cluster_14 <- inner_join(cluster_14_SNPs, snps_genes_chr, by = "SNP", suffix = c("", ""))

head(genes_cluster_14)
```

```{r save-genes-cluster-14-xlsx}
write_xlsx(
  genes_cluster_14,
  here(
    "output", "ldna","genes_cluster_14.xlsx"
    )
  )
```


##### 8.1.2 Genes with differential expression

```{r join-cluster-14-snps-de-genes}
DE_genes_cluster_14 <- inner_join(cluster_14_SNPs, snps_expression, by = "SNP", suffix = c("", ""))

head(DE_genes_cluster_14)
```

```{r save-de-genes-cluster-14-xlsx}
write_xlsx(
  DE_genes_cluster_14,
  here(
    "output", "ldna","DE_genes_cluster_14.xlsx"
    )
  )
```


### 8.2 Cluster 6 genes

```{r select-cluster-6-snp-ids}
cluster_6_SNPs <-
  cluster_6 |>
  select(
    SNP
  )

head(cluster_6_SNPs)
```


##### 8.2.1 Genes from coding regions

```{r join-cluster-6-snps-genes}
genes_cluster_6 <- inner_join(cluster_6_SNPs, snps_genes_chr, by = "SNP", suffix = c("", ""))

head(genes_cluster_6)
```

```{r save-genes-cluster-6-xlsx}
write_xlsx(
  genes_cluster_6,
  here(
    "output", "ldna","genes_cluster_6.xlsx"
    )
  )
```


##### 8.2.2 Genes with differential expression

```{r join-cluster-6-snps-de-genes}
DE_genes_cluster_6 <- inner_join(cluster_6_SNPs, snps_expression, by = "SNP", suffix = c("", ""))

head(DE_genes_cluster_6)
```

```{r save-de-genes-cluster-6-xlsx}
write_xlsx(
  DE_genes_cluster_6,
  here(
    "output", "ldna","DE_genes_cluster_6.xlsx"
    )
  )
```
